Choline supplementation in early life improves and low levels of choline can impair outcomes in a mouse model of Alzheimer’s disease

Version 1

Abstract

Abstract

Maternal choline supplementation (MCS) improves cognition in Alzheimer’s disease (AD) models. However, effects of MCS on neuronal hyperexcitability in AD are unknown. We investigated effects of MCS in a well-established mouse model of AD with hyperexcitability, the Tg2576 mouse. The most common type of hyperexcitability in Tg2576 mice, and many other mouse models and AD patients, are generalized EEG spikes (interictal spikes; IIS). Hyperexcitability is also reflected by elevated expression of the transcription factor ΔFosB in the principal cells of the dentate gyrus (DG), granule cells (GCs). We also studied the hilus of the DG because hilar neurons regulate GC excitability. We found reduced expression of the neuronal marker NeuN within hilar neurons in Tg2576 mice, which other studies have shown is a sign of oxidative stress or other pathology.

Tg2576 breeding pairs received a diet with a relatively low, intermediate or high concentration of choline. After weaning, all mice received the intermediate diet. In offspring of mice fed the high choline diet, IIS frequency declined, GC ΔFosB expression was reduced, and NeuN expression restored. Spatial memory improved. In contrast, offspring exposed to the relatively low choline diet had several adverse effects, such as increased mortality. The results provide new evidence that a diet high in choline in early life can improve outcomes in a mouse model of AD, and relatively low choline can have negative consequences. This is the first study showing that dietary choline can regulate hyperexcitability, hilar neurons, and spatial memory in an animal model of AD.

Introduction

Diet has been suggested to influence several aspects of brain health. One dietary intervention that has been studied extensively in rodents and humans is supplementation of the maternal diet with the nutrient choline, (maternal choline supplementation; MCS). MCS improves many aspects of brain health in humans (Zeisel & da Costa 2009, Jiang et al. 2014). In normal rats, MCS also is beneficial, with numerous studies showing improved behavior in the adult offspring (for review, see (for review see Meck & Williams 2003)).

In Alzheimer’s disease (AD), changes to the diet have been recommended (Bourre 2006, Power et al. 2019, Mao et al. 2021, Lobo et al. 2022), including MCS (Strupp et al. 2016, Velazquez et al. 2020). The recommendation to use MCS is based primarily on data from the Ts65Dn mouse model of Down syndrome (DS) and AD, where MCS led to improved memory and attention in Ts65Dn offspring (Strupp et al. 2016, Powers et al. 2017, Powers et al. 2021). In addition, degeneration of basal forebrain cholinergic neurons (BFCNs) in Ts65Dn mice, a hallmark of DS and AD, was reduced (Kelley et al. 2016, Powers et al. 2017).

Based on the studies of Ts65Dn mice, we asked if MCS would improve another impairment that has been found in AD patients and mouse models of AD, hyperexcitability. We chose to study Tg2576 mice because hyperexcitability is robust, and BFCNs are likely to play a role (Kam et al. 2016, Lisgaras & Scharfman 2023). In our past work, the primary type of hyperexcitability was interictal spikes (IIS), which are named because they occur in between seizures (ictal events) in epilepsy. They are studied by EEG. We also studied hyperexcitability using ΔFosB expression, a transcription factor that is increased when neurons have been highly active over the prior 10-14 days (McClung et al. 2004). We studied ΔFosB in granule cells (GCs) of the dentate gyrus (DG), because numerous ΔFosB-expressing GCs occur when there are IIS (You et al. 2017, You et al. 2018).

We also asked if behavior would improve if MCS reduced IIS frequency in Tg2576 mice. The basis for this question is in studies of epilepsy, where IIS disrupt cognition (Rausch et al. 1978, Aarts et al. 1984, Holmes & Lenck-Santini 2006, Kleen et al. 2010, Kleen et al. 2013, Gelinas et al. 2016). Also, IIS in the Tg2576 mouse start early in life (by 1 month of age), before memory impairment, which occurs at 3-4 months of age (Duffy et al. 2015). Therefore, IIS might disrupt activity sufficiently to promote behavioral impairments. We previously found that the novel object location (NOL) task is impaired at 3-4 months of age in Tg2576 mice (Duffy et al. 2015), so we evaluated NOL, and a related task, novel object recognition (NOR; Vogel-Ciernia & Wood 2014). We also studied expression of NeuN, a neuronal marker that is reduced in numerous disease conditions (Buckingham et al. 2008, Kadriu et al. 2009, Matsuda et al. 2009, Duffy et al. 2011, Duffy et al. 2015). Here we report loss of NeuN-ir in the DG hilus of Tg2576 mice, which is important because impaired hilar neurons could cause GC hyperexcitability (Sperk et al. 2007, Scharfman & Myers 2012).

We fed dams one of three diets, which were relatively low, intermediate or high in choline. The high choline diet provided levels of choline similar to other studies of MCS in rodents (Meck et al. 1988, Loy et al. 1991, Holler et al. 1996, Meck & Williams 1999, Sandstrom et al. 2002, Mellott et al. 2004, Glenn et al. 2012, Kelley et al. 2019). After weaning the intermediate diet was used. Offspring were implanted with electrodes for EEG at 1 month of age, and 24 hr-long recordings began 1/month starting at 1 week after surgery. NOL and NOR were tested at 3 months and again at 6 months, and then we perfusion-fixed mice, sectioned the brains, and evaluated ΔFosB and NeuN protein expression (Figure 1A).

Schematic of the behavioral procedures for the NOL and NOR tasks.

A. Experimental timeline. Dams were fed one of three diets for mating, gestation, and offspring consumed the diet until weaning at 25-30 days of age. At 1 month-old, offspring were implanted with electrodes for EEG and recorded for a 24 hr period at 1.2, 2, 3, 4, 5, and 6 months of age. At 3 and 6 months of age behavior was tested. At 6 months of age, after behavioral testing, mice were perfused, brains were sectioned, and sections were processed with antibodies against NeuN and ΔFosB.

B. Prior to NOL or NOR, animals were acclimated to the testing arena. There were three acclimation sessions separated by 24hrs during which animals were allowed to freely explore for 5 min.

C. In the NOL task, animals were placed in a cage with two identical objects and allowed to freely explore for 5 min (Training). After one hour they were brought back to the cage, where one object was displaced, and allowed to freely explore for 5 min (Testing).

D. In the NOR task. animals were placed in a cage with two identical objects and allowed to freely explore for 5 min (Training). After one hour they were brought back to the cage, where one object was replaced, and allowed to freely explore for 5 min (Testing).

The results showed a remarkable effect of the high choline diet. IIS and spatial memory were improved, as was ΔFosB and NeuN expression. Interestingly, the relatively low choline diet had mixed effects, reducing IIS frequency, but making memory, ΔFosB, and NeuN worse. The mice also died prematurely. In summary, we make a strong argument for choline supplementation in early life to improve outcomes in an AD model, and potentially AD.

Methods

Animals

All experimental procedures followed the guidelines set by the National Institute of Health (NIH) and were approved by the Institutional Animal Care and Use Committee (IACUC) of the Nathan Kline Institute for Psychiatric Research. Mice were housed in standard mouse cages with corn cob bedding and a 12-hr light-dark cycle. Food and water were provided ad libitum.

Tg2576 mice express a mutant form of human APP (isoform 695) with a mutation found in a Swedish cohort (Lys670Arg, Met671Leu). The expression is driven by the hamster prion protein promoter. Mice were bred in-house from heterozygous Tg2576 males and non-transgenic female mice (C57BL6/SJL F1 hybrid). Wildtype (WT) mice were obtained from Jackson Laboratories (Stock# 100012). At weaning, genotypes were determined using a protocol to detect APP695.

Breeding pairs were randomly assigned to receive one of three diets with different concentrations of choline chloride: 1.1 g/kg (AIN-76A, Dyets), 2.0 g/kg (Purina 5008, W.F. Fisher and Son Inc.) and 5.0 g/kg (AIN-76A, Dyets; Supplemental Table 1). These diets (relatively “low choline” diet, “intermediate choline” diet and “high choline” diet, respectively) were used until weaning (25-30 days of age). After weaning, all mice were fed the intermediate diet. Mice were housed with others of the same sex (1-4 per cage).

Behavior

General information

Mice were housed in the room where they would be tested behaviorally starting at least 24 hr prior to all experiments. Both NOL and NOR included three acclimations (5 min each) followed by a training and a testing session (5 min each; Figure 1B-D). The interval between training and testing was 60 minutes and therefore tested short-term memory (Vogel-Ciernia & Wood 2014). NOL and NOR tests were separated by 7 days to ensure effects of one task did not affect the next. Prior studies have failed to find an effect on one task on the next when a 7 day-long interval is used (Botterill et al. 2021). The order of testing (NOL before NOR or NOR before NOL) was randomized.

Video recordings of all training and testing sessions were captured using a USB camera (Logitech HD Pro C920, Logitech). Acclimations were conducted between 10:00 a.m. and 12:00 p.m., while training and testing was conducted between 1:00 p.m. and 4:00 p.m. All equipment was cleaned using 70% ethanol before each use. In pilot studies, we chose objects that would be explored equally by animals. Indeed, animals showed no preference for either of the objects (i.e., LEGO object and pineapple, Supplemental Figure 1) during training, as reported in the Results.

NOL

NOL was conducted in a standard rat cage (26 cm wide x 40 cm long x 20 cm high) with different pictures on 3 sides of the cage to provide a consistent context and therefore foster acclimation. Pictures included several shapes and colors. The dimensions of the pictures were 1) 10 cm x 21 cm, 2) 16 cm x 18 cm, and 3) 17 cm x 20 cm. During the acclimations, animals were allowed to freely explore the cage. During the training session, mice were placed in the cage where they had been acclimated, with two identical objects, 1 in a corner of the cage (e.g., left top) and the second in the adjacent corner (e.g., left bottom, Supplemental Figure 1). The mice were then removed and placed in their home cage for 1 hr. During the test session, one of the objects was moved to the opposite end of the cage (left top to right top; Supplemental Figure 1). The two objects were identical pineapple-like objects made of metal that were painted and approximately 3 cm x 3 cm x 5 cm (Supplemental Figure 1; Botterill et al. 2021).

NOR

NOR was conducted in a standard rat cage as described above. During the acclimations, animals were allowed to freely explore the cage. During the training session, mice were placed in the cage used for acclimation with two identical objects centered along the shortest cage wall (Supplemental Figure 1). The mouse was then removed and placed in their home cage for 1 hr. During the testing session, one of the objects was replaced with a different one. The objects that were identical were 2 pineapple-like objects, and the new object was made of 2 green LEGO pieces approximately 5 cm x 7 cm x 7 cm (Supplemental Fig. 1; Botterill et al. 2021).

Quantification

The experimenter who conducted the analysis was blind to the genotype of the offspring and which object was novel or familiar. Videos of the training and testing sessions were analyzed offline. Exploration was quantified based on the guidelines of Vogel-Ciernia & Wood (2014), but in addition to time spent exploring, the number of approaches to an object was also quantified. Time spent exploring each object was defined as the duration of time the nose was pointed at the object and the nose was located within 2 cm of the object. In addition, exploration time included the time animals spent on top of the object if they were looking down and sniffing it. An approach was defined by a movement toward the object and ending at least 2 cm from the edge of the object.

Animals that remembered the objects explored during the training phase were expected to demonstrate an increased preference for novelty in the test phase (Ennaceur & Delacour 1988). In other words, exploration of the novel object during testing was expected to be higher than 50% of the total time of object exploration.

Anatomy

Perfusion-fixation and sectioning

Mice were initially anesthetized by isoflurane inhalation (NDC# 07-893-1389, Patterson Veterinary), followed by urethane (2.5 g/kg; i.p.; Cat# U2500, Sigma-Aldrich). Under deep anesthesia, the abdominal and heart cavities were opened and the animal was transcardially-perfused using 10 mL of room temperature (RT) saline (0.9% sodium chloride in double distilled (dd) H2O) using a peristaltic pump (Minipuls1; Gilson), followed by 30 mL of 4°C 4% paraformaldehyde (PFA, Cat# 19210, Electron Microscopy Sciences) in 0.1 M phosphate buffer (PB; pH 7.4). The brains were immediately extracted, hemisected, and post-fixed for at least 24 hr in 4% PFA (at 4°C).

Following post-fixation, the left hemisphere was sectioned in the coronal plane and the right hemisphere was sectioned in the horizontal plane (50 µm-thick sections) using a vibratome (VT 1000P, Leica Biosystems). Sections were collected serially and stored at 4°C in 24-well tissue culture plates containing cryoprotectant solution (30% glycerol, 30% ethylene glycol, 40% 0.1 M phosphate buffer, pH 6.7). Throughout the hippocampus, every 6th section was processed with an antibody to NeuN or ΔFosB.

NeuN and ΔFosB immunohistochemistry

Free floating sections were washed in 0.1 M Tris Buffer (TB; 3 x 5 min), followed by a 3 min wash in 1% H2O2. Sections were then washed in 0.1 M TB (3 x 5 min) and incubated for 60 min in 5% horse serum for NeuN (Cat# S-2000, Vector) or 5% goat serum for ΔFosB (Cat# S-1000, Vector), diluted in a solution of 0.25% (volume/volume or v/v) Triton-X 100, and 1% (weight/volume or w/v) bovine serum albumin in 0.1 M TB. Sections were then incubated overnight at 4°C in primary antibody to NeuN (mouse monoclonal, 1:5000; Cat# MAB377, Chemicon International) or anti-ΔFosB (rabbit monoclonal, 1:1000; Cat# D3S8R, Cell Signaling), diluted in a solution of 0.25% (v/v) Triton-X 100, and 1% (w/v) bovine serum albumin in 0.1M TB. Both NeuN and ΔFosB have been well-characterized as antigens (Mullen et al. 1992, Wolf et al. 1996, Chen et al. 1997, Sarnat et al. 1998) and the antibodies we used have been commonly employed in the past (Chen et al. 1997, Duffy et al. 2013, Corbett et al. 2017).

On the following day, sections were washed in 0.1 M TB (3 x 5 min) and then incubated for 60 min in biotinylated horse anti-mouse secondary antibody for NeuN (1:500, Cat# BA-2000, Vector) or biotinylated goat anti-rabbit secondary antibody for ΔFosB (1:500, Cat# BA-1000), diluted in a solution of 0.25% (v/v) Triton-X 100, and 1% (w/v) bovine serum albumin in 0.1 M TB. The sections were then washed in 0.1 M TB (3 x 5 min) and incubated in avidin-biotin complex for 2 hrs (1:1000; Cat# PK-6100, Vector). They were washed in 0.1 M TB (3 x 5 min) and then reacted in a solution containing 0.5 mg/mL 3, 3’-diaminobenzidine (DAB; Cat# D5905, Sigma-Aldrich), 40 µg/mL ammonium chloride (Cat# A4514, Sigma-Aldrich), 25 mg/mL D(+)-glucose (Cat# G5767, Sigma-Aldrich), and 3 g/mL glucose oxidase (Cat# G2133, Sigma-Aldrich) in 0.1 M TB. This method slowed the reaction time so that the reaction could be temporally controlled, meaning that the reaction could be stopped when the immunoreactivity was robust but background was still low. The sections were then washed in 0.1 M TB (3 x 5 min), mounted on gelatin-coated slides (1% bovine gelatin, Cat# G9391, Sigma-Aldrich) and dried at RT overnight. The following day they were dehydrated in increasing concentrations of ethanol (90%, 10 min; 95% 10 min; 100%, 10 min; 100% again, 10 min), washed in Xylene (10 min; Cat# 534056, Sigma-Aldrich) and cover-slipped with Permount (Cat# 17986-01, Electron Microscopy Sciences).

Analysis

Photomicrographs were acquired using ImagePro Plus V7.0 (Media Cybernetics) and a digital camera (Model RET 2000R-F-CLR-12, Q-Imaging). NeuN and ΔFosB staining were quantified from micrographs using ImageJ (V1.44, National Institutes of Health). All images were first converted to grayscale and in each section, the hilus was traced, defined by zone 4 of Amaral (1978). A threshold was then calculated to identify the NeuN-stained cell bodies but not background. Then NeuN-stained cell bodies in the hilus were quantified manually. Note that the threshold was defined in ImageJ using the distribution of intensities in the micrograph. A threshold was then set to identify the dark cells but not the background slide using A slider was pushed from the low level of staining corresponding to background to the location where staining intensity makes a sharp rise, reflecting strongly stained cells. Below that was considered background and above that was defined as above threshold.

To count ΔFosB-stained cells, images were converted to grayscale and in each section, the dentate gyrus GCL was outlined and defined as a region of interest (ROI). A threshold was then set to identify the dark cells but not the background, analogous to the method to define threshold in the preceding paragraph. Two thresholds were used. The first threshold was most inclusive of immunoreactive cells. Presumably this group reflected cells with less neuronal activity. Then counts were made a second time with a higher threshold. This group probably corresponded to cells with the highest activity. ImageJ was used to calculate the area (in pixels) within the ROI that was above threshold.

Video-electroencephalographic (video-EEG) recordings

Stereotaxic surgery

EEG electrodes (stainless steel jeweler’s screws) were implanted in Tg2576 animals at 4 weeks of age. WT littermates were not implanted because previous work from our laboratory has shown that WT mice have no IIS (Kam et al. 2016). The animals were anesthetized by isoflurane inhalation (3% isoflurane. 2% oxygen for induction). For maintenance during surgery, isoflurane was < 1.75% and flow rate was 1.2 L/min. Mice were placed in a stereotaxic apparatus (David Kopf Instruments). Prior to the implantation of the electrodes, animals were injected with the analgesic Buprenex (0.2 mg/kg, s.c.; buprenorphine hydroxide, NDC# 12496-0757-5, Reckitt Benckiser), which was diluted in saline (0.03 mg/mL in sterile 0.9% sodium chloride solution, Vedco Inc.). The skull was exposed with a midline incision and six holes were drilled for the placement of subdural screw electrodes (Cat# 8209, 0.10” stainless steel screws, Pinnacle Technology).

The coordinates for the electrode placement were: right occipital cortex (AP - 3.5 mm, ML 2.0 mm), left frontal cortex (AP -0.5 mm, ML -1.5 mm), left hippocampus (AP -2.5 mm, ML -2.0 mm), and right hippocampus (AP -2.5 mm, ML 2.0 mm (Kam et al. 2016). Two additional screws were placed, one over the right olfactory bulb as ground (AP 2.3 mm, ML 1.8 mm) and another at the midline over the cerebellum as a reference (relative to Lambda: AP -1.5 mm, ML -0.5 mm). Screws were attached to an eight-pin connector (Cat# ED85100-ND, Digi-Key Corporation), which was placed over the skull and secured with dental cement (Cat# 4734FIB, Lang Dental Mfg. Co., Inc.).

After surgery, animals were placed on a heating blanket overnight and injected with lactated Ringer’s solution (50 mL/kg at 31°C; NDC# 099355000476, Aspen Veterinary Resources Ltd). They were then transferred to the room where the video-EEG is recorded. Implanted animals were allowed to recover for one week following the surgery.

Video-EEG recording

Video-EEG was recorded stating at 1.25 months (5 weeks) of age. The animals were then recorded at 2, 3, 4, 5 and 6 months of age. Each recording session lasted 24 hr so that a long period of sleep could be acquired, since IIS occur primarily in sleep (Kam et al. 2016).

Mice were placed into 21 cm x 19 cm transparent cages with food and water provided ad libitum and corncob bedding. A pre-amplifier was inserted into the eight-pin connector on the skull, which was connected to a multichannel commutator (Cat# 8408, Pinnacle Technology). This arrangement allowed for free range of movement throughout the recording. EEG signals were acquired at 2000 Hz and bandpass filtered at 0.5 – 200 Hz using Sirenia Acquisition) V2.0.4, Pinnacle Technology). Simultaneous video recordings were captured using an infrared camera (Cat# AP-DCS100W, Apex CCTV).

Analysis

EEG recordings were analyzed offline with Neuroscore V3.2.1 (Data Science International). IIS were defined as large amplitude, 10-75 msec deflections occurring synchronously in all four leads (Kam et al. 2016). They were quantified by first setting an amplitude threshold (calculated separately for each recording) and a duration criterion (10-75 msec). To determine the amplitude threshold, we calculated the root mean square (RMS) amplitude for noise in each recording during a 60 sec segment of baseline activity that did not include any artifacts. The threshold for IIS was set at 9 standard deviations above the RMS amplitude. This threshold was selected because it identified IIS extremely well and excluded movement or other artifacts. The IIS in one of the hippocampal channels was used because of data suggesting IIS begin in the hippocampus (Kam et al. 2016). Following the automatic analysis of the IIS, the software’s detection accuracy was verified manually for each recording to ensure that all spikes appeared in all four channels and that no artifacts were included in the final IIS count.

Statistical comparisons

Data are expressed as mean ± standard error of the mean (SEM). P-value was set to <0.05 prior to all experiments. Parametric data comparing two groups used unpaired two-tailed t-tests. For >2 groups, one-way ANOVA was used followed by Tukey-Kramer post-hoc tests that corrected for multiple comparisons. For data with two main factors, two-way ANOVA was followed by Tukey-Kramer post-hoc tests. Interactions are not reported unless they were significant.

For the analysis of IIS frequency, a repeated measures ANOVA was not possible because animal numbers were not the same at all ages. This was due to mortality in low choline-treated mice or other unidentified reasons. Therefore, a mixed effects analysis was conducted.

For non-parametric data, Mann-Whitney U tests were used to compare two groups and Kruskal-Wallis for >2 groups. The post-hoc tests were Dunn’s test for multiple comparisons. Tests were conducted using Prism (V 9.4.1, Graphpad Software).

Results

Behavior

NOL and the NOR, unlike other tasks previously used to evaluate cognitive deficits in >9 months-old Tg2576 mouse model (i.e., Morris Water Maze, Radial Arm Water Maze, Y-Maze; Yassine et al. 2013, Wolf et al. 2016), provided an opportunity to capture cognitive deficits in animals as young as three months-old (Duffy et al. 2015). Figure 1B-D shows a schematic of the specific experimental procedures for NOL and NOR.

NOL

The results of the NOL task are presented in Figure 2. Exploration of the novel object is expressed as a percentage of the total exploration of both objects. During the training session, animals did not show preference for one object over the other and this result was independent of genotype. Thus, a two-way ANOVA revealed no main effects of genotype on novel object exploration during training at three (F(1,97) = 0.40, p = 0.53) or six months of age (F(1,70) = 1.94, p = 0.17). There also was no effect of maternal diet on training at three (F(2,97) = 0.06, p = 0.94) or six months of age (F(2,70) = 2.94, p = 0.06).

Choline enrichment reduced spatial memory deficits.

A. Three months-old mice.

1. WT and Tg2576 mice exposed to the low choline diet showed spatial memory deficits.

2. Tg2576 exposed to the intermediate diet showed spatial memory deficits but not WT.

3. Spatial memory was improved in animals exposed to the high choline diet.

B. Six months-old mice.

1. WT and Tg2576 mice exposed to the low choline diet showed spatial memory deficits.

2. WT and Tg2576 mice exposed to the intermediate diet showed spatial memory deficits.

3. WT and Tg2576 mice exposed to the high choline diet had improved spatial memory.

Next, we asked whether Tg2576 mice were deficient in the test phase of NOL. First we studied the 3 months-old mice. We used a two-way ANOVA with genotype and task phase (training versus testing) as main factors. In animals that received the intermediate diet, a two-way ANOVA revealed a significant main effect of the phase of the task (training or testing; F(1,24) = 7.48, p = 0.01). Consistent with previous studies (Duffy et al. 2015), post-hoc analyses revealed a significant increase in the exploration of the novel object during testing in WT (p = 0.02) but not Tg2576 mice (p = 0.50; Figure 2A2). Thus, Tg2576 mice were impaired in NOL when fed the intermediate diet.

In animals that received the high choline diet, a two-way ANOVA revealed a main effect of the task phase (F(1,36) = 48.23, p < 0.001). Post-hoc analyses revealed a significant increase in the exploration of the novel object during testing in both WT (p < 0.001) and Tg2576 mice (p < 0.001; Figure 2A3). Therefore, high choline supplementation improved memory in Tg2576 mice.

In animals treated with the low choline diet, there was no effect of phase of the task (F(1,37) = 1.39, p = 0.25). These data suggesting that relatively low levels of choline during early life impaired spatial memory in both WT and Tg2576 mice (Figure 2A). Thus, low choline had a significant adverse effect. Adverse effects are further supported by survival plots showing that there was more mortality at earlier ages in offspring exposed to the low choline diet (Supplemental Figure 1).

Figure 2B shows the results in 6 months-old WT and Tg2576 mice. In animals that received the intermediate diet, a two-way ANOVA showed no main effect of genotype (F(1,18) = 0.01, p = 0.92) or task phase (F(1,18) = 3.40, p = 0.08), revealing age-related memory deficits even in WT mice at six months of age (Figure 2B). In contrast, the high choline group showed a main effect of task phase (F(1,29) = 28.32, p < 0.001). Post-hoc analyses showed a significant increase in the exploration of the novel object during testing in both WT (p < 0.001) and Tg2576 (p = 0.02; Figure 2B). In animals that received the relatively low choline diet, there was no effect of task phase (F(1,25) = 3.00, p = 0.10; Figure 2B). Thus, low choline had adverse effects on NOL in both WT and Tg2576 mice.

Taken together, these results demonstrate the lasting beneficial effects of the high choline diet and adverse effects of low choline on offspring in a spatial memory task in Tg2576 as well as WT littermates.

NOR

We first confirmed that animals did not show preference for one object over the other in training. A two-way ANOVA revealed no main effects of genotype at three months (F(1,71) = 2.59, p = 0.54) and no effect of diet (F(2,71) = 0.22, p = 0.81). The same result was obtained at six months of age (genotype: (F(1,66) = 0.11, p = 0.75); diet: (F(2,66) = 0.98, p = 0.38). Thus, novel object exploration during training approached 50% in all treatment groups, independent of genotype and maternal diet, and at both ages.

Figure 3A shows the results of the testing phase of the NOR task in 3 months-old WT and Tg2576 mice. In animals that received the intermediate choline diet, a two-way ANOVA revealed a significant main effect of task phase (F(1,19) = 40.53, p < 0.01) and no effect of genotype (F(1,19) = 0.78, p = 0.39). Post-hoc analyses revealed a significant increase in the exploration of the novel object during testing compared to training for both WT (p < 0.01) and Tg2576 (p < 0.01; Figure 3A). These results suggest that Tg2576 did not have a deficit in NOR at 3 months of age.

WT and Tg2576 mice showed recognition memory regardless of diet.

A. Three months-old mice.

1-3. WT and Tg2576 mice treated performed the NOR task.

B. Six months-old mice.

1-3. WT and Tg2576 mice performed the NOR task.

Mice fed the high choline diet also showed no deficit in NOR at 3 months of age. Thus, a two-way ANOVA revealed a main effect of task phase (F(1,26) = 17.51, p < 0.01) but not genotype (F(1,26) = 3.53, p = 0.07). Post-hoc analyses revealed a significant increase in the exploration of the novel object during testing in both WT (p < 0.01) and Tg2576 (p = 0.03; Figure 3A).

In animals treated with the low choline diet in early life, there was no impairment in NOR at 3 months of age (task phase (F(1,26) = 22.61, p < 0.01; genotype (F(1,26) = 0.02, p = 0.89). There was a significant increase in the exploration of the novel object in testing both in WT (p = 0.03) and Tg2576 (p < 0.01).

Figure 3B shows that there were no impairments at 6 months of age. In animals that received the intermediate diet, a two-way ANOVA revealed a significant main effect of task phase (F(1,19) = 13.65, p < 0.01) but no main effect of genotype (F(1,19) = 0.28, p = 0.60). Animals explored the novel object significantly more during testing in WT (p = 0.03) and Tg2576 mice (p = 0.04; Figure 3B). In animals that received the other diets, results were the same. For the high choline group, there was an effect of task phase (F(1,27) = 16.26, p < 0.01) but not genotype (F(1,27) = 0.24, p = 0.63) and a significant increase in the exploration of the novel object during testing in both WT (p = 0.04) and Tg2576 (p < 0.01; Figure 3B). Similarly, in animals treated with the low choline diet, there was a main effect of task phase (F(1,26) = 32.66, p < 0.01), no effect of genotype (F(1, 6) = 0.05, p = 0.82), and more exploration of the novel object during the test phase (WT, p < 0.01; Tg2576, p < 0.01).

Thus, at both 3 and 6 months of age, WT and Tg2576 mice performed well in the NOR task. These results suggest that young Tg2576 mice are less sensitive to NOR than NOL. The greater sensitivity to NOL is consistent with past demonstrations that the dentate gyrus contributes to NOL (Sahay et al. 2011, Kesner et al. 2015, Spyrka & Hess 2018, Vandrey et al. 2020, Gulmez Karaca et al. 2021, GoodSmith et al. 2022), and early signs of adverse effects in the dentate gyrus in mouse models of AD (impaired LTP; Jacobsen et al. 2006, altered gene expression; Palop et al. 2011). Our implementation of NOL may have increased the dentate gyrus-dependence of the task by making the object locations relatively close together, because the dentate gyrus is critical to distinguishing small differences in locations. In studies by Pofahl and colleagues, it was shown that distances between objects like those we used made NOL dentate gyrus-dependent (Pofahl et al. 2021).

Exploration time

Total object exploration (TOE) was measured to address any effects of genotype or diet on the total duration of exploration of objects (Figure 4).

Tg2576 mice treated with the choline supplemented diet spent more time with the objects at each approach.

A-B. NOL.

There was no effect of diet on time exploring or approaches to the novel object in the NOL task at 3 months of age (A) but there were significant differences at 6 months of age (B). Mice exposed to the high choline diet spent more time exploring than mice that had been fed the low choline or the intermediate diet.

C-D. NOR.

There was no effect of diet on time exploring or approaches to the novel object in the NOR task at 3 months of age (C) but there were significant differences at 6 months of age (D). Mice that had been fed the high choline diet spent more time exploring than mice that had been fed the low choline or the intermediate diet.

3 months of age

For NOL at 3 months of age, there were effects of genotype (F(1,67) = 6.891, p = 0.01) but not maternal diet (F(2,87) = 0.67, p = 0.63; Figure 4A). Tukey-Kramer post-hoc tests showed that Tg2576 mice fed the low choline diet showed more exploration than WT mice fed the same diet but the effect was on the border of significance (p = 0.049; data not shown). Three months-old mice tested with NOR showed no significant effect of genotype (F(1,76) = 1.35, p = 0.25) or diet (F(2,76) = 0.30, p = 0.61; Figure 4C).

6 months of age

Exploration

NOL and NOR showed significant effects of diet on exploration at 6 months of age (Figure 4B,D). For NOL, there was no effect of genotype (F(1,66) = 0.33; p = 0.57) but there was an effect of maternal diet (F(2,66) = 8.82; p = 0.0004). For NOR, the same was true (genotype: F(1,72) = 0.96, p = 0.33; diet: F(2,72) = 8.50, p = 0.0005). We then pooled genotypes to determine which diets differed. For NOL, mice treated with the relatively low choline diet had less exploration than mice fed the high choline diet (one-way ANOVA, F(2,77) = 6.90; p = 0.002; Tukey post-hoc test, p = 0.005; Figure 4B) and the mice that were fed the intermediate diet also had less exploration than mice fed the high choline diet (Tukey post-hoc test, p = 0.008; Figure 4B). Results for NOR were the same: mice treated with the low choline diet had less exploration than mice fed the high choline diet (one-way ANOVA, F(2,74) =4.81; p = 0.02; Tukey post-hoc test, p = 0.02; Figure 4D) and the mice that were fed the intermediate diet also had less exploration than mice fed the high choline diet (Tukey post-hoc test, p = 0.03; Figure 4D).

Approaches

To gain insight into the potential reason for differences in exploration, we measured total object approaches (TOA; Figure 4A-D). TOA was defined as the number of approaches to the familiar object + number of approaches to the novel object. There were no differences by genotype (two-way ANOVA, F(1,73) = 2.930, p = 0.09) or maternal diet (F(2,73) = 2.36, p = 0.10; Figure 4A-D).

Taken together, the results indicate that, in 6 months-old mice, animals that received the high choline diet spent more time with objects at each approach. This may underlie increased object location memory in high choline-treated mice by increasing time for information processing during an approach. It may also help in the NOR task with object recognition memory, although it did not seem impaired in mice fed other diets. Another possibility is that as Tg2576 mice age they show compensatory changes that enhance memory. This has been suggested for procedural learning (Middei et al. 2004) but not for memory of objects. A third possibility is that the high choline diet reduces anxiety (Glenn et al. 2012, Langley et al. 2015, McCall et al. 2015). Reduced anxiety may lessen fear of exploring objects, and as a result, animals may spend more time with objects.

For 3 months-old mice there were no effects of maternal diet on exploration, suggesting that increased time with objects for a given approach is a change that develops with age. To our knowledge, this age-related change has not been previously reported. It could be a strategy to cope with impairments that are developing as AD pathology manifests.

Anatomy

NeuN

To further analyze the effects of dietary choline early in life, we used an antibody against a neuronal nuclear antigen, NeuN. Previous studies have found that reduced expression of NeuN often occurs in neurons after insults or injury (e.g., ischemia, toxicity and even aging; Lind et al. 2005, Portiansky et al. 2006, Buckingham et al. 2008, Kadriu et al. 2009, Matsuda et al. 2009, Won et al. 2009, Duffy et al. 2011). For example, when NeuN is reduced after a reduction in ankyrin-rich membrane spanning kinase D-interacting substrate of 220K, ARMS/Kidins, a protein critical to neurotrophin signaling, neurons in the entorhinal cortex demonstrate signs of toxicity such as pkynosis (Duffy et al. 2011, Duffy et al. 2015). Since NeuN is reduced when Aβ levels are elevated (Wu et al. 2016) and NeuN is reduced in AD patients (Camporez et al. 2021), we used NeuN expression to ask if the high choline diet could rescue the loss of NeuN-ir.

As shown in Figure 5, relatively weak NeuN-ir was observed in Tg2576 animals in the hilus, an area known to be highly vulnerable to insult and injury (Scharfman 1999). The most NeuN loss appeared to be in mice that received the low choline diet and the intermediate diet compared to the high choline diet (Figure 5B). Therefore, we conducted a two-way ANOVA with diet and septotemporal levels as main factors. To examine the septotemporal axis, quantification from 2-3 dorsal coronal sections were averaged to provide a value for dorsal DG and 2-3 ventral horizontal sections were averaged for a ventral value. There was no effect of diet (F(2,30) = 2.11, p = 0.14) nor septotemporal level (F(1,30) = 0.02, p = 0.88) but there was a significant interaction of diet and septotemporal level (F(2,30) = 3.88, p = 0.04). A one-way ANOVA using dorsal values showed a significant effect of diet (F(2,21)= 7.58, p = 0.003) but this was not the case for ventral values (F(2,23) = 0.13; p = 0.88; Figure 5D). In the dorsal DG, there was less NeuN-ir in mice fed the low choline diet versus the high choline diet (Tukey-Kramer post hoc test, p = 0.03) and less NeuN in mice fed the intermediate vs high choline diet (p = 0.003; Figure 5D). The results suggest that choline enrichment protected hilar neurons from NeuN loss in Tg2576 mice.

Choline supplementation improves NeuN immunoreactivity (ir) in hilar cells in Tg2576 animals.

A. A sagittal view of the brain illustrating the coronal sectioning plane.

B. Representative images of NeuN-ir staining in the dorsal dentate gyrus of Tg2576 animals from each treatment group.

C. Quantification methods. Representative images demonstrating the thresholding criteria established to quantify NeuN-ir.

1. A NeuN-stained section shows an inset (arrow) expanded at the top left to show 3 hilar cells. NeuN-ir cells are marked by blue arrows. NeuN-ir that was too close to background staining to be considered positive is marked by a green arrow.

2. After converting the image to grayscale, a threshold was set so that all cells stained well with NeuN antibody were red. The inset shows that the two cells that were darkest met the threshold while the weakly stained cell did not.

D. Six months-old Tg2576 mice treated with the high choline diet had significantly more NeuN-ir cells in the hilus in the dorsal dentate gyrus (1) compared to other treatment groups but not ventral dentate gyrus (2).

ΔFosB

To complement the information from the video-EEG recordings we used a marker of elevated neuronal activity, ΔFosB (Figure 6). ΔFosB is a truncated variant of the transcription factor FosB, which is increased by enhanced neuronal activity; ΔFosB has a half-life of approximately 8 days (Ulery-Reynolds et al., 2009), so when ΔFosB is elevated, it reflects increased neuronal activity over the last 10-14 days (McClung et al. 2004). Previous studies have shown that when the J20 mouse model of AD is examined with an antibody to ΔFosB, the GC layer shows extremely high levels of ΔFosB expression (Corbett et al. 2017), similar to a mouse with chronic spontaneous seizures (epilepsy; Morris et al. 2000). This is not surprising since the J20 mice have recurrent seizures (Palop et al. 2007). Therefore, we asked if Tg2576 mice would have robust ΔFosB in the GC layer, and choline supplementation would reduce it.

Choline supplementation reduced ΔFosB expression in dorsal GCs of Tg2576 mice.

A. A sagittal view of the brain illustrating the coronal sectioning plane.

B. Representative images of ΔFosB staining in the anterior dentate gyrus of Tg2576 animals from each treatment group.

C. Quantification methods. Representative images demonstrating the thresholding criteria established to quantify ΔFosB.

1. A ΔFosB -stained section shows strongly-stained cells (white arrows).

2. A strict thresholding criteria was used to make only the darkest stained cells red.

D. Six months-old Tg2576 mice treated with the choline supplemented diet had significantly more ΔFosB -ir pixels above threshold in the dorsal dentate gyrus (1) compared to other treatment groups but not ventral dentate gyrus (2).

E. Methods are shown using a threshold that was less strict.

1. Some of the stained cells that were included are not as dark as those used for the strict threshold (white arrows).

2. All cells above the less conservative threshold are shown in red.

F. With the less conservative threshold, the results were the same as D. Mice that were fed the high choline diet had less ΔFosB-ir pixels than the mice that were fed the other diets. However, this was true for dorsal (1) but not ventral (2) sections.

There was strong expression of ΔFosB in Tg2576 GCs especially in mice fed the low choline diet (Figure 6). Two-way ANOVA showed a main effect of maternal diet (F(2,18) = 3.76, p = 0.04) but there was not a main effect of septotemporal level (F(1,18) = 1.3, p = 0.26).

To clarify the effect of maternal diet, a one-way ANOVA was conducted. There was an effect of diet (F(2,9) = 10.74, p = 0.004) with post-hoc tests showing significantly less GC ΔFosB expression in high choline-treated mice compared to low choline (p = 0.003). Low choline-treated mice were not significantly different from mice that were fed the intermediate diet (p = 0.08). These data used dorsal sections so we repeated the analysis for the ventral DG, and there also was no significant effect of diet (F2,9) = 0.79, p = 0.52; Figure 6).

ΔFosB quantification was repeated with a lower threshold to define ΔFosB-ir GCs (see Methods) and results were the same (Figure 6). One-way ANOVA showed a significant effect of diet (F(2,41) = 14.66, p < 0.0001), and low choline mice had greater ΔFosB-ir than high choline mice (p =0.003). Mice fed the intermediate diet had higher expression than mice fed the high choline diet (p = 0.0001). These data suggest that high choline in the diet early in life can reduce hyperexcitability of GCs in offspring later in life. In addition, low choline has an opposite effect, again suggesting this maternal diet has adverse effects.

IIS

Previous research has shown that Tg2576 mice exhibit IIS (Figure 7A) starting at very young ages: 4 weeks (Kam et al. 2016) or 6 weeks (Bezzina et al. 2015). Therefore, we performed video-EEG using cortical (left frontal cortex, right occipital cortex) and hippocampal (left and right) electrodes in WT and Tg2576 mice (Figure 7B-C). Animals were recorded for 24 hr each session so that there was sufficient time to capture major behavioral states (exploration, awake rest, and sleep). Consistent with previous studies (Bezzina et al. 2015, Kam et al. 2016) we observed IIS in Tg2576 mice but not WT littermates.

The high choline diet reduced IIS frequency in Tg2576 animals.

A. 1. Representative example of an IIS (red arrow). IIS were defined as occurring in all four channels (generalized) to distinguish them from focal spikes that can often be artifact. 2-3. The IIS shown in A1 is expanded.

B. 1. Scatter plot of IIS frequency at each age of recording.

1. Means and sem are plotted. The high choline diet group had fewer IIS than the intermediate diet group at ages 1-3 months ($, p<0.05) and the low choline group had less IIS than the intermediate diet at ages 1-4 months (@, p<0.05). The high and low choline diet groups were significantly different at ages 3 and 4 months (*, p<0.05).

As shown in Figure 7, animals that received the intermediate diet had a significantly higher number of IIS in the 24 hr recording periods compared to animals that received the high choline and the low choline diets. A mixed model analysis showed that there was a significant effect of age (F(2,37) = 3.38; p = 0.04) and maternal diet (F(2,36) = 8.12; p = 0.01). At the 1.25-month recording, Tukey post-hoc analyses showed that IIS frequency in animals treated with the intermediate diet was higher compared to animals treated with the low (p = 0.03) or high choline diets (p = 0.04). This also was true for 2 months and 3 months (low choline, high choline, p <0.05). At 4 months, the low choline groups had significantly reduced IIS frequency compared to the intermediate diet (p = 0.009) but not the high choline group (p = 0.98). At 5-6 months, IIS frequencies were not significantly different in the mice fed the maternal diets (all p > 0.05).

As shown previously (Kam et al. 2016), and in Figure 7, at 4-6 months of age there was high variability in IIS frequency between animals. It is possible that IIS frequency becomes more variable because we found that seizures occur in the Tg2576 mice after 2 months of age (Kam et al. 2016), and there was a sharp increase in IIS during the day before and after a seizure (Supplemental Figure 3).

Seizures and premature mortality in mice fed the low choline diet

We found mice fed the low choline diet had greater ΔFosB-ir in GCs and the hilus showed very low NeuN-ir. Therefore we asked whether low choline-treated mice had more seizures than mice fed the other diets. We recorded 8 mice by video-EEG for 5 days each (at 6 months of age) to examine seizures and found 2 mice from the low choline group had seizures (11 seizures over 2 of the 5 days in one mouse, 1 seizure in the other mouse) whereas none of the other mice had seizures (n=0/4, 2 intermediate and 2 high choline, data not shown).

These values are probably an underestimate for the low choline group because many mice in this group appeared to die in a severe seizure prior to 6 months of age (Supplemental Figure 2). Therefore, the survivors at 6 months probably were the subset with few seizures.

The reason that low choline-treated mice appeared to die in a seizure was that they were found in a specific posture in their cage which occurs when a severe seizure leads to death (Supplemental Figure 2). They were found in a prone posture with extended, rigid limbs (Supplemental Figure 2). Regardless of how the mice died, there was greater mortality in the low choline group compared to mice that had been fed the high choline diet (Supplemental Figure 2).

Sex differences

As shown in Figures 2-3, there appeared to be no sex differences in NOL or NOR. For confirmation, we studied NOL further, since that was the task that showed effects of diet. A three-way ANOVA with sex, genotype and task phase as factors showed no significant effect of sex at three months of age (low choline diet, F(1,35) = 0.0001; p = 0.99). Results were the same for the three months-old mice fed the intermediate diet (F(1,22) = 0.062; p = 0.81) and three months-old mice that were fed the high choline diet (F(1,34) = 0.43; p = 0.52). At six months of age there also were no significant effects of sex (low choline diet, F(1,23) = 0.34; p = 0.57; intermediate diet, F(1,16) = 2.49; p = 0.13; high choline diet, F(1,27) = 0.29; p = 0.87).

Regarding IIS, we did not find sex differences (Supplemental Figure 4). A two-way ANOVA with sex and diet as factors showed no effect of sex (F(1,20) =2.48; p = 0.13) or diet (F(1,20) = 3.56; p = 0.07). Note that for this comparison we used the low choline and high choline groups only because sex information was only available for these mice. The results for months 2-3 also showed no effect of sex (2 months: F(1,17) = 0.64; p = 0.43; 3 months: F(1,15) = 0.19; p = 0.67).

Discussion

Summary

This study showed that early life choline supplementation had several beneficial effects in Tg2576 mice. The high choline diet led to reduced IIS frequency, improved behavior in the NOL task, and improved expression of ΔFosB.

We showed for the first time that Tg2576 mice have pathology in the hilus at six months of age in that hilar neurons are deficient in immunostaining with an anti-NeuN antibody. Tg2576 offspring on the high choline diet showed a restoration of hilar NeuN staining. Because hilar neurons play a role in NOL (Bui et al. 2018), it is possible that rescue of hilar neurons by the high choline diet led to improved spatial memory. Hilar neurons have been suggested to contribute to excitability of GCs (Sloviter 1994, Scharfman 1999, Scharfman & Myers 2012, Jinde et al. 2013, Bui et al. 2018, Botterill et al. 2019), so phenotypic rescue of hilar neurons may also have contributed to the ability of the high choline diet to reduce hyperexcitability.

We also found surprising effects of treating mice with relatively low choline in the maternal diet. On the one hand, IIS frequency was reduced. However, these mice showed premature mortality. Furthermore, in WT mice where no impairments in NOL normally occur, WT mice fed the low choline diet were impaired. In Tg2576 mice, the group that had the low maternal choline diet had the highest ΔFosB expression and the lowest NeuN expression of any experimental group. Potential reasons for these surprising effects are discussed below.

Benefits of high choline

The results of this study are consistent with previous reports that prenatal or postnatal choline supplementation improves object memory. One study of iron deficiency showed that choline supplementation improved NOR (Kennedy et al. 2014). Wistar rats treated with a diet high in choline had improved NOR relative to rats that had been fed a diet low in choline (Moreno et al. 2013). In rats that were aged to 24 months, animals that had been fed a high choline diet showed improved NOR, but only in females (Glenn et al. 2008).

Although few studies have examined the effects of MCS on hyperexcitability, our results showing the benefits of high choline are consistent with reports that a methyl-enriched diet (including choline but also betaine, folic acid, vitamin B12, L-methionine and zinc) reduced spike wave discharges in the offspring (Sarkisova et al. 2023) and audiogenic seizure severity (Poletaeva et al. 2014). Perinatal ethanol treatment that increased excitability was mitigated by choline chloride injection from postnatal day 10-30 (Grafe et al. 2022).

To our knowledge our study is the first to show that MCS can exert effects on IIS and ΔFosB. However, prior studies have shown that MCS in rats improved memory after severe seizures. Using a convulsant to induce several hours of severe seizures (status epilepticus, SE), there is usually extensive hippocampal neuronal loss and when tested in the subsequent days and weeks, memory impairment. MCS reduced this impairment following SE in the Morris water maze (Yang et al. 2000, Holmes et al. 2002, Wong-Goodrich et al. 2011).

Our results are consistent with previous studies of mouse models of AD. In the APP/PS1 mouse model, it was shown that life-long choline supplementation, to the level we used (5.0 g/kg) improved memory in aged mice compared to the 1.1 g/kg diet (Velazquez et al. 2019). Postnatal choline supplementation also improved memory in APP/PS1 mice (Wang et al. 2019). Other improvements were also shown in the APP/PS1 mice, such as improved cholinergic expression via upregulation of choline acetyltransferase (Mellott et al. 2017, Velazquez et al. 2019).

Adverse effects of low choline

The effects of the relatively low choline diet were surprising because several adverse effects were found. This is unexpected because 1.1 g/kg choline is not considered very low in several studies of the past, and some publications consider it normal (e.g., Lamoureux et al. 2008). The majority of past studies used rats (Meck et al. 1988, Meck & Williams 1999, 2003, Meck et al. 2007, Lamoureux et al. 2008), however, so there could be a species difference. Also, some of the past studies used the diet for a shorter timespan than we did (e.g., Lamoureux et al. 2008) because of evidence that perinatal choline supplementation exerts its maximal effects during critical periods (Meck & Williams 2003, Meck et al. 2007).

It was also surprising to find high mortality in mice exposed to the low choline diet in early life. Also, the offspring of mothers fed the low choline diet had very high ΔFosB expression in the GCs, suggesting they had chronic seizures. The observation that mice died from seizures is consistent with the ΔFosB data. There also was weak NeuN in the hilus. When WT mice were tested, NOL was impaired in the low choline group. These data suggest that the offspring of mothers fed the low choline diet were unhealthy, possibly leading to or because of seizures.

On the other hand, IIS were reduced in frequency in the same mice. There are several possible explanations for this apparent paradox. The most parsimonious explanation is the mice were unhealthy. If there was disruption of sleep, IIS might be reduced since IIS occur in sleep primarily. However, there is no evidence that MCS influences sleep in humans or laboratory animals. The one study that looked at the sleep/wake cycle in rats did not find effects (Cheng et al. 2008).

A more radical explanation is that the intermediate diet promoted IIS rather than low choline reducing IIS. The idea here is that choline is not critical to IIS because both low and high choline groups had low IIS frequency compared to the intermediate group, at least for 3 months of life. Instead of choline, a constituent of the intermediate diet may have promoted IIS. The reason why the effect of diet waned over time could have been due to the fact that after weaning all animals had the intermediate diet.

More seizures could explain premature mortality in the low choline group. However, why it would decrease IIS is not clear. One reason seizures could cause IIS to be reduced is that seizures in the Tg2576 mice typically started during sleep and woke mice from sleep. The arousal from sleep would interrupt the behavioral state when IIS usually occur as mentioned above. Also, seizures in the Tg2576 mice were followed by a depression of the EEG that would reduce IIS (postictal depression; Supplemental Figure 3).

NeuN

As mentioned above, NeuN is a neuronal nuclear antigen that can be phosphorylated, and when that occurs the antibody to NeuN no longer binds to NeuN. NeuN is phosphorylated in response to oxidative damage, brain injury, and toxicity (Lind et al. 2005). Therefore, it was of interest when we saw that hilar neurons of the dentate gyrus showed reduced NeuN in Tg2576 mice. Hilar neurons are mainly glutamatergic mossy cells and somatostatin (SOM)/neuropeptide Y (NPY)-expressing GABAergic neurons (HIPP cells; (Houser 2007, Scharfman & Myers 2012, Scharfman 2016), and both neuronal types are implicated in spatial memory functions of the GCs, as well as their excitability (Myers & Scharfman 2009, 2011, Scharfman & Myers 2012, Jinde et al. 2013, Scharfman 2016, Raza et al. 2017, Bui et al. 2018, GoodSmith et al. 2019, Li et al. 2021). If damaged by the high intracellular APP/Aβ levels in young Tg2576 mice, the hilar neurons would be expected to show NeuN loss. This idea is consistent with deficits in SOM and NPY-stained cells (Chan-Palay 1987), indicating a vulnerability.

The reason for rescue of dorsal hilar but not ventral hilar NeuN with the high choline diet is unclear. However, there was a trend for the high choline mice to have greater NeuN in ventral tissue sections. The greater sensitivity of dorsal neurons may be related to differences in dorsal and ventral dentate gyrus gene expression patterns, since some of the genes that are differentially expressed could affect vulnerability to insult or injury (Cembrowski et al. 2016, Zhang et al. 2018). Resilience of vulnerable neuronal populations, especially in the hippocampus, is a topic of high interest in AD research.

ΔFosB

Given that seizures are associated with elevated GC expression of ΔFosB (Chen et al. 1997, McClung et al. 2004, Corbett et al. 2017, You et al. 2017), the differences in ΔFosB expression levels observed here could be the result of increased seizures in animals that received the low choline diet. Therefore, we suggest that the high ΔFosB in Tg2576 mice fed the low choline diet reflects more seizures than in the other two diets.

ΔFosB is a transcription factor that is linked to cognition. In the J20 mouse model of AD, elevated ΔFosB in GCs led to reduced cognition, and when ΔFosB was selectively reduced the cognition improved (Corbett et al. 2017). Therefore, the reduction in ΔFosB by the high choline diet was important to show hyperexcitability was reduced and also important because it showed how high choline may benefit cognition.

Additional considerations

Although the low and high choline diets only had differences in choline, the intermediate diet had different amounts of choline as well as other constituents (Supplementary Table 1). Therefore, the only diet comparison with a difference restricted to choline is the low versus high choline diets (Supplemental Table 1). The intermediate diet was useful however, because numerous studies in AD mouse models employ this standard diet. Moreover, it may have promoted IIS as discussed above.

Although we failed to detect sex differences, there may actually have been differences if females had been analyzed according to their stage of the estrous cycle at death. This possibility is raised by prior data showing that rats and mice during diestrus have low excitability in hippocampal slices; at other cycle stages hyperexcitability occurs (Scharfman et al. 2003).

The Tg2576 mouse model is one of many murine models of AD, and as with all models there are inherent limitations. The Tg2576 model of cerebral amyloid overexpression lacks tau pathology and recapitulates familial AD, whereas the majority of AD is sporadic. However, MCS has now demonstrated structural/functional benefits in several AD-relevant models, namely Tg2576 (this report), APP-PS1 (Mellott et al. 2017), and Ts65Dn (Powers et al. 2016, Strupp et al. 2016, Powers et al. 2017, Alldred et al. 2021, Alldred et al. 2023).

eLife Assessment

In this fundamental work, the authors demonstrated that maternal choline supplementation improved spatial memory, reduced hyperexcitability, and restored NeuN expression in a familial Alzheimer's disease mouse model. Interestingly, choline deficiency increased mortality, while paradoxically reduced hyperexcitability. Using behavior, electrophysiological, and histological measures, the authors present solid and convincing evidence supporting the significant role of maternal choline supplementation in protecting hippocampal functions vulnerable to Alzheimer's disease.

Reviewer #1 (Public Review):

Summary:

Chartampila et al. describe the effect of early-life choline supplementation on cognitive functions and epileptic activity in a mouse model of Alzheimer's disease. The cognitive abilities are assessed by the novel object recognition test and the novel object location test, performed in the same cohort of mice at 3 months and 6 months of age. Neuronal loss was tested using NeuN immunoreactivity, and neuronal hyperexcitability was examined using FosB and video-EEG recordings.

Strengths:

The study was designed as a 6-month follow-up, with repeated behavioral and EEG measurements through disease development, providing valuable and interesting findings on AD progression and the effect of early-life choline supplantation. Moreover, the behavioral data that suggest an adverse effect of low choline in WT mice are interesting and important beyond the context of AD.

Weaknesses:

1. The multiple headings and subheadings, focusing on the experimental method rather than the narrative, reduce the readability.

2. Quantification of NeuN and FosB in WT littermates is needed to demonstrate rescue of neuronal death and hyperexcitability by high choline supplementation and also to gain further insights into the adverse effect of low choline on the performance of WT mice in the behavioral test.

3. Quantification of the discrimination ratio of the novel object and novel location tests can facilitate the comparison between the different genotypes and diets.

4. The longitudinal analyses enable the performance of multi-level correlations between the discrimination ratio in NOR and NOL, NeuN and Fos levels, multiple EEG parameters, and premature death. Such analysis can potentially identify biomarkers associated with AD progression. These can be interesting in different choline supplementation, but also in the standard choline diet.

Reviewer #2 (Public Review):

Summary:

The authors demonstrated that maternal choline supplementation (MCS) improved spatial memory, reduced a marker of hyperexcitability/epilepsy (FosB expression), and reduced oxidative stress (as measured by restored NeuN expression) in an Alzheimer's disease mouse model. This multidisciplinary study spanned behavior, EEG, and histological measures and constituted a large amount of work. Overall, the results supported that MCS does have important effects on hippocampal function, which may substantially impact human AD.

Strengths:

The strength of the group was the ability to monitor the incidence of interictal spikes (IIS) over the course of 1.2-6 months in the Tg2576 Alzheimer's disease model, combined with meaningful behavioral and histological measures. The authors were able to demonstrate MCS had protective effects in Tg2576 mice, which was particularly convincing in the hippocampal novel object location task.

Weaknesses:

Although choline deficiency was associated with impaired learning and elevated FosB expression, consistent with increased hyperexcitability, IIS was reduced with both low and high choline diets. Although not necessarily a weakness, it complicates the interpretation and requires further evaluation.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Strengths:

The study was designed as a 6-month follow-up, with repeated behavioral and EEG measurements through disease development, providing valuable and interesting findings on AD progression and the effect of early-life choline supplantation. Moreover, the behavioral data that suggest an adverse effect of low choline in WT mice are interesting and important beyond the context of AD.

Thank you for identifying several strengths.

Weaknesses:

(1) The multiple headings and subheadings, focusing on the experimental method rather than the narrative, reduce the readability.

We have reduced the number of headings.

(2) Quantification of NeuN and FosB in WT littermates is needed to demonstrate rescue of neuronal death and hyperexcitability by high choline supplementation and also to gain further insights into the adverse effect of low choline on the performance of WT mice in the behavioral test.

We agree and have added WT data for the NeuN and ΔFosB analyses. These data are included in the text and figures. For NeuN, the Figure is Figure 6. For ΔFosB it is Figure 7. In brief, the high choline diet restored NeuN and ΔFosB to the levels of WT mice.

Below is Figure 6 and its legend to show the revised presentation of data for NeuN. Afterwards is the revised figure showing data for ΔFosB. After that are the sections of the Results that have been revised.

Choline supplementation improved NeuN immunoreactivity (ir) in hilar cells in Tg2576 animals.

A. Representative images of NeuN-ir staining in the anterior DG of Tg2576 animals. (1) A section from a Tg2576 mouse fed the low choline diet. The area surrounded by a box is expanded below. Red arrows point to NeuN-ir hilar cells. Mol=molecular layer, GCL=granule cell layer, HIL=hilus. Calibration for the top row, 100 µm; for the bottom row, 50 µm. (2) A section from a Tg2576 mouse fed the intermediate diet. Same calibrations as for 1. (3) A section from a Tg2576 mouse fed the high choline diet. Same calibrations as for 1. B. Quantification methods. Representative images demonstrate the thresholding criteria used to quantify NeuN-ir. (1) A NeuN-stained section. The area surrounded by the white box is expanded in the inset (arrow) to show 3 hilar cells. The 2 NeuN-ir cells above threshold are marked by blue arrows. The 1 NeuN-ir cell below threshold is marked by a green arrow. (2) After converting the image to grayscale, the cells above threshold were designated as red. The inset shows that the two cells that were marked by blue arrows are red while the cell below threshold is not. (3) An example of the threshold menu from ImageJ showing the way the threshold was set. Sliders (red circles) were used to move the threshold to the left or right of the histogram of intensity values. The final position of the slider (red arrow) was positioned at the onset of the steep rise of the histogram. C. NeuN-ir in Tg2576 and WT mice. Tg2576 mice had either the low, intermediate, or high choline diet in early life. WT mice were fed the standard diet (intermediate choline). (1) Tg2576 mice treated with the high choline diet had significantly more hilar NeuN-ir cells in the anterior DG compared to Tg2576 mice that had been fed the low choline or intermediate diet. The values for Tg2576 mice that received the high choline diet were not significantly different from WT mice, suggesting that the high choline diet restored NeuN-ir. (2) There was no effect of diet or genotype in the posterior DG, probably because the low choline and intermediate diet did not appear to lower hilar NeuN-ir.

Choline supplementation reduced ∆FosB expression in dorsal GCs of Tg2576 mice.

A. Representative images of ∆FosB staining in GCL of Tg2576 animals from each treatment group. (1) A section from a low choline-treated mouse shows robust ∆FosB-ir in the GCL. Calibration, 100 µm. Sections from intermediate (2) and high choline (3)-treated mice. Same calibration as 1. B. Quantification methods. Representative images demonstrating the thresholding criteria established to quantify ∆FosB. (1) A ∆FosB -stained section shows strongly-stained cells (white arrows). (2) A strict thresholding criteria was used to make only the darkest stained cells red. C. Use of the strict threshold to quantify ∆FosB-ir. (1) Anterior DG. Tg2576 mice treated with the choline supplemented diet had significantly less ∆FosB-ir compared to the Tg2576 mice fed the low or intermediate diets. Tg2576 mice fed the high choline diet were not significantly different from WT mice, suggesting a rescue of ∆FosB-ir. (2) There were no significant differences in ∆FosB-ir in posterior sections. D. Methods are shown using a threshold that was less strict. (1) Some of the stained cells that were included are not as dark as those used for the strict threshold (white arrows). (2) All cells above the less conservative threshold are shown in red. E. Use of the less strict threshold to quantify ∆FosB-ir. (1) Anterior DG. Tg2576 mice that were fed the high choline diet had less ΔFosB-ir pixels than the mice that were fed the other diets. There were no differences from WT mice, suggesting restoration of ∆FosB-ir by choline enrichment in early life. (2) Posterior DG. There were no significant differences between Tg2576 mice fed the 3 diets or WT mice.

Results, Section C1, starting on Line 691:

“To ask if the improvement in NeuN after MCS in Tg256 restored NeuN to WT levels we used WT mice. For this analysis we used a one-way ANOVA with 4 groups: Low choline Tg2576, Intermediate Tg2576, High choline Tg2576, and Intermediate WT (Figure 5C). Tukey-Kramer multiple comparisons tests were used as the post hoc tests. The WT mice were fed the intermediate diet because it is the standard mouse chow, and this group was intended to reflect normal mice. The results showed a significant group difference for anterior DG (F(3,25)=9.20; p=0.0003; Figure 5C1) but not posterior DG (F(3,28)=0.867; p=0.450; Figure 5C2). Regarding the anterior DG, there were more NeuN-ir cells in high choline-treated mice than both low choline (p=0.046) and intermediate choline-treated Tg2576 mice (p=0.003). WT mice had more NeuN-ir cells than Tg2576 mice fed the low (p=0.011) or intermediate diet (p=0.003). Tg2576 mice that were fed the high choline diet were not significantly different from WT (p=0.827).”

Results, Section C2, starting on Line 722:

“There was strong expression of ∆FosB in Tg2576 GCs in mice fed the low choline diet (Figure 7A1). The high choline diet and intermediate diet appeared to show less GCL ΔFosB-ir (Figure 7A2-3). A two-way ANOVA was conducted with the experimental group (Tg2576 low choline diet, Tg2576 intermediate choline diet, Tg2576 high choline diet, WT intermediate choline diet) and location (anterior or posterior) as main factors. There was a significant effect of group (F(3,32)=13.80, p=<0.0001) and location (F(1,32)=8.69, p=0.006). Tukey-Kramer post-hoc tests showed that Tg2576 mice fed the low choline diet had significantly greater ΔFosB-ir than Tg2576 mice fed the high choline diet (p=0.0005) and WT mice (p=0.0007). Tg2576 mice fed the low and intermediate diets were not significantly different (p=0.275). Tg2576 mice fed the high choline diet were not significantly different from WT (p>0.999). There were no differences between groups for the posterior DG (all p>0.05).”

“∆FosB quantification was repeated with a lower threshold to define ∆FosB-ir GCs (see Methods) and results were the same (Figure 7D). Two-way ANOVA showed a significant effect of group (F(3,32)=14.28, p< 0.0001) and location (F(1,32)=7.07, p=0.0122) for anterior DG but not posterior DG (Figure 7D). For anterior sections, Tukey-Kramer post hoc tests showed that low choline mice had greater ΔFosB-ir than high choline mice (p=0.0024) and WT mice (p=0.005) but not Tg2576 mice fed the intermediate diet (p=0.275); Figure 7D1). Mice fed the high choline diet were not significantly different from WT (p=0.993; Figure 7D1). These data suggest that high choline in the diet early in life can reduce neuronal activity of GCs in offspring later in life. In addition, low choline has an opposite effect, suggesting low choline in early life has adverse effects.”

(3) Quantification of the discrimination ratio of the novel object and novel location tests can facilitate the comparison between the different genotypes and diets.

We have added the discrimination index for novel object location to the paper. The data are in a new figure: Figure 3. In brief, the results for discrimination index are the same as the results done originally, based on the analysis of percent of time exploring the novel object.

Below is the new Figure and legend, followed by the new text in the Results.

Novel object location results based on the discrimination index.

A. Results are shown for the 3 months-old WT and Tg2576 mice based on the discrimination index. (1) Mice fed the low choline diet showed object location memory only in WT. (2) Mice fed the intermediate diet showed object location memory only in WT. (3) Mice fed the high choline diet showed memory both for WT and Tg2576 mice. Therefore, the high choline diet improved memory in Tg2576 mice. B. The results for the 6 months-old mice are shown. (1-2) There was no significant memory demonstrated by mice that were fed either the low or intermediate choline diet. (3) Mice fed a diet enriched in choline showed memory whether they were WT or Tg2576 mice. Therefore, choline enrichment improved memory in all mice.

Results, Section B1, starting on line 536:

“The discrimination indices are shown in Figure 3 and results led to the same conclusions as the analyses in Figure 2. For the 3 months-old mice (Figure 3A), the low choline group did not show the ability to perform the task for WT or Tg2576 mice. Thus, a two-way ANOVA showed no effect of genotype (F(1,74)=0.027, p=0.870) or task phase (F(1,74)=1.41, p=0.239). For the intermediate diet-treated mice, there was no effect of genotype (F(1,50)=0.3.52, p=0.067) but there was an effect of task phase (F(1,50)=8.33, p=0.006). WT mice showed a greater discrimination index during testing relative to training (p=0.019) but Tg2576 mice did not (p=0.664). Therefore, Tg2576 mice fed the intermediate diet were impaired. In contrast, high choline-treated mice performed well. There was a main effect of task phase (F(1,68)=39.61, p=<0.001) with WT (p<0.0001) and Tg2576 mice (p=0.0002) showing preference for the moved object in the test phase. Interestingly, there was a main effect of genotype (F(1,68)=4.50, p=0.038) because the discrimination index for WT training was significantly different from Tg2576 testing (p<0.0001) and Tg2576 training was significantly different from WT testing (p=0.0003).”

“The discrimination indices of 6 months-old mice led to the same conclusions as the results in Figure 2. There was no evidence of discrimination in low choline-treated mice by two-way ANOVA (no effect of genotype, (F(1,42)=3.25, p=0.079; no effect of task phase, F(1,42)=0.278, p=0.601). The same was true of mice fed the intermediate diet (genotype, F(1,12)=1.44, p=0.253; task phase, F(1,12)=2.64, p=0.130). However, both WT and Tg2576 mice performed well after being fed the high choline diet (effect of task phase, (F(1,52)=58.75, p=0.0001, but not genotype (F(1,52)=1.197, p=0.279). Tukey-Kramer post-hoc tests showed that both WT (p<0.0001) and Tg2576 mice that had received the high choline diet (p=0.0005) had elevated discrimination indices for the test session.”

(4) The longitudinal analyses enable the performance of multi-level correlations between the discrimination ratio in NOR and NOL, NeuN and Fos levels, multiple EEG parameters, and premature death. Such analysis can potentially identify biomarkers associated with AD progression. These can be interesting in different choline supplementation, but also in the standard choline diet.

We agree and added correlations to the paper in a new figure (Figure 9). Below is Figure 9 and its legend. Afterwards is the new Results section.

Correlations between IIS, Behavior, and hilar NeuN-ir.

A. IIS frequency over 24 hrs is plotted against the preference for the novel object in the test phase of NOL. A greater preference is reflected by a greater percentage of time exploring the novel object. (1) The mice fed the high choline diet (red) showed greater preference for the novel object when IIS were low. These data suggest IIS impaired object location memory in the high choline-treated mice. The low choline-treated mice had very weak preference and very few IIS, potentially explaining the lack of correlation in these mice. (2) There were no significant correlations for IIS and NOR. However, there were only 4 mice for the high choline group, which is a limitation. B. IIS frequency over 24 hrs is plotted against the number of dorsal hilar cells expressing NeuN. The dorsal hilus was used because there was no effect of diet on the posterior hilus. (1) Hilar NeuN-ir is plotted against the preference for the novel object in the test phase of NOL. There were no significant correlations. (2) Hilar NeuN-ir was greater for mice that had better performance in NOR, both for the low choline (blue) and high choline (red) groups. These data support the idea that hilar cells contribute to object recognition (Kesner et al. 2015; Botterill et al. 2021; GoodSmith et al. 2022).

Results, Section F, starting on Line 801:

“F. Correlations between IIS and other measurements

As shown in Figure 9A, IIS were correlated to behavioral performance in some conditions. For these correlations, only mice that were fed the low and high choline diets were included because mice that were fed the intermediate diet did not have sufficient EEG recordings in the same mouse where behavior was studied. IIS frequency over 24 hrs was plotted against the preference for the novel object in the test phase (Figure 9A). For NOL, IIS were significantly less frequent when behavior was the best, but only for the high choline-treated mice (Pearson’s r, p=0.022). In the low choline group, behavioral performance was poor regardless of IIS frequency (Pearson’s r, p=0.933; Figure 9A1). For NOR, there were no significant correlations (low choliNe, p=0.202; high choline, p=0.680) but few mice were tested in the high choline-treated mice (Figure 9B2).

We also tested whether there were correlations between dorsal hilar NeuN-ir cell numbers and IIS frequency. In Figure 9B, IIS frequency over 24 hrs was plotted against the number of dorsal hilar cells expressing NeuN. The dorsal hilus was used because there was no effect of diet on the posterior hilus. For NOL, there was no significant correlation (low choline, p=0.273; high choline, p=0.159; Figure 9B1). However, for NOR, there were more NeuN-ir hilar cells when the behavioral performance was strongest (low choline, p=0.024; high choline, p=0.016; Figure 9B2). These data support prior studies showing that hilar cells, especially mossy cells (the majority of hilar neurons), contribute to object recognition (Botterill et al. 2021; GoodSmith et al. 2022).”

We also noted that all mice were not possible to include because they died or other reasons, such a a loss of the headset (Results, Section A, Lines 463-464): Some mice were not possible to include in all assays either because they died before reaching 6 months or for other reasons.

Reviewer #2 (Public Review):

Strengths:

The strength of the group was the ability to monitor the incidence of interictal spikes (IIS) over the course of 1.2-6 months in the Tg2576 Alzheimer's disease model, combined with meaningful behavioral and histological measures. The authors were able to demonstrate MCS had protective effects in Tg2576 mice, which was particularly convincing in the hippocampal novel object location task.

We thank the Reviewer for identifying several strengths.

Weaknesses:

Although choline deficiency was associated with impaired learning and elevated FosB expression, consistent with increased hyperexcitability, IIS was reduced with both low and high choline diets. Although not necessarily a weakness, it complicates the interpretation and requires further evaluation.

We agree and we revised the paper to address the evaluations that were suggested.

Reviewer #1 (Recommendations For The Authors):

(1) A reference directing to genotyping of Tg2576 mice is missing.

We apologize for the oversight and added that the mice were genotyped by the New York University Mouse Genotyping core facility.

Methods, Section A, Lines 210-211: “Genotypes were determined by the New York University Mouse Genotyping Core facility using a protocol to detect APP695.”

(2) Which software was used to track the mice in the behavioral tests?

We manually reviewed videos. This has been clarified in the revised manuscript. Methods, Section B4, Lines 268-270: Videos of the training and testing sessions were analyzed manually. A subset of data was analyzed by two independent blinded investigators and they were in agreement.

(3) Unexpectedly, a low choline diet in AD mice was associated with reduced frequency of interictal spikes yet increased mortality and spontaneous seizures. The authors attribute this to postictal suppression.

We did not intend to suggest that postictal depression was the only cause. It was a suggestion for one of many potential explanations why seizures would influence IIS frequency. For postictal depression, we suggested that postictal depression could transiently reduce IIS. We have clarified the text so this is clear (Discussion, starting on Line 960):

If mice were unhealthy, IIS might have been reduced due to impaired excitatory synaptic function. Another reason for reduced IIS is that the mice that had the low choline diet had seizures which interrupted REM sleep. Thus, seizures in Tg2576 mice typically started in sleep. Less REM sleep would reduce IIS because IIS occur primarily in REM. Also, seizures in the Tg2576 mice were followed by a depression of the EEG (postictal depression; Supplemental Figure 3) that would transiently reduce IIS. A different, radical explanation is that the intermediate diet promoted IIS rather than low choline reducing IIS. Instead of choline, a constituent of the intermediate diet may have promoted IIS.

However, reduced spike frequency is already evident at 5 weeks of age, a time point with a low occurrence of premature death. A more comprehensive analysis of EEG background activity may provide additional information if the epileptic activity is indeed reduced at this age.

We did not intend to suggest that premature death caused reduced spike frequency. We have clarified the paper accordingly. We agree that a more in-depth EEG analysis would be useful but is beyond the scope of the study.

(4) Supplementary Fig. 3 depicts far more spikes / 24 h compared to Fig. 7B (at least 100 spikes/24h in Supplementary Fig. 3 and less than 10 spikes/24h in Fig. 7B).

We would like to clarify that before and after a seizure the spike frequency is unusually high. Therefore, there are far more spikes than prior figures.

We clarified this issue by adding to the Supplemental Figure more data. The additional data are from mice without a seizure, showing their spikes are low in frequency.

All recordings lasted several days. We included the data from mice with a seizure on one of the days and mice without any seizures. For mice with a seizure, we graphed IIS frequency for the day before, the day of the seizure, and the day after. For mice without a seizure, IIS frequency is plotted for 3 consecutive days. When there was a seizure, the day before and after showed high numbers of spikes. When there was no seizure on any of the 3 days, spikes were infrequent on all days.

The revised figure and legend are shown below. It is Supplemental Figure 4 in the revised submission.

IIS frequency before and after seizures.

A. Representative EEG traces recorded from electrodes implanted in the skull over the left frontal cortex, right occipital cortex, left hippocampus (Hippo) and right hippocampus during a spontaneous seizure in a 5 months-old Tg2576 mouse. Arrows point to the start (green arrow) and end of the seizure (red arrow), and postictal depression (blue arrow). B. IIS frequency was quantified from continuous video-EEG for mice that had a spontaneous seizure during the recording period and mice that did not. IIS frequency is plotted for 3 consecutive days, starting with the day before the seizure (designated as day 1), and ending with the day after the seizure (day 3). A two-way RMANOVA was conducted with the day and group (mice with or without a seizure) as main factors. There was a significant effect of day (F(2,4)=46.95, p=0.002) and group (seizure vs no seizure; F(1,2)=46.01, p=0.021) and an interaction of factors (F(2,4)=46.68, p=0.002)..Tukey-Kramer post-hoc tests showed that mice with a seizure had significantly greater IIS frequencies than mice without a seizure for every day (day 1, p=0.0005; day 2, p=0.0001; day 3, p=0.0014). For mice with a seizure, IIS frequency was higher on the day of the seizure than the day before (p=0.037) or after (p=0.010). For mice without a seizure, there were no significant differences in IIS frequency for day 1, 2, or 3. These data are similar to prior work showing that from one day to the next mice without seizures have similar IIS frequencies (Kam et al., 2016).

In the text, the revised section is in the Results, Section C, starting on Line 772:

“At 5-6 months, IIS frequencies were not significantly different in the mice fed the different diets (all p>0.05), probably because IIS frequency becomes increasingly variable with age (Kam et al. 2016). One source of variability is seizures, because there was a sharp increase in IIS during the day before and after a seizure (Supplemental Figure 4). Another reason that the diets failed to show differences was that the IIS frequency generally declined at 5-6 months. This can be appreciated in Figure 8B and Supplemental Figure 6B. These data are consistent with prior studies of Tg2576 mice where IIS increased from 1 to 3 months but then waxed and waned afterwards (Kam et al., 2016).”

(5) The data indicating the protective effect of high choline supplementation are valuable, yet some of the claims are not completely supported by the data, mainly as the analysis of littermate WT mice is not complete.

We added WT data to show that the high choline diet restored cell loss and ΔFosB expression to WT levels. These data strengthen the argument that the high choline diet was valuable. See the response to Reviewer #1, Public Review Point #2.

• Line 591: "The results suggest that choline enrichment protected hilar neurons from NeuN loss in Tg2576 mice." A comparison to NeuN expression in WT mice is needed to make this statement.

These data have been added. See the response to Reviewer #1, Public Review Point #2.

• Line 623: "These data suggest that high choline in the diet early in life can reduce hyperexcitability of GCs in offspring later in life. In addition, low choline has an opposite effect, again suggesting this maternal diet has adverse effects." Also here, FosB quantification in WT mice is needed.

These data have been added. See the response to Reviewer #1, Public Review Point #2.

(7) Was the effect of choline associated with reduced tauopathy or A levels?

The mice have no detectable hyperphosphorylated tau. The mice do have intracellular A before 6 months. This is especially the case in hilar neurons, but GCs have little (Criscuolo et al., eNeuro, 2023). However, in neurons that have reduced NeuN, we found previously that antibodies generally do not work well. We think it is because the neurons become pyknotic (Duffy et al., 2015), a condition associated with oxidative stress which causes antigens like NeuN to change conformation due to phosphorylation. Therefore, we did not conduct a comparison of hilar neurons across the different diets.

(8) Since the mice were tested at 3 months and 6 months, it would be interesting to see the behavioral difference per mouse and the correlation with EEG recording and immunohistological analyses.

We agree that would be valuable and this has been added to the paper. Please see response to Reviewer #1, Public Review Point #4.

Reviewer #2 (Recommendations For The Authors):

There were several areas that could be further improved, particularly in the areas of data analysis (particularly with images and supplemental figures), figure presentation, and mechanistic speculation.

Major points:

(1) It is understandable that, for the sake of labor and expense, WT mice were not implanted with EEG electrodes, particularly since previous work showed that WT mice have no IIS (Kam et al. 2016). However, from a standpoint of full factorial experimental design, there are several flaws - purists would argue are fatal flaws. First, the lack of WT groups creates underpowered and imbalanced groups, constraining statistical comparisons and likely reducing the significance of the results. Also, it is an assumption that diet does not influence IIS in WT mice. Secondly, with a within-subject experimental design (as described in Fig. 1A), 6-month-old mice are not naïve if they have previously been tested at 3 months. Such an experimental design may reduce effect size compared to non-naïve mice. These caveats should be included in the Discussion. It is likely that these caveats reduce effect size and that the actual statistical significance, were the experimental design perfect, would be higher overall.

We agree and have added these points to the Limitations section of the Discussion. Starting on Line 1050: In addition, groups were not exactly matched. Although WT mice do not have IIS, a WT group for each of the Tg2576 groups would have been useful. Instead, we included WT mice for the behavioral tasks and some of the anatomical assays. Related to this point is that several mice died during the long-term EEG monitoring of IIS.

(2) Since behavior, EEG, NeuN and FosB experiments seem to be done on every Tg2576 animal, it seems that there are missed opportunities to correlate behavior/EEG and histology on a per-mouse basis. For example, rather than speculate in the discussion, why not (for example) directly examine relationships between IIS/24 hours and FosB expression?

We addressed this point above in responding to Reviewer #1, Public Review Point #4.

(3) Methods of image quantification should be improved. Background subtraction should be considered in the analysis workflow (see Fig. 5C and Fig. 6C background). It would be helpful to have a Methods figure illustrating intermediate processing steps for both NeuN and FosB expression.

We added more information to improve the methods of quantification. We did use a background subtraction approach where ImageJ provides a histogram of intensity values, and it determines when there is a sharp rise in staining relative to background. That point is where we set threshold. We think it is a procedure that has the least subjectivity.

We added these methods to the Methods section and expanded the first figure about image quantification, Figure 6B. That figure and legend are shown above in response to Reviewer #1, Point #2.

This is the revised section of the Methods, Section C3, starting on Line 345:

“Photomicrographs were acquired using ImagePro Plus V7.0 (Media Cybernetics) and a digital camera (Model RET 2000R-F-CLR-12, Q-Imaging). NeuN and ∆FosB staining were quantified from micrographs using ImageJ (V1.44, National Institutes of Health). All images were first converted to grayscale and in each section, the hilus was traced, defined by zone 4 of Amaral (1978). A threshold was then calculated to identify the NeuN-stained cell bodies but not background. Then NeuN-stained cell bodies in the hilus were quantified manually. Note that the threshold was defined in ImageJ using the distribution of intensities in the micrograph. A threshold was then set using a slider in the histogram provided by Image J. The slider was pushed from the low level of staining (similar to background) to the location where staining intensity made a sharp rise, reflecting stained cells. Cells with labeling that was above threshold were counted.”

(4) This reviewer is surprised that the authors do not speculate more about ACh-related mechanisms. For example, choline deficiency would likely reduce Ach release, which could have the same effect on IIS as muscarinic antagonism (Kam et al. 2016), and could potentially explain the paradoxical effects of a low choline diet on reducing IIS. Some additional mechanistic speculation would be helpful in the Discussion.

We thank the Reviewer for noting this so we could add it to the Discussion. We had not because we were concerned about space limitations.

The Discussion has a new section starting on Line 1009:

“Choline and cholinergic neurons

There are many suggestions for the mechanisms that allow MCS to improve health of the offspring. One hypothesis that we are interested in is that MCS improves outcomes by reducing IIS. Reducing IIS would potentially reduce hyperactivity, which is significant because hyperactivity can increase release of A. IIS would also be likely to disrupt sleep since it represents aberrant synchronous activity over widespread brain regions. The disruption to sleep could impair memory consolidation, since it is a notable function of sleep (Graves et al. 2001; Poe et al. 2010). Sleep disruption also has other negative consequences such as impairing normal clearance of A (Nedergaard and Goldman 2020). In patients, IIS and similar events, IEDs, are correlated with memory impairment (Vossel et al. 2016).

How would choline supplementation in early life reduce IIS of the offspring? It may do so by making BFCNs more resilient. That is significant because BFCN abnormalities appear to cause IIS. Thus, the cholinergic antagonist atropine reduced IIS in vivo in Tg2576 mice. Selective silencing of BFCNs reduced IIS also. Atropine also reduced elevated synaptic activity of GCs in young Tg2576 mice in vitro. These studies are consistent with the idea that early in AD there is elevated cholinergic activity (DeKosky et al. 2002; Ikonomovic et al. 2003; Kelley et al. 2014; Mufson et al. 2015; Kelley et al. 2016), while later in life there is degeneration. Indeed, the chronic overactivity could cause the degeneration.

Why would MCS make BFCNs resilient? There are several possibilities that have been explored, based on genes upregulated by MCS. One attractive hypothesis is that neurotrophic support for BFCNs is retained after MCS but in aging and AD it declines (Gautier et al. 2023). The neurotrophins, notably nerve growth factor (NGF) and brain-derived neurotrophic factor (BDNF) support the health of BFCNs (Mufson et al. 2003; Niewiadomska et al. 2011).”

Minor points:

(1) The vendor is Dyets Inc., not Dyets.

Thank you. This correction has been made.

(2) Anesthesia chamber not specified (make, model, company).

We have added this information to the Methods, Section D1, starting on Line 375: The animals were anesthetized by isoflurane inhalation (3% isoflurane. 2% oxygen for induction) in a rectangular transparent plexiglas chamber (18 cm long x 10 cm wide x 8 cm high) made in-house.

(3) It is not clear whether software was used for the detection of behavior. Was position tracking software used or did blind observers individually score metrics?

We have added the information to the paper. Please see the response to Reviewer #1, Recommendations for Authors, Point #2.

(4) It is not clear why rat cages and not a true Open Field Maze were used for NOL and NOR.

We used mouse cages because in our experience that is what is ideal to detect impairments in Tg2576 mice at young ages. We think it is why we have been so successful in identifying NOL impairments in young mice. Before our work, most investigators thought behavior only became impaired later. We would like to add that, in our experience, an Open Field Maze is not the most common cage that is used.

(5) Figure 1A is not mentioned.

It had been mentioned in the Introduction. Figure B-D was the first Figure mentioned in the Results so that is why it might have been missed. We now have added it to the first section of the Results, Line 457, so it is easier to find.

1. Although Fig 7 results are somewhat complicated compared to Fig. 5 and 6 results, EEG comes chronologically earlier than NeuN and FosB expression experiments.

We have kept the order as is because as the Reviewer said, the EEG is complex. For readability, we have kept the EEG results last.

(7) Though the statistical analysis involved parametric and nonparametric tests, It is not clear which normality tests were used.

We have added the name of the normality tests in the Methods, Section E, Line 443: Tests for normality (Shapiro-Wilk) and homogeneity of variance (Bartlett’s test) were used to determine if parametric statistics could be used. We also added after this sentence clarification: When data were not normal, non-parametric data were used. When there was significant heteroscedasticity of variance, data were log transformed. If log transformation did not resolve the heteroscedasticity, non-parametric statistics were used. Because we added correlations and analysis of survival curves, we also added the following (starting on Line 451): For correlations, Pearson’s r was calculated. To compare survival curves, a Log rank (Mantel-Cox) test was performed.

Figures:

(1) In Fig. 1A, Anatomy should be placed above the line.

We changed the figure so that the word “Anatomy” is now aligned, and the arrow that was angled is no longer needed.

In Fig. 1C and 1D, the objects seem to be moved into the cage, not the mice. This schematic does not accurately reflect the Fig. 1C and 1D figure legend text.

Thank you for the excellent point. The figure has been revised. We also updated it to show the objects more accurately.

Please correct the punctuation in the Fig. 1D legend.

Thank you for mentioning the errors. We corrected the legend.

For ease of understanding, Fig. 1C and 1D should have training and testing labeled in the figure.

Thank you for the suggestion. We have revised the figure as suggested.

(2) In Figure 2, error bars for population stats (bar graphs) are not obvious or missing. Same for Figure 3.

We added two supplemental figures to show error bars, because adding the error bars to the existing figures made the symbols, colors, connecting lines and error bars hard to distinguish. For novel object location (Fig. 2) the error bars are shown in Supp. Fig. 2. For novel object recognition, the error bars are shown in Supplemental Fig. 3.

(3) The authors should consider a Methods figure for quantification of NeuN and deltaFOSB (expansions of Fig. 5C and Fig. 6C).

Please see Reviewer #1, Public Review Point #2.

(4) In Figure 5, A should be omitted and mentioned in the Methods/figure legend. B should be enlarged. C should be inset, zoomed-in images of the hilus, with an accompanying analysis image showing a clear reduction in NeuN intensity in low choline conditions compared to intermediate and high choline conditions. In D, X axes could delineate conditions (figure legend and color unnecessary). Figure 5C should be moved to a Methods figure.

We thank the review for the excellent suggestions. We removed A as suggested. We expanded B and included insets. We used different images to show a more obvious reduction of cells for the low choline group. We expanded the Methods schematics. The revised figure is Figure 6 and shown above in response to Reviewer 1, Public Review Point #2.

(5) In Figure 6, A should be eliminated and mentioned in the Methods/figure legend. B should be greatly expanded with higher and lower thresholds shown on subsequent panels (3x3 design).

We removed A as suggested. We expanded B as suggested. The higher and lower thresholds are shown in C. The revised figure is Figure 7 and shown above in response to Reviewer 1, Public Review Point #2.

(6) In Figure 7, A2 should be expanded vertically. A3 should be expanded both vertically and horizontally. B 1 and 2 should be increased, particularly B1 where it is difficult to see symbols. Perhaps colored symbols offset/staggered per group so that the spread per group is clearer.

We added a panel (A4) to show an expansion of A2 and A3. However, we did not see that a vertical expansion would add information so we opted not to add that. We expanded B1 as suggested but opted not to expand B2 because we did not think it would enhance clarity. The revised figure is below.

(7) Supplemental Figure 1 could possibly be combined with Figure 1 (use rounded corner rat cage schematic for continuity).

We opted not to combine figures because it would make one extremely large figure. As a result, the parts of the figure would be small and difficult to see.

(8) Supplemental Figure 2 - there does not seem to be any statistical analysis associated with A mentioned in the Results text.

We added the statistical information. It is now Supplemental Figure 4:

Mortality was high in mice treated with the low choline diet.

A. Survival curves are shown for mice fed the low choline diet and mice fed the high choline diet. The mice fed the high choline diet had a significantly less severe survival curve. B. Left: A photo of a mouse after sudden unexplained death. The mouse was found in a posture consistent with death during a convulsive seizure. The area surrounded by the red box is expanded below to show the outstretched hindlimb (red arrow). Right: A photo of a mouse that did not die suddenly. The area surrounded by the box is expanded below to show that the hindlimb is not outstretched.

The revised text is in the Results, Section E, starting on Line 793:

“The reason that low choline-treated mice appeared to die in a seizure was that they were found in a specific posture in their cage which occurs when a severe seizure leads to death (Supplemental Figure 5). They were found in a prone posture with extended, rigid limbs (Supplemental Figure 5). Regardless of how the mice died, there was greater mortality in the low choline group compared to mice that had been fed the high choline diet (Log-rank (Mantel-Cox) test, Chi square 5.36, df 1, p=0.021; Supplemental Figure 5A).”

Also, why isn't intermediate choline also shown?

We do not have the data from the animals. Records of death were not kept, regrettably.

Perhaps labeling of male/female could also be done as part of this graph.

We agree this would be very interesting but do not have all sex information.

B is not very convincing, though it is understandable once one reads about posture.

We have clarified the text and figure, as well as the legend. They are above.

Are there additional animals that were seen to be in a specific posture?

There are many examples, and we added them to hopefully make it more convincing.

We also added posture in WT mice when there is a death to show how different it is.

Is there any relationship between seizures detected via EEG, as shown in Supplemental Figure 3, and death?

Several mice died during a convulsive seizure, which is the type of seizure that is shown in the Supplemental Figure.

(9) Supplemental Figure 3 seems to display an isolated case in which EEG-detected seizures correlate with increased IIEs. It is not clear whether there are additional documented cases of seizures that could be assembled into a meaningful population graph. If this data does not exist or is too much work to include in this manuscript, perhaps it can be saved for a future paper.

We have added other cases and revised the graph. This is now Supplemental Figure 4 and is shown above in response to Reviewer #1, Recommendation for Authors Point #4.

Frontal is misspelled.

We checked and our copy is not showing a misspelling. However, we are very grateful to the Reviewer for catching many errors and reading the manuscript carefully.

(10) Supplemental Figure 4 seems incomplete in that it does not include EEG data from months 4, 5, and 6 (see Fig. 7B).

We have added data for these ages to the Supplemental Figure (currently Supplemental Figure 6) as part B. In part A, which had been the original figure, only 1.2, 2, and 3 months-old mice were shown because there were insufficient numbers of each sex at other ages. However, by pooling 1.2 and 2 months (Supplemental Figure 6B1), 3 and 4 months (B2) and 5 and 6 months (B3) we could do the analysis of sex. The results are the same – we detected no sex differences.

Author response image 9.

A. IIS frequency was similar for each sex. A. IIS frequency was compared for females and males at 1.2 months (1), 2 months (2), and 3 months (3). Two-way ANOVA was used to analyze the effects of sex and diet. Female and male Tg2576 mice were not significantly different. B. Mice were pooled at 1.2 and 2 months (1), 3 and 4 months (2) and 5 and 6 months (3). Two-way ANOVA analyzed the effects of sex and diet. There were significant effects of diet for (1) and (2) but not (3). There were no effects of sex at any age.

(1) There were significant effects of diet (F(2,47)=46.21, p<0.0001) but not sex (F(1,47)=0.106, p=0.746). Female and male mice fed the low choline diet or high choline diet were significantly different from female and male mice fed the intermediate diet (all p<0.05, asterisk).

(2) There were significant effects of diet (F(2,32)=10.82, p=0.0003) but not sex (F(1,32)=1.05, p=0.313). Both female and male mice of the low choline group were significantly different from male mice fed the intermediate diet (both p<0.05, asterisk) but no other pairwise comparisons were significant.

(3) There were no significant differences (diet, F(2,23)=1.21, p=0.317); sex, F(1,23)=0.844, p=0.368).

The data are discussed the Results, Section G, tarting on Line 843:

In Supplemental Figure 6B we grouped mice at 1-2 months, 3-4 months and 5-6 months so that there were sufficient females and males to compare each diet. A two-way ANOVA with diet and sex as factors showed a significant effect of diet (F(2,47)=46.21; p<0.0001) at 1-2 months of age, but not sex (F1,47)=0.11, p=0.758). Post-hoc comparisons showed that the low choline group had fewer IIS than the intermediate group, and the same was true for the high choline-treated mice. Thus, female mice fed the low choline diet differed from the females (p<0.0001) and males (p<0.0001) fed the intermediate diet. Male mice that had received the low choline diet different from females (p<0.0001) and males (p<0.0001) fed the intermediate diet. Female mice fed the high choline diet different from females (p=0.002) and males (p<0.0001) fed the intermediate diet, and males fed the high choline diet difference from females (p<0.0001) and males (p<0.0001) fed the intermediate diet.

For the 3-4 months-old mice there was also a significant effect of diet (F(2,32)=10.82, p=0.0003) but not sex (F(1,32)=1.05, p=0.313). Post-hoc tests showed that low choline females were different from males fed the intermediate diet (p=0.007), and low choline males were also significantly different from males that had received the intermediate diet (p=0.006). There were no significant effects of diet (F(2,23)=1.21, p=0.317) or sex (F(1,23)=0.84, p=0.368) at 5-6 months of age.

Version 2

Abstract

Abstract

Maternal choline supplementation (MCS) improves cognition in Alzheimer’s disease (AD) models. However, effects of MCS on neuronal hyperexcitability in AD are unknown. We investigated effects of MCS in a well-established mouse model of AD with hyperexcitability, the Tg2576 mouse. The most common type of hyperexcitability in Tg2576 mice, and many other mouse models and AD patients, are generalized EEG spikes (interictal spikes; IIS). Hyperexcitability is also reflected by elevated expression of the transcription factor ΔFosB in the granule cells (GCs) of the dentate gyrus (DG), which are the principal cell type. We also studied the hilus of the DG because hilar neurons regulate GC excitability. We found reduced expression of the neuronal marker NeuN within hilar neurons in Tg2576 mice, which other studies have shown is a sign of oxidative stress or other pathology.

Tg2576 breeding pairs received a diet with a relatively low, intermediate or high concentration of choline. After weaning, all mice received the intermediate diet. In offspring of mice fed the high choline diet, IIS frequency declined, GC ΔFosB expression was reduced, and NeuN expression was restored. Spatial memory improved using the novel object location task. In contrast, offspring exposed to the relatively low choline diet had several adverse effects, such as increased mortality. They had the weakest hilar NeuN immunoreactivity and greatest GC ΔFosB. However, their IIS frequency was low, which was surprising. The results provide new evidence that a diet high in choline in early life can improve outcomes in a mouse model of AD, and relatively low choline can have mixed effects. This is the first study showing that dietary choline can regulate hyperexcitability, hilar neurons, ΔFosB and spatial memory in an animal model of AD.

Introduction

Diet has been suggested to influence several aspects of brain health. One dietary intervention that has been studied extensively in rodents and humans is supplementation of the maternal diet with the nutrient choline (maternal choline supplementation; MCS). MCS improves many aspects of brain health in humans (Zeisel and da Costa 2009; Jiang et al. 2014). In normal rats, MCS also is beneficial, with numerous studies showing improved behavior in the adult offspring (for review see Meck and Williams 2003).

In Alzheimer’s disease (AD), changes to the diet have been recommended (Bourre 2006; Power et al. 2019; Mao et al. 2021; Lobo et al. 2022), including MCS (Strupp et al. 2016; Velazquez et al. 2020; Dave et al. 2023; Judd et al. 2023a; b). One reason for the recommendation is that serum levels of choline are low in individuals with AD (Dave et al. 2023; Judd et al. 2023b). In addition, using mouse models of AD, increased choline improved many characteristics of the disease, ranging from inflammation to glucose metabolism and the hallmark amyloid and tau pathology (Dave et al. 2023; Judd et al. 2023a; b). In the Ts65Dn mouse model of Down syndrome (DS) and AD, MCS led to improved memory and attention in Ts65Dn offspring (Strupp et al. 2016; Powers et al. 2017; Powers et al. 2021). In addition, degeneration of basal forebrain cholinergic neurons (BFCNs) in Ts65Dn mice, a hallmark of DS and AD, was reduced (Kelley et al. 2016; Powers et al. 2017; Alldred et al. 2023; Gautier et al. 2023).

Here we asked if MCS would improve Tg2576 mice, a model of familial AD where a mutation in the precursor to amyloid β (Aβ), amyloid precursor protein (APP) is expressed by the hamster prion protein promoter (Hsiao et al. 1996). This is a commonly used mouse model that simulates aspects of AD.

One of the reasons to use Tg2576 mice was to ask if MCS would improve the hyperexcitability found in the mice. The Tg2576 mouse is ideal because hyperexcitability is robust, and BFCNs are likely to play a role in the hyperexcitabililty (Bezzina et al. 2015; Kam et al. 2016; Lisgaras and Scharfman 2023). In our past work, the primary type of hyperexcitability was interictal spikes (IIS), which are named because they occur in between seizures (ictal events) in epilepsy. They are studied by EEG. Importantly, IIS are found in numerous mouse models of AD (J20, Brown et al. 2018; Ts65Dn, PS2-/-, Lisgaras and Scharfman 2023; APP/PS1, Shoob et al. 2023), as well as patients (Sanchez et al. 2012; Vossel et al. 2016; Beagle et al. 2017; Vossel et al. 2017; Vossel and Karageorgiou 2021; Vossel 2023).

We also studied hyperexcitability using ΔFosB expression, a transcription factor that is increased when neurons have been highly active over the prior 10-14 days (McClung et al. 2004). We studied ΔFosB in granule cells (GCs) of the dentate gyrus (DG), because numerous ΔFosB-expressing GCs occur when there are IIS (You et al. 2017; You et al. 2018). We also focused on GCs because we found that the cell layer of GCs (GCL) is where IIS are largest relative to area CA1 and overlying neocortex (Lisgaras and Scharfman 2023). Moreover, closed loop optogenetic silencing of GCs reduced IIS (Lisgaras and Scharfman 2023).

We also asked if behavior would improve if MCS reduced IIS frequency in Tg2576 mice. The basis for this question is in studies of epilepsy, where IIS disrupt cognition (Rausch et al. 1978; Aarts et al. 1984; Holmes and Lenck-Santini 2006; Kleen et al. 2010; Kleen et al. 2013; Gelinas et al. 2016). Also, both IIS and an epileptiform discharge that is usually longer lasting, interictal epileptiform discharges (IEDs), are correlated with cognitive dysfunction in AD patients (Vossel et al. 2016).

We fed dams one of 3 diets, which were relatively low, intermediate or high in choline (Figure 1A; Supplemental Table 1). The high choline diet provided levels of choline similar to other studies of MCS in rodents (Meck et al. 1988; Loy et al. 1991; Holler et al. 1996; Meck and Williams 1999; Sandstrom et al. 2002; Mellott et al. 2004; Glenn et al. 2012; Kelley et al. 2019). After weaning the intermediate diet was used. Offspring were implanted with electrodes for EEG at 1 month of age, and 24 hrs-long recordings began 1/month starting at 1 week after surgery. After 6 months the recordings ended. The time span (1-6 months of age) was selected because this is when cognitive impairment and IIS first develop. Therefore we could determine if these characteristics were delayed.

Schematic of the behavioral procedures for the NOL and NOR tasks.

A. Experimental timeline. Dams were fed one of 3 diets for mating, gestation, and offspring consumed the diet until weaning at 25-30 days of age. At 1 month-old, offspring were implanted with electrodes for EEG and recorded for a 24 hrs-long period at 1.2, 2, 3, 4, 5, and 6 months of age. At 3 and 6 months of age behavior was tested. At 6 months of age, after behavioral testing, mice were perfused, brains were sectioned, and sections were processed with antibodies against NeuN and ΔFosB.

B. Prior to NOL or NOR, animals were acclimated to the testing arena. There were 3 acclimation sessions separated by 24 hrs during which animals were allowed to freely explore for 5 min.

C. In the NOL task, animals were placed in a cage with two identical objects and allowed to freely explore for 5 min (Training). After one hour they were brought back to the cage, where one object was displaced, and allowed to freely explore for 5 min (Testing).

D. In the NOR task. animals were placed in a cage with two identical objects and allowed to freely explore for 5 min (Training). After one hour they were brought back to the cage, where one object was replaced, and allowed to freely explore for 5 min (Testing).

We previously found that the novel object location (NOL) task is impaired at 3-4 months of age in Tg2576 mice (Duffy et al. 2015), so we evaluated NOL at 3 months, and a related task, novel object recognition (NOR; Vogel-Ciernia and Wood 2014). To understand the persistence of any effects of the diets, we repeated NOL and NOR at 6 months. The interval between testing was sufficiently long that it is unlikely that testing at 3 months affected testing at 6 months, but we can not exclude the possibility. At 6 months of age we perfusion-fixed mice, sectioned the brains, and evaluated NeuN and ΔFosB protein expression. NeuN was studied because it is a neuronal marker that is reduced in numerous pathological conditions (Buckingham et al. 2008; Kadriu et al. 2009; Matsuda et al. 2009; Duffy et al. 2011; Duffy et al. 2015). We also assayed ΔFosB because it is a method complementary to EEG recordings of hyperexcitabililty. It also can be used to assess hyperexcitability of GCs.

The results showed a remarkable effect of the high choline diet. IIS and spatial memory were improved, as was NeuN and ΔFosB expression. Interestingly, the relatively low choline diet had mixed effects, reducing IIS frequency, but making NOL, ΔFosB, and NeuN worse. The mice also died prematurely relative to mice that were fed the high choline diet. We also report for the first time that there is loss of NeuN-ir in the DG hilus of Tg2576 mice, which is important because impaired hilar neurons could cause GC hyperexcitability (Sperk et al. 2007; Scharfman and Myers 2012). In summary, we make a strong argument for choline supplementation in early life to improve outcomes in an AD model, especially the DG.

Methods

Animals

All experimental procedures followed the guidelines set by the National Institute of Health (NIH) and were approved by the Institutional Animal Care and Use Committee (IACUC) of the Nathan Kline Institute for Psychiatric Research. Mice were housed in standard mouse cages (26 cm wide x 40 cm long x 20 cm high) with corn cob bedding and a 12 hrs-long light-dark cycle. Food and water were provided ad libitum.

Tg2576 mice express a mutant form of human APP (isoform 695) with a mutation found in a Swedish cohort (Lys670Arg, Met671Leu), driven by the hamster prion protein promoter (Hsiao et al. 1996). Mice were bred in-house from heterozygous Tg2576 males and non-transgenic female mice (C57BL6/SJL F1 hybrid). Wildtype (WT) mice were obtained from Jackson Laboratories (Stock# 100012). Genotypes were determined by the New York University Mouse Genotyping Core facility using a protocol to detect APP695.

Breeding pairs were randomly assigned to receive one of 3 diets with different concentrations of choline chloride: 1.1 g/kg (AIN-76A, Dyets Inc.), 2.0 g/kg (Purina 5008, W.F. Fisher and Son Inc.) and 5.0 g/kg (AIN-76A, Dyets Inc.; Supplemental Table 1). These diets (“low choline” diet, “intermediate choline” diet and “high choline” diet, respectively) were used until weaning (25-30 days of age). After weaning, all mice were fed the intermediate diet. Mice were housed with others of the same sex (1-4 per cage).

Behavior

General information

Mice were housed in the room where they would be tested behaviorally starting at least 24 hrs prior to all experiments. Both NOL and NOR included 3 acclimations (5 min each) followed by a training and a testing session (5 min each; Figure 1B-D). The interval between training and testing was 60 minutes and therefore tested short-term memory (Vogel-Ciernia and Wood 2014). NOL and NOR tests were separated by 7 days to ensure effects of one task did not affect the next. Prior studies have failed to find an effect on one task on the next when a 7 day-long interval is used (Botterill et al. 2021). The order of testing (NOL before NOR or NOR before NOL) was randomized.

Video recordings of all training and testing sessions were captured using a USB camera (Logitech HD Pro C920, Logitech). Acclimations were conducted between 10:00 a.m. and 12:00 p.m., while training and testing was conducted between 1:00 p.m. and 4:00 p.m. All equipment was cleaned using 70% ethanol before each use. In pilot studies, we chose objects that would be explored equally by animals. Indeed, animals showed no preference for either of the objects (i.e., LEGO object and pineapple, Supplemental Figure 1) during training, as reported in the Results.

NOL

NOL was conducted in a standard rat cage (26 cm wide x 40 cm long x 20 cm high) with different pictures on 3 sides of the cage to provide a consistent context and therefore foster acclimation. Pictures included several shapes and colors. The dimensions of the pictures were 1) 10 cm x 21 cm, 2) 16 cm x 18 cm, and 3) 17 cm x 20 cm. During the acclimations, animals were allowed to freely explore the cage. During the training session, mice were placed in the cage where they had been acclimated, with two identical objects, 1 in a corner of the cage (e.g., left top) and the second in the adjacent corner (e.g., left bottom, Supplemental Figure 1). The mice were then removed and placed in their home cage for 1 hr. During the test session, one of the objects was moved to the opposite end of the cage (left top to right top; Supplemental Figure 1). The two objects were identical pineapple-like objects made of metal that were painted and approximately 3 cm x 3 cm x 5 cm (Supplemental Figure 1; Botterill et al. 2021).

NOR

NOR was conducted in the cage described above for NOL. As for NOL, during the acclimations animals were allowed to freely explore the cage. During the training session, mice were placed in the cage used for acclimation with two identical objects centered along the shortest cage wall (Supplemental Figure 1). The mouse was then removed and placed in their home cage for 1 hr. During the testing session, one of the objects was replaced with a different one. The objects that were identical were 2 pineapple-like objects, and the new object was made of 2 green LEGO pieces approximately 5 cm x 7 cm x 7 cm (Supplemental Fig. 1; Botterill et al. 2021).

Quantification

The experimenter who conducted the analysis was blind to the genotype of the offspring and which object was novel or familiar. Videos of the training and testing sessions were analyzed manually. A subset of data was analyzed by two independent blinded investigators and they were in agreement. Exploration was quantified based on the guidelines of Vogel-Ciernia and Wood (2014). However, in addition to time spent exploring, the number of approaches to an object was quantified. Time spent exploring each object was defined as the duration of time the nose was pointed at the object and the nose was located within 2 cm of the object. In addition, exploration time included the time animals spent on top of the object if they were looking down and sniffing it. An approach was defined by a movement toward the object and ending at least 2 cm from the edge of the object.

Animals that remembered the objects that were explored during the training phase were expected to demonstrate an increased preference for novelty in the test phase (Ennaceur and Delacour 1988). In other words, exploration of the novel object during testing was expected to be higher than 50% of the total time of object exploration.

Anatomy

Perfusion-fixation and sectioning

Mice were initially anesthetized by isoflurane inhalation (NDC# 07-893-1389, Patterson Veterinary), followed by urethane (2.5 g/kg; i.p.; Cat# U2500, Sigma-Aldrich). Under deep anesthesia, the abdominal and heart cavities were opened with surgical scissors and a 23g needle inserted into the heart. The atria was clipped, the needle was clamped in place with a hemostat, and the animal was transcardially-perfused using 10 mL of room temperature (RT) saline (0.9% sodium chloride in double distilled (dd) H2O) using a peristaltic pump (Minipuls1; Gilson), followed by 30 mL of 4°C 4% paraformaldehyde (PFA, Cat# 19210, Electron Microscopy Sciences) in 0.1 M phosphate buffer (PB; pH 7.4). The brains were immediately extracted, hemisected, and post-fixed for at least 24 hrs in 4% PFA (at 4°C).

Following post-fixation, 50 µm-thick coronal sections were made using a vibratome (VT 1000P, Leica Biosystems). Sections were collected serially and stored at 4°C in 24-well tissue culture plates containing cryoprotectant solution (30% glycerol, 30% ethylene glycol, 40% 0.1 M phosphate buffer, pH 6.7). Throughout the hippocampus, every 6th section was processed with an antibody to NeuN or ΔFosB. For analysis of anterior DG, 2-3 sections that were 300 µm apart were selected from ∼1.5-2.5 mm posterior to Bregma. For posterior DG, 2-3 sections that were 300 µm apart were selected starting at ∼3.5 mm posterior to Bregma. The values of the 2-3 sections were averaged so that there was one measurement for anterior and one for posterior for each mouse.

NeuN and ΔFosB immunohistochemistry

Free floating sections were washed in 0.1 M Tris Buffer (TB; 3 x 5 min), followed by a 3 min wash in 1% H2O2. Sections were then washed in 0.1 M TB (3 x 5 min) and incubated for 60 min in 5% horse serum for NeuN (Cat# S-2000, Vector) or 5% goat serum for ΔFosB (Cat# S-1000, Vector), diluted in a solution of 0.25% (volume/volume or v/v) Triton-X 100, and 1% (weight/volume or w/v) bovine serum albumin in 0.1 M TB. Sections were then incubated overnight at 4°C in primary antibody to NeuN (mouse monoclonal, 1:5000; Cat# MAB377, Chemicon International) or anti-ΔFosB (rabbit monoclonal, 1:1000; Cat# D3S8R, Cell Signaling), diluted in a solution of 0.25% (v/v) Triton-X 100, and 1% (w/v) bovine serum albumin in 0.1M TB. Both NeuN and ΔFosB have been well-characterized as antigens (Mullen et al. 1992; Wolf et al. 1996; Chen et al. 1997; Sarnat et al. 1998) and the antibodies we used have been commonly employed in the past (Chen et al. 1997; Duffy et al. 2013; Corbett et al. 2017). On the following day, sections were washed in 0.1 M TB (3 x 5 min) and then incubated for 60 min in biotinylated horse anti-mouse secondary antibody for NeuN (1:500, Cat# BA-2000, Vector) or biotinylated goat anti-rabbit secondary antibody for ΔFosB (1:500, Cat# BA-1000), diluted in a solution of 0.25% (v/v) Triton-X 100, and 1% (w/v) bovine serum albumin in 0.1 M TB. The sections were then washed in 0.1 M TB (3 x 5 min) and incubated in avidin-biotin complex for 2 hrs (1:1000; Cat# PK-6100, Vector). They were washed in 0.1 M TB (3 x 5 min) and then reacted in a solution containing 0.5 mg/mL 3, 3’-diaminobenzidine (DAB; Cat# D5905, Sigma-Aldrich), 40 µg/mL ammonium chloride (Cat# A4514, Sigma-Aldrich), 25 mg/mL D(+)-glucose (Cat# G5767, Sigma-Aldrich), and 3 g/mL glucose oxidase (Cat# G2133, Sigma-Aldrich) in 0.1 M TB. This method slowed the reaction time so that the reaction could be temporally controlled, meaning that the reaction could be stopped when the immunoreactivity was robust but background was still low. The sections were then washed in 0.1 M TB (3 x 5 min), mounted on gelatin-coated slides (1% bovine gelatin, Cat# G9391, Sigma-Aldrich) and dried at RT overnight. The following day they were dehydrated in increasing concentrations of ethanol (90%, 10 min; 95% 10 min; 100%, 10 min; 100% again, 10 min), washed in Xylene (10 min; Cat# 534056, Sigma-Aldrich) and cover-slipped with Permount (Cat# 17986-01, Electron Microscopy Sciences).

Analysis

Photomicrographs were acquired using ImagePro Plus V7.0 (Media Cybernetics) and a digital camera (Model RET 2000R-F-CLR-12, Q-Imaging). NeuN and ΔFosB staining were quantified from micrographs using ImageJ (V1.44, National Institutes of Health). All images were first converted to grayscale and in each section, the hilus was traced, defined by zone 4 of Amaral (1978). A threshold was then calculated to identify the NeuN-stained cell bodies but not background. Then NeuN-stained cell bodies in the hilus were quantified manually. Note that the threshold was defined in ImageJ using the distribution of intensities in the micrograph. A threshold was then set using a slider in the histogram provided by Image J. The slider was pushed from the low level of staining (similar to background) to the location where staining intensity made a sharp rise, reflecting stained cells. Cells with labeling that was above threshold were counted.

To quantify ΔFosB-stained cells, images were converted to grayscale and in each section, the GCL was outlined and defined as a region of interest (ROI). A threshold was then set to identify the well-stained cells but not the background, analogous to the method to define threshold in the for NeuN. Two thresholds were used because some cells were lightly stained and others much more robust in their staining. The first threshold was most inclusive of immunoreactive cells. Presumably this group reflected cells with less neuronal activity as well as those with intense activity. Then a second analysis was done with the same sections using a higher threshold. This group corresponded to cells with the highest activity. ImageJ was used to calculate the area (in pixels) within the ROI that was above threshold.

Video-electroencephalographic (video-EEG) recordings

Stereotaxic surgery

EEG electrodes (stainless steel jeweler’s screws) were implanted in Tg2576 animals at 4 weeks of age. WT littermates were not implanted because previous work from our laboratory has shown that WT mice have no IIS (Kam et al. 2016). The animals were anesthetized by isoflurane inhalation (3% isoflurane. 2% oxygen for induction) in a rectangular transparent plexiglas chamber (18 cm long x 10 cm wide x 8 cm high) made in-house. For maintenance during surgery, isoflurane was < 1.75% and flow rate was 1.2 L/min. Mice were placed in a stereotaxic apparatus (David Kopf Instruments). Prior to the implantation of the electrodes, animals were injected with the analgesic Buprenex (0.2 mg/kg, s.c.; buprenorphine hydroxide, NDC# 12496-0757-5, Reckitt Benckiser), which was diluted in saline (0.03 mg/mL in sterile 0.9% sodium chloride solution, Vedco Inc.). The skull was exposed with a midline incision and 6 holes were drilled for the placement of subdural screw electrodes (Cat# 8209, 0.10” stainless steel screws, Pinnacle Technology).

The coordinates for the electrode placement were: right occipital cortex (AP - 3.5 mm, ML 2.0 mm), left frontal cortex (AP -0.5 mm, ML -1.5 mm), left hippocampus (AP -2.5 mm, ML -2.0 mm), and right hippocampus (AP -2.5 mm, ML 2.0 mm (Kam et al. 2016). Two additional screws were placed, one over the right olfactory bulb as ground (AP 2.3 mm, ML 1.8 mm) and another at the midline over the cerebellum as a reference (relative to Lambda: AP -1.5 mm, ML -0.5 mm). Screws were attached to an 8-pin connector (Cat# ED85100-ND, Digi-Key Corporation), which was placed over the skull and secured with dental cement (Cat# 4734FIB, Lang Dental Mfg. Co., Inc.).

After surgery, animals were placed on a heating blanket overnight and injected with lactated Ringer’s solution (50 mL/kg at 31°C; NDC# 099355000476, Aspen Veterinary Resources Ltd). They were then transferred to the room where the video-EEG is recorded. Implanted animals were allowed to recover for one week following the surgery.

Video-EEG recording

Video-EEG was recorded stating at 1.25 months (5 weeks) of age. The animals were then recorded at 2, 3, 4, 5 and 6 months of age. Each recording session lasted 24 hrs so that a long period of sleep could be acquired, since IIS occur primarily in sleep (Kam et al. 2016).

Mice were placed into 21 cm x 19 cm transparent cages with food and water provided ad libitum and corncob bedding. A pre-amplifier was inserted into the 8-pin connector on the skull, which was connected to a multichannel commutator (Cat# 8408, Pinnacle Technology). This arrangement allowed for free range of movement throughout the recording. EEG signals were acquired at 2000 Hz and bandpass filtered at 0.5 – 200 Hz using Sirenia Acquisition )V2.0.4, Pinnacle Technology). Simultaneous video recordings were captured using an infrared camera (Cat# AP-DCS100W, Apex CCTV).

Analysis

EEG recordings were analyzed offline with Neuroscore V3.2.1 (Data Science International). IIS were defined as large amplitude, 10-75 msec deflections occurring synchronously in all 4 leads (Kam et al. 2016). They were quantified by first setting an amplitude threshold (calculated separately for each recording) and a duration criterion (10-75 msec). To determine the amplitude threshold, we calculated the root mean square (RMS) amplitude for noise in each recording during a 60 sec segment of baseline activity that did not include any artifacts. The threshold for IIS was set at 9 standard deviations above the RMS amplitude. This threshold was selected because it identified IIS extremely well and excluded movement or other artifacts. The IIS in one of the hippocampal channels was used because of data suggesting IIS begin in the hippocampus (Kam et al. 2016; Lisgaras and Scharfman 2023). Following the automatic analysis of the IIS, the software’s detection accuracy was verified manually for each recording to ensure that all spikes appeared in all 4 channels and that no artifacts were included in the final IIS count.

Statistical comparisons

Data are expressed as mean ± standard error of the mean (SEM). The significant was set to <0.05 prior to all experiments. Tests were conducted using Prism (V 9.4.1, Graphpad Software).

Parametric data comparing two groups used unpaired two-tailed t-tests. For >2 groups, one-way ANOVA was used followed by Tukey-Kramer post-hoc tests that corrected for multiple comparisons. For data with two main factors, two-way ANOVA was followed by Tukey-Kramer post-hoc tests. Interactions are not reported in the Results unless they were significant. For the analysis of IIS frequency from 1.2 months of age up to 6 months-of age, a repeated measures ANOVA (RMANOVA) was used.

Tests for normality (Shapiro-Wilk) and homogeneity of variance (Bartlett’s test) were used to determine if parametric statistics could be used. When data were not normal, non-parametric data were used. When there was significant heteroscedasticity of variance, data were log transformed. If log transformation did not resolve the heteroscedasticity, non-parametric statistics were used. For non-parametric data, Mann-Whitney U tests were used to compare two groups and Kruskal-Wallis for >2 groups. The post-hoc tests were Dunn’s test for multiple comparisons.

For correlations, Pearson’s r was calculated. To compare survival curves, a Log rank (Mantel-Cox) test was performed.

Results

Approach

As shown in Figure 1A, animals were implanted with electrodes at 1 month of age and recorded for a continuous 24 hrs-long period at 1.2, 2, 3, 4, 5,and 6 months of age. At 3 and 6 months of age, mice were tested for NOL and NOR. Afterwards they were perfused and immunocytochemistry was conducted for NeuN and ΔFosB. Sections were made in the coronal plane and relatively anterior and posterior levels were compared to sample different parts of hippocampus. Some mice were not possible to include in all assays either because they died before reaching 6 months or for other reasons.

Behavior

NOL and the NOR were selected because they show deficits in Tg2576 mice at just 3 months of age (Duffy et al. 2015). This is before deficits have been shown in other tasks in >9 months-old Tg2576 mice (i.e., Morris Water Maze, Radial Arm Water Maze, Y-Maze; Yassine et al. 2013; Wolf et al. 2016). Figure 1B-D shows a schematic of the specific experimental procedures for NOL and NOR.

NOL

The results of the NOL task are presented in Figures 2-4. In Figure 2 the diets, age, sex, and individual animals are presented. Lines connect the behavior of a given animal in training with the behavior during testing. Behavior is measured as the time spent exploring the object that is moved during the testing period. Exploration of the novel object is expressed as a percentage of the total exploration of both objects. When the slope of the line increases between training and testing, the time spent exploring the moved object increased in testing relative to training. This increase reflects a preference to explore the moved object during the testing period, and suggests the mouse can recall the old object locations and can recognize the new object location. Supplemental Figure 2 shows the data with mean ± sem.

Choline enrichment reduced spatial memory deficits.

A. Three months-old mice.

1. WT and Tg2576 mice exposed to the low choline diet showed spatial memory deficits.

2. Tg2576 exposed to the intermediate diet showed spatial memory deficits but not WT.

3. Spatial memory was improved in animals exposed to the high choline diet.

B. Six months-old mice.

1. WT and Tg2576 mice exposed to the low choline diet showed spatial memory deficits.

2. WT and Tg2576 mice exposed to the intermediate diet showed spatial memory deficits.

3. WT and Tg2576 mice exposed to the high choline diet had improved spatial memory.

An important foundation for this task is equal preference for the objects during training. The lack of preference reflects no inherent bias for one object vs. the other. We confirmed the lack of bias for each genotype and each diet. Thus, a two-way ANOVA with genotype or age (3 or 6 months) revealed no effect of genotype on object exploration during training (F(1,97)=0.40, p=0.530).There also was no effect of age (F(1,70)=1.94, p=0.167). When the factors were diet and age there also was no effect of diet (F(2,97)=0.06, p=0.941) or age (F(2,70)=2.94, p=0.058) on object exploration during training.

Next, we asked whether Tg2576 mice were deficient in the test phase of NOL when they were fed the standard mouse diet (the intermediate diet). First we studied 3 months-old mice (Figure 2A). We used a two-way ANOVA with genotype and task phase (training versus testing) as main factors. In animals that received the intermediate diet, a two-way ANOVA revealed a significant main effect of the task phase (F(1,48)=8.51, p=0.005) and a trend for a genotype effect (F(1,48)=3.80, p=0.057). Consistent with previous studies (Duffy et al. 2015), Tukey-Kramer post-hoc analyses revealed a significant increase in the exploration of the novel object during testing in WT (p=0.017) but not Tg2576 mice (p=0.676; Figure 2A). Thus, Tg2576 mice were impaired in NOL when fed the intermediate diet.

In contrast to Tg2576 mice fed the intermediate diet, Tg2576 mice that received the high choline diet showed memory for object location. Thus, a two-way ANOVA revealed a main effect of the task phase (F(1,72)=46.16, p<0.0001) but not genotype (F(1,72)=1.74, p=0.191). Tukey-Kramer post-hoc analyses revealed a significant increase in the exploration of the novel object during testing in both WT (p=0.0003) and Tg2576 mice (p<0.0001; Figure 2A3). Therefore, high choline supplementation improved memory in Tg2576 mice.

In animals treated with the low choline diet, there was no effect of the phase of the task (F(1,74)=1.41, p=0.250) or genotype (F(1,74)=0.027, p=0.871). These data suggest that relatively low levels of choline during early life impaired spatial memory in both WT and Tg2576 mice. Thus, low choline had a significant adverse effect. Adverse effects are further supported by survival plots showing that there was more mortality at earlier ages in offspring exposed to the low choline diet (Supplemental Figure 1).

Figure 2B shows the results in 6 months-old WT and Tg2576 mice. In animals that received the intermediate diet, a two-way ANOVA showed no effect of genotype (F(1,36)=0.01, p=0.907) or task phase (F(1,36)=2.36, p=0.133), revealing memory deficits in WT and Tg2576 mice (Figure 2B). In contrast, the high choline group showed a main effect of task phase (F(1,56)=22.18, p<0.0001) but not genotype (F(1,56)=1.78, p=0.188). Tukey-Kramer post-hoc analyses showed a significant increase in the exploration of the novel object during testing in both WT (p<0.001) and Tg2576 (p=0.020; Figure 2B). Thus, the high choline-treated mice showed object location memory but the mice fed the intermediate diet did not. The mice that received the relatively low choline diet showed an effect of genotype (F(1,50)=4.36, p=0.042; Figure 2B) in that training in WT mice differed from testing in Tg2576 mice (p=0.031). However, there was no main effect of task phase (F(1,50)=3.75, p=0.058). WT mice did not differ between training and testing (p=0.114) and Tg2576 mice did not either (p=0.921). Therefore, mice fed the low choline and intermediate diets were impaired and the high choline-treated mice were not.

The discrimination indices are shown in Figure 3 and results led to the same conclusions as the analyses in Figure 2. For the 3 months-old mice (Figure 3A), the low choline group did not show the ability to perform the task for WT or Tg2576 mice. Thus, a two-way ANOVA showed no effect of genotype (F(1,74)=0.027, p=0.870) or task phase (F(1,74)=1.41, p=0.239). For the intermediate diet-treated mice, there was no effect of genotype (F(1,50)=0.3.52, p=0.067) but there was an effect of task phase (F(1,50)=8.33, p=0.006). WT mice showed a greater discrimination index during testing relative to training (p=0.019) but Tg2576 mice did not (p=0.664). Therefore, Tg2576 mice fed the intermediate diet were impaired. In contrast, high choline-treated mice performed well. There was a main effect of task phase (F(1,68)=39.61, p=<0.001) with WT (p<0.0001) and Tg2576 mice (p=0.0002) showing preference for the moved object in the test phase. Interestingly, there was a main effect of genotype (F(1,68)=4.50, p=0.038) because the discrimination index for WT training was significantly different from Tg2576 testing (p<0.0001) and Tg2576 training was significantly different from WT testing (p=0.0003).

Novel object location results based on the discrimination index.

A. Results are shown for the 3 months-old WT and Tg2576 mice based on the discrimination index.

1. Mice fed the low choline diet showed object location memory only in WT.

2. Mice fed the intermediate diet showed object location memory only in WT.

3. Mice fed the high choline diet showed memory both for WT and Tg2576 mice. Therefore, the high choline diet improved memory in Tg2576 mice.

B. The results for the 6 months-old mice are shown.

1-2. There was no significant memory demonstrated by mice that were fed either the low or intermediate choline diet.

3. Mice fed a diet enriched in choline showed memory whether they were WT or Tg2576 mice. Therefore, choline enrichment improved memory in all mice.

The discrimination indices of 6 months-old mice led to the same conclusions as the results in Figure 2. There was no evidence of discrimination in low choline-treated mice by two-way ANOVA (no effect of genotype, (F(1,42)=3.25, p=0.079; no effect of task phase, F(1,42)=0.278, p=0.601). The same was true of mice fed the intermediate diet (genotype, F(1,12)=1.44, p=0.253; task phase, F(1,12)=2.64, p=0.130). However, both WT and Tg2576 mice performed well after being fed the high choline diet (effect of task phase, (F(1,52)=58.75, p=0.0001, but not genotype (F(1,52)=1.197, p=0.279). Tukey-Kramer post-hoc tests showed that both WT (p<0.0001) and Tg2576 mice that had received the high choline diet (p=0.0005) had elevated discrimination indices for the test session.

Taken together, these results demonstrate the lasting beneficial effects of the high choline diet and adverse effects of low choline on offspring in a spatial memory task in Tg2576 as well as WT littermates.

NOR

We first confirmed that animals did not show preference for one object over the other in training. Indeed, a two-way ANOVA revealed no main effects of genotype at 3 months (F(1,71)=2.59, p=0.536) and no effect of diet (F(2,71)=0.22, p=0.809). The same result was obtained at 6 months of age (genotype, (F(1,66)=0.11, p=0.746); diet, (F(2,66)=0.98, p=0.376). Thus, novel object exploration during training approached 50% in all treatment groups, independent of genotype and maternal diet, and at both ages.

Figure 4 shows the results for the NOR task. In this figure the diets, age, sex, and individual animals are presented. Lines connect the performance of a given animal in training and in testing. Supplemental Figure 3 shows the data with mean ± sem. Figure 4A shows that there were no impairments at 3 months of age. In animals treated with the low choline diet, there was a main effect of task phase (F(1,52)=19.81, p<0.0001), no effect of genotype (F(1,52)=0.020, p=0.887), and more exploration of the novel object during the test phase (WT, p=0.007; Tg2576, p<0.002). In animals that received the intermediate diet, there also was a significant main effect of task phase (F(1,38)=30.88, p<0.0001) and no effect of genotype (F(1,38)=0.97, p=0.330). Animals explored the novel object significantly more during testing in WT (p=0.014) and Tg2576 mice (p=0.002; Figure 3B). For the high choline group, there also was an effect of task phase (F(1,26)=17.51, p=0.0003) and not genotype (F(1,26)=3.53, p=0.072) and a significant increase in the exploration of the novel object during testing in both WT (p=0.006) and Tg2576 (p=0.027; Figure 3B).

WT and Tg2576 mice showed object recognition memory regardless of diet.

A. Three months-old mice.

1-3. WT and Tg2576 mice treated performed the NOR task.

B. Six months-old mice.

1-3. WT and Tg2576 mice performed the NOR task.

Figure 4B shows the data for 6 months of age. Animals that received the low choline diet performed the NOR task at 3 months of age (two-way ANOVA, task phase (F(1,26)=32.66, p<0.0001; genotype (F(1,26)=0.048, p=0.821). Tukey-Kramer post-hoc tests showed that there was a significant increase in the exploration of the novel object in testing both in WT (p<0.001) and Tg2576 (p= 0.003). For mice that had been fed the intermediate choline diet, they also performed the NOR task at 3 months of age (two-way ANOVA, task phase (F(1,19)=13.65, p=0.002); genotype (F(1,19)=0.28, p=0.604). Tukey-Kramer post-hoc analyses revealed a significant increase in the exploration of the novel object during testing compared to training for both WT (p=0.0027) and Tg2576 (p=0.039; Figure 4B). Mice fed the high choline diet also showed no deficit in NOR at 3 months of age. Thus, a two-way ANOVA revealed a main effect of task phase (F(1,27)=16.26, p=0.0004) but not genotype (F(1,27)=0.24, p=0.625). Tukey-Kramer post-hoc-hoc analyses revealed a significant increase in the exploration of the novel object during testing in both WT (p=0.003) and Tg2576 (p=0.007; Figure 4B).

Thus, at both 3 and 6 months of age, WT and Tg2576 mice performed well in the NOR task. These results suggest that young Tg2576 mice are less sensitive to NOR than NOL. The greater sensitivity to NOL is consistent with past demonstrations that the DG contributes to NOL (Sahay et al. 2011; Kesner et al. 2015; Spyrka and Hess 2018; Vandrey et al. 2020; Gulmez Karaca et al. 2021; GoodSmith et al. 2022). Also, our implementation of NOL may have increased the DG-dependence of the task by making the object locations relatively close together, because in studies by Pofahl and colleagues, it was shown that distances between objects like those we used made NOL DG-dependent (Pofahl et al. 2021).

Exploration time

Total object exploration (TOE) was measured to address any effects of genotype or diet on the total duration of exploration of objects (Figure 5).

Tg2576 mice treated with the choline supplemented diet spent more time with the objects at each approach.

A-B. NOL.

There was no effect of diet on time exploring or approaches to the novel object in the NOL task at 3 months of age (A) but there were significant differences at 6 months of age (B). Mice exposed to the high choline diet spent more time exploring than mice that had been fed the low choline or the intermediate diet.

C-D. NOR.

There was no effect of diet on time exploring or approaches to the novel object in the NOR task at 3 months of age (C) but there were significant differences at 6 months of age (D). Mice that had been fed the high choline diet spent more time exploring than mice that had been fed the low choline or the intermediate diet.

For NOL TOE at 3 months of age, there were effects of genotype (F(1,67)=6.891, p=0.01) but not maternal diet (F(2,87)=0.67, p=0.63). Tukey-Kramer post-hoc tests showed that Tg2576 mice fed the low choline diet showed more exploration than WT mice fed the same diet but the effect was on the border of significance (p=0.049; data not shown). Three months-old mice tested with NOR showed no significant effect of genotype (F(1,76)=1.35, p=0.25) or diet (F(2,76)=0.30, p=0.61). Since there were weak or no effects of genotype, all genotypes are pooled for Figure 5A1 (NOL) and Figure 5B1 (NOR).

For NOL TOE at 6 months of age, there was no effect of genotype (F(1,66)=0.33; p=0.57). For NOR TOE, the same was true (genotype: F(1,72)=0.96, p=0.33; diet: F(2,72)=8.50, p=0.0005). Because genotype was not a significant factor, we pooled genotypes (NOL, Figure 5C1; NOR, Figure 5D2).

With genotypes pooled we determined how diets differed. For NOL TOE, mice treated with the low choline diet had less exploration than mice fed the high choline diet (one-way ANOVA, F(2,77)=6.90; p=0.002; Tukey post-hoc test, p=0.005; Figure 5C1) and the mice that were fed the intermediate diet also had less exploration than mice fed the high choline diet (Tukey post-hoc test, p=0.008; Figure 5C1). Results for NOR were similar: mice treated with the low choline diet had less exploration than mice fed the high choline diet (one-way ANOVA, F(2,74)=4.81; p=0.020; Tukey post-hoc test, p=0.016; Figure 5D1) and the mice that were fed the intermediate diet also had less exploration than mice fed the high choline diet (Tukey post-hoc test, p=0.03; Figure 5D1).

To gain insight into the potential reason for the effect of diet onexploration at 6 months of age, we measured total object approaches (TOATOA was defined as the number of approaches to the familiar object + number of approaches to the novel object (Figure 5). There was no effect of diet on TOA for NOL (F(2,75)=2.88; p=0.092; Figure 5C2) or NOR (F(2,74)=1.81, p=0.171; Figure 5D2).

Taken together, the results indicate that, in 6 months-old mice, animals that received the high choline diet spent more time with objects at each approach. This may underlie increased object memory in high choline-treated mice by increasing time for information processing during an approach. Another possibility is that as Tg2576 mice age they show compensatory changes that enhance memory. This has been suggested for procedural learning (Middei et al. 2004) but to the best of our knowledge, it has not been shown for memory of objects. A third possibility is that the high choline diet reduces anxiety (Glenn et al. 2012; Langley et al. 2015; McCall et al. 2015). Reduced anxiety may lessen fear of exploring objects, and as a result, animals may spend more time with objects.

Anatomy

NeuN

To further analyze the effects of dietary choline early in life, we used an antibody against a neuronal nuclear antigen, NeuN. Previous studies have found that reduced expression of NeuN often occurs in neurons after insults or injury (e.g., ischemia, toxicity and even aging; Lind et al. 2005; Portiansky et al. 2006; Buckingham et al. 2008; Kadriu et al. 2009; Matsuda et al. 2009; Won et al. 2009; Duffy et al. 2011). For example, when NeuN is reduced after a reduction in ankyrin-rich membrane spanning kinase D-interacting substrate of 220K, ARMS/Kidins, a protein critical to neurotrophin signaling, neurons in the entorhinal cortex demonstrate signs of toxicity such as pyknosis (Duffy et al. 2011; Duffy et al. 2015). Since NeuN is reduced when Aβ levels are elevated (Wu et al. 2016) and NeuN is reduced in AD patients (Camporez et al. 2021), we used NeuN expression to ask if the high choline diet could rescue the loss of NeuN-ir.

As shown in Figure 6A, relatively weak NeuN-ir was observed in Tg2576 that received the low choline diet and the intermediate diet compared to the high choline diet (Figure 6A). This observation is consistent with the vulnerability of the hilus to insult and injury (Scharfman 1999). Therefore, we conducted a two-way ANOVA with diet and septotemporal levels as main factors. To examine the septotemporal axis, quantification from 2-3 anterior coronal sections were averaged to provide a value for the septal pole of the DG and 2-3 posterior sections were averaged to assess the caudal DG. The most ventral, temporal levels were not sampled. There was no effect of diet (F(2,30)=2.11, p=0.137) nor rostral-caudal level (F(1,30)=0.02, p=0.877) but there was a significant interaction of diet and rostral-caudal level (F(2,30)=3.88, p=0.036). A one-way ANOVA using anterior values showed a significant effect of diet (F(2,21)=7.58, p=0.003) but this was not the case for posterior values (F(2,23)=0.13; p=0.876). In the anterior DG, there was less NeuN-ir in mice fed the low choline diet versus the high choline diet (Tukey-Kramer post-hoc test, p=0.029) and less NeuN in mice fed the intermediate vs high choline diet (p=0.003). The results suggest that choline enrichment protected dorsal hilar neurons from NeuN loss in Tg2576 mice.

Choline supplementation improved NeuN immunoreactivity (ir) in hilar cells in Tg2576 animals

A. Representative images of NeuN-ir staining in the anterior DG of Tg2576 animals.

1. A section from a Tg2576 mouse fed the low choline diet. The area surrounded by a box is expanded below. Red arrows point to NeuN-ir hilar cells. Mol=molecular layer, GCL=granule cell layer, HIL=hilus. Calibration for the top row, 100 µm; for the bottom row, 50 µm.

2. A section from a Tg2576 mouse fed the intermediate diet. Same calibrations as for 1.

3. A section from a Tg2576 mouse fed the high choline diet. Same calibrations as for 1.

B. Quantification methods. Representative images demonstrate the thresholding criteria used to quantify NeuN-ir.

1. A NeuN-stained section. The area surrounded by the white box is expanded in the inset (arrow) to show 3 hilar cells. The 2 NeuN-ir cells above threshold are marked by blue arrows. The 1 NeuN-ir cell below threshold is marked by a green arrow.

2. After converting the image to grayscale, the cells above threshold were designated as red. The inset shows that the two cells that were marked by blue arrows are red while the cell below threshold is not.

3. An example of the threshold menu from ImageJ showing the way the threshold was set. Sliders (red circles) were used to move the threshold to the left or right of the histogram of intensity values. The final position of the slider (red arrow) was positioned at the onset of the steep rise of the histogram.

C. NeuN-ir in Tg2576 and WT mice. Tg2576 mice had either the low, intermediate, or high choline diet in early life. WT mice were fed the standard diet (intermediate choline).

1. Tg2576 mice treated with the high choline diet had significantly more hilar NeuN-ir cells in the anterior DG compared to Tg2576 mice that had been fed the low choline or intermediate diet. The values for Tg2576 mice that received the high choline diet were not significantly different from WT mice, suggesting that the high choline diet restored NeuN-ir.

2. There was no effect of diet or genotype in the posterior DG, probably because the low choline and intermediate diet did not appear to lower hilar NeuN-ir.

To ask if the improvement in NeuN after MCS in Tg256 restored NeuN to WT levels we used WT mice. For this analysis we used a one-way ANOVA with 4 groups: Low choline Tg2576, Intermediate Tg2576, High choline Tg2576, and Intermediate WT (Figure 5C). Tukey-Kramer multiple comparisons tests were used as the post hoc tests. The WT mice were fed the intermediate diet because it is the standard mouse chow, and this group was intended to reflect normal mice. The results showed a significant group difference for anterior DG (F(3,25)=9.20; p=0.0003; Figure 5C1) but not posterior DG (F(3,28)=0.867; p=0.450; Figure 5C2). Regarding the anterior DG, there were more NeuN-ir cells in high choline-treated mice than both low choline (p=0.046) and intermediate choline-treated Tg2576 mice (p=0.003). WT mice had more NeuN-ir cells than Tg2576 mice fed the low (p=0.011) or intermediate diet (p=0.003). Tg2576 mice that were fed the high choline diet were not significantly different from WT (p=0.827).

ΔFosB

To complement the information from the video-EEG recordings (see below) we used a marker of elevated neuronal activity, ΔFosB (Figure 7). ΔFosB is a truncated variant of the transcription factor FosB, which is increased by enhanced neuronal activity; ΔFosB has a half-life of approximately 8 days (Ulery-Reynolds et al., 2009), so when ΔFosB is elevated, it reflects increased neuronal activity over the last 10-14 days (McClung et al. 2004). Previous studies have shown that when the J20 mouse model of AD is examined with an antibody to ΔFosB, the GC layer shows extremely high levels of ΔFosB expression (Corbett et al. 2017), similar to a mouse with chronic spontaneous seizures (Morris et al. 2000). This is not surprising since the J20 mice have recurrent seizures (Palop et al. 2007). Therefore, we asked if Tg2576 mice would have robust ΔFosB in the GC layer, and choline supplementation would reduce it. We also included WT mice. These WT mice would allow us to address the possibility that the high choline diet restored ΔFosB back to normal. As for the NeuN immunocytochemistry, the WT mice were fed the intermediate diet.

Choline supplementation reduced ΔFosB expression in dorsal GCs of Tg2576 mice.

A. Representative images of ΔFosB staining in GCL of Tg2576 animals from each treatment group.

1. A section from a low choline-treated mouse shows robust ΔFosB-ir in the GCL. Calibration, 100 µm.

2-3. Sections from intermediate (2) and high choline (3)-treated mice. Same calibration as 1.

B. Quantification methods. Representative images demonstrating the thresholding criteria established to quantify ΔFosB.

1. A ΔFosB -stained section shows strongly-stained cells (white arrows).

2. A strict thresholding criteria was used to make only the darkest stained cells red.

C. Use of the strict threshold to quantify ΔFosB-ir.

1. Anterior DG. Tg2576 mice treated with the choline supplemented diet had significantly less ΔFosB-ir compared to the Tg2576 mice fed the low or intermediate diets. Tg2576 mice fed the high choline diet were not significantly different from WT mice, suggesting a rescue of ΔFosB-ir.

2. There were no significant differences in ΔFosB-ir in posterior sections.

D. Methods are shown using a threshold that was less strict.

1. Some of the stained cells that were included are not as dark as those used for the strict threshold (white arrows).

2. All cells above the less conservative threshold are shown in red.

E. Use of the less strict threshold to quantify ΔFosB-ir.

1. Anterior DG. Tg2576 mice that were fed the high choline diet had less ΔFosB-ir pixels than the mice that were fed the other diets. There were no differences from WT mice, suggesting restoration of ΔFosB-ir by choline enrichment in early life.

2. Posterior DG. There were no significant differences between Tg2576 mice fed the 3 diets or WT mice.

There was strong expression of ΔFosB in Tg2576 GCs in mice fed the low choline diet (Figure 7A1). The high choline diet and intermediate diet appeared to show less GCL ΔFosB-ir (Figure 7A2-3). A two-way ANOVA was conducted with the experimental group (Tg2576 low choline diet, Tg2576 intermediate choline diet, Tg2576 high choline diet, WT intermediate choline diet) and location (anterior or posterior) as main factors. There was a significant effect of group (F(3,32)=13.80, p=<0.0001) and location (F(1,32)=8.69, p=0.006). Tukey-Kramer post-hoc tests showed that Tg2576 mice fed the low choline diet had significantly greater ΔFosB-ir than Tg2576 mice fed the high choline diet (p=0.0005) and WT mice (p=0.0007). Tg2576 mice fed the low and intermediate diets were not significantly different (p=0.275). Tg2576 mice fed the high choline diet were not significantly different from WT (p>0.999). There were no differences between groups for the posterior DG (all p>0.05).

ΔFosB quantification was repeated with a lower threshold to define ΔFosB-ir GCs (see Methods) and results were the same (Figure 7D). Two-way ANOVA showed a significant effect of group (F(3,32)=14.28, p< 0.0001) and location (F(1,32)=7.07, p=0.0122) for anterior DG but not posterior DG (Figure 7D). For anterior sections, Tukey-Kramer post hoc tests showed that low choline mice had greater ΔFosB-ir than high choline mice (p=0.0024) and WT mice (p=0.005) but not Tg2576 mice fed the intermediate diet (p=0.275); Figure 7D1). Mice fed the high choline diet were not significantly different from WT (p=0.993; Figure 7D1). These data suggest that high choline in the diet early in life can reduce neuronal activity of GCs in offspring later in life. In addition, low choline has an opposite effect, suggesting low choline in early life has adverse effects.

IIS

Previous research has shown that Tg2576 mice exhibit IIS (Figure 8A) starting at very young ages: 4 weeks (Kam et al. 2016) or 6 weeks (Bezzina et al. 2015). Therefore, we performed video-EEG using cortical (left frontal cortex, right occipital cortex) and hippocampal (left and right) electrodes in WT and Tg2576 mice (Figure 8B-C). Animals were recorded for 24 hrs each session so that the major behavioral states (exploration, awake rest, and sleep) were sampled well. Consistent with previous studies (Bezzina et al. 2015; Kam et al. 2016) we observed IIS in Tg2576 mice but not WT littermates. Therefore, analyses below were only in Tg2576 mice.

The high choline diet reduced IIS frequency in Tg2576 animals.

A. 1. Representative example of an IIS (red arrow). IIS were defined as occurring in all 4 channels (generalized) to distinguish them from focal spikes that can often be artifact.

2-4. The IIS shown in A1 is expanded.

B. 1. Scatter plot of IIS frequency at each age of recording.

2. Means and sem are plotted. The high choline diet group had fewer IIS than the intermediate diet group at ages 1-3 months ($, p<0.05) and the low choline group had less IIS than the intermediate diet at ages 1-4 months (@, p<0.05). The high and low choline diet groups were significantly different at ages 3 and 4 months (*, p<0.05).

As shown in Figure 8, animals that received the intermediate diet had a significantly higher number of IIS in the 24 hrs-long recording periods compared to animals that received the high choline and the low choline diets. A two-way ANOVA (mixed model analysis) showed that there was a significant effect of age (F(2,37)=3.38; p=0.036) and maternal diet (F(2,36)=8.12; p=0.089). At the 1.25-month recording, Tukey-Kramer post-hoc analyses showed that IIS frequency in animals treated with the intermediate diet was higher compared to animals treated with the low (p=0.027) or high choline diets (p=0.038). This also was true for 2 months and 3 months (low choline, high choline, p<0.05). At 4 months, the low choline group had significantly reduced IIS frequency compared to the mice that had received the intermediate diet (p=0.009) but this was not the case for the high choline group compared to the intermediate group (p=0.976). At 5-6 months, IIS frequencies were not significantly different in the mice fed the different diets (all p>0.05), probably because IIS frequency becomes increasingly variable with age (Kam et al. 2016). One source of variability is seizures, because there was a sharp increase in IIS during the day before and after a seizure (Supplemental Figure 4). Another reason that the diets failed to show differences was that the IIS frequency generally declined at 5-6 months. This can be appreciated in Figure 8B and Supplemental Figure 6B. These data are consistent with prior studies of Tg2576 mice where IIS increased from 1 to 3 months but then waxed and waned afterwards (Kam et al., 2016).

Seizures and premature mortality in mice fed the low choline diet

We found mice fed the low choline diet had greater ΔFosB-ir in GCs and the hilus showed very low NeuN-ir. Therefore, we asked whether low choline-treated mice had more seizures than mice fed the other diets. We recorded 8 mice by video-EEG for 5 days each (at 6 months of age) to examine seizures and found 2 mice from the low choline group had seizures (11 seizures over 2 of the 5 days in one mouse, 1 seizure in the other mouse) whereas none of the other mice had seizures (n=0/4, 2 intermediate and 2 high choline, data not shown).

These values are probably an underestimate for the low choline group because many mice in this group appeared to die in a severe seizure prior to 6 months of age (Supplemental Figure 5). Therefore, the survivors at 6 months probably were the subset with few seizures.

The reason that low choline-treated mice appeared to die in a seizure was that they were found in a specific posture in their cage which occurs when a severe seizure leads to death (Supplemental Figure 5). They were found in a prone posture with extended, rigid limbs (Supplemental Figure 5). Regardless of how the mice died, there was greater mortality in the low choline group compared to mice that had been fed the high choline diet (Log-rank (Mantel-Cox) test, Chi square 5.36, df 1, p=0.021; Supplemental Figure 5A).

Correlations between IIS and other measurements

As shown in Figure 9A, IIS were correlated to behavioral performance in some conditions. For these correlations, only mice that were fed the low and high choline diets were included because mice that were fed the intermediate diet did not have sufficient EEG recordings in the same mouse where behavior was studied. IIS frequency over 24 hrs was plotted against the preference for the novel object in the test phase (Figure 9A). For NOL, IIS were significantly less frequent when behavior was the best, but only for the high choline-treated mice (Pearson’s r, p=0.022). In the low choline group, behavioral performance was poor regardless of IIS frequency (Pearson’s r, p=0.933; Figure 9A1). For NOR, there were no significant correlations (low choliNe, p=0.202; high choline, p=0.680) but few mice were tested in the high choline-treated mice (Figure 9B2).

Correlations between IIS, Behavior, and hilar NeuN-ir.

A. IIS frequency over 24 hrs is plotted against the preference for the novel object in the test phase of NOL. A greater preference is reflected by a greater percentage of time exploring the novel object.

1. The mice fed the high choline diet (red) showed greater preference for the novel object when IIS were low. These data suggest IIS impaired object location memory in the high choline-treated mice. The low choline-treated mice had very weak preference and very few IIS, potentially explaining the lack of correlation in these mice.

2. There were no significant correlations for IIS and NOR. However, there were only 4 mice for the high choline group, which is a limitation.

B. IIS frequency over 24 hrs is plotted against the number of dorsal hilar cells expressing NeuN. The dorsal hilus was used because there was no effect of diet on the posterior hilus..

1. Hilar NeuN-ir is plotted against the preference for the novel object in the test phase of NOL. There were no significant correlations.

2. Hilar NeuN-ir was greater for mice that had better performance in NOR, both for the low choline (blue) and high choline (red) groups. These data support the idea that hilar cells contribute to object recognition (Kesner et al. 2015; Botterill et al. 2021; GoodSmith et al. 2022).

We also tested whether there were correlations between dorsal hilar NeuN-ir cell numbers and IIS frequency. In Figure 9B, IIS frequency over 24 hrs was plotted against the number of dorsal hilar cells expressing NeuN. The dorsal hilus was used because there was no effect of diet on the posterior hilus. For NOL, there was no significant correlation (low choline, p=0.273; high choline, p=0.159; Figure 9B1). However, for NOR, there were more NeuN-ir hilar cells when the behavioral performance was strongest (low choline, p=0.024; high choline, p=0.016; Figure 9B2). These data support prior studies showing that hilar cells, especially mossy cells (the majority of hilar neurons), contribute to object recognition (Botterill et al. 2021; GoodSmith et al. 2022).

Sex differences

As shown in Figures 2-3, there appeared to be no sex differences in NOL or NOR. For confirmation, we studied NOL further, since that was the task that showed effects of diet. A 3-way ANOVA with sex, genotype and task phase as factors showed no significant effect of sex at 3 months of age (low choline diet, F(1,35)=0.0001; p=0.99). Results were the same for the 3 months-old mice fed the intermediate diet (F(1,22)=0.062; p=0.810) and 3 months-old mice that were fed the high choline diet (F(1,34)=0.432; p=0.522). At 6 months of age there also were no significant effects of sex (low choline diet, F(1,23)=0.343; p=0.571; intermediate diet, F(1,16)=2.49; p=0.130; high choline diet, F(1,27)=0.29; p=0.873).

Regarding IIS, we did not find sex differences (Supplemental Figure 6). In Supplemental Figure 6A, comparisons are made of males and females fed the low or high choline diet at 1.2, 2, and 3 months-of age. A two-way ANOVA with sex and diet as factors showed no effect of sex (F(1,20) =2.48; p=0.132) or diet (F(1,20)=3.56; p=0.067). Note that for this comparison we used the low choline and high choline groups only because there were few mice for each sex at all ages. The results for months 2-3 also showed no effect of sex (2 months: F(1,17)=0.637; p=0.429; 3 months: F(1,15)=0.178; p=0.668).

In Supplemental Figure 6B we grouped mice at 1-2 months, 3-4 months and 5-6 months so that there were sufficient females and males to compare each diet. A two-way ANOVA with diet and sex as factors showed a significant effect of diet (F(2,47)=46.21; p<0.0001) at 1-2 months of age, but not sex (F1,47)=0.11, p=0.758). Post-hoc comparisons showed that the low choline group had fewer IIS than the intermediate group, and the same was true for the high choline-treated mice. Thus, female mice fed the low choline diet differed from the females (p<0.0001) and males (p<0.0001) fed the intermediate diet. Male mice that had received the low choline diet different from females (p<0.0001) and males (p<0.0001) fed the intermediate diet. Female mice fed the high choline diet different from females (p=0.002) and males (p<0.0001) fed the intermediate diet, and males fed the high choline diet difference from females (p<0.0001) and males (p<0.0001) fed the intermediate diet.

For the 3-4 months-old mice there was also a significant effect of diet (F(2,32)=10.82, p=0.0003) but not sex (F(1,32)=1.05, p=0.313). Post-hoc tests showed that low choline females were different from males fed the intermediate diet (p=0.007), and low choline males were also significantly different from males that had received the intermediate diet (p=0.006). There were no significant effects of diet (F(2,23)=1.21, p=0.317) or sex (F(1,23)=0.84, p=0.368) at 5-6 months of age.

Discussion

Summary

This study showed that choline supplementation in early life had several beneficial effects in Tg2576 mice. The high choline diet led to improved behavior in the NOL task, improved expression of ΔFosB, rand educed IIS frequency.

We also found surprising effects of treating mice with relatively low choline in the maternal diet. The low choline-treated Tg2576 mice had the least hilar NeuN and most GCL ΔFosB expression. There was an impairment in the NOL task not only in low choline-treated Tg2576 mice but also WT mice. In addition, the low choline-treated Tg2576 mice showed premature mortality compared to Tg2576 mice fed the high choline diet. These results were surprising because the relatively low levels of choline have been considered by others to be a control (Moreno et al. 2013; Mellott et al. 2017; Velazquez et al. 2019). Our data suggest the low choline diet led to several adverse effects. Yet the IIS frequency was low in the low choline-treated Tg2576 mice, suggesting a beneficial effect. It is possible the chronic overexpression of GC ΔFosB led to reduced IIS by altering GC gene expression, since ΔFosB regulates the GC transcriptome (Corbett et al. 2017; You et al. 2018; Stephens et al. 2020). Alterations in GC transcription could impair IIS generation because the GCs appear to be a site where IIS are generated (Lisgaras and Scharfman 2023). Altered GC gene expression might also explain the reduced performance in the NOL task of low choline-treated mice, because the DG plays an important role in this detecting novel object locations normally (Lee et al. 2005).

We showed for the first time that Tg2576 mice have pathology in the hilus at 6 months of age in that hilar neurons are deficient in immunostaining with an anti-NeuN antibody. Tg2576 offspring on the high choline diet showed a restoration of hilar NeuN staining. Because hilar neurons play a role in NOL (Bui et al. 2018), it is possible that rescue of hilar neurons by the high choline diet led to improved spatial memory. However, there was no significant correlation between the number of NeuN-ir hilar cells and testing for the NOL task. There was a significant correlation for the NOR task, with both low and high choline groups performing better when hilar NeuN-ir cell numbers were greater. The significant correlation with NOR test performance is consistent with a role of somatostatin-expressing hilar GABAergic neurons (Nagarajan et al. 2024) and hilar glutamatergic mossy cells in NOR and object-related activity (Botterill et al. 2021; GoodSmith et al. 2022). Somatostatin-expressing hilar neurons (Tallent 2007; Savanthrapadian et al. 2014; Hofmann et al. 2016) and mossy cells (Sloviter 1994; Scharfman 1999; Scharfman and Myers 2012; Jinde et al. 2013; Bui et al. 2018; Botterill et al. 2019) have been suggested to contribute to excitability of GCs so phenotypic rescue of mossy cells may also have contributed to the ability of the high choline diet to reduce hyperexcitability.

Benefits of high choline

The results of this study are consistent with previous reports that prenatal or postnatal choline supplementation improves object memory. One study of iron deficiency showed that choline supplementation improved NOR (Kennedy et al. 2014). Wistar rats treated with a diet high (5.0 g/kg) in choline had improved NOR relative to rats that had been fed a diet lower (1.1 g/kg) in choline (Moreno et al. 2013). In rats that were aged to 24 months, animals that had been fed a high choline diet showed improved NOR, but only in females (Glenn et al. 2008).

Although few studies have examined the effects of MCS on hyperexcitability, our results showing the benefits of high choline are consistent with reports that a methyl-enriched diet (including choline but also betaine, folic acid, vitamin B12, L-methionine and zinc) reduced spike wave discharges in the offspring (Sarkisova et al. 2023) and audiogenic seizure severity (Poletaeva et al. 2014). Perinatal ethanol treatment that increased excitability was mitigated by choline chloride injection from postnatal day 10-30 (Grafe et al. 2022).

To our knowledge our study is the first to show that MCS can exert effects on IIS and ΔFosB. However, prior studies have shown that MCS in rats improved memory after severe seizures. Using a convulsant to induce several hours of severe seizures (status epilepticus, SE), there is usually extensive hippocampal neuronal loss and when tested in the subsequent days and weeks, memory impairment. MCS reduced this impairment following SE in the Morris water maze (Yang et al. 2000; Holmes et al. 2002; Wong-Goodrich et al. 2011).

Our results are consistent with previous studies of mouse models of AD. In the APP/PS1 mouse model, it was shown that life-long choline supplementation, to the level we used (5.0 g/kg) improved memory in aged mice compared to the 1.1 g/kg diet (Velazquez et al. 2019). Postnatal choline supplementation also improved memory in APP/PS1 mice (Wang et al. 2019). Other improvements were also shown in the APP/PS1 mice, such as increased choline acetyltransferase, the major enyzme for acetylcholine synthesis (Mellott et al. 2017; Velazquez et al. 2019). In 3xFAD mice, a diet without choline had numerous deleterious consequences, including increased Aβ and tau phosporylation (Dave et al. 2023)

Adverse effects of the diet with relatively low choline

The relatively low choline diet had several adverse effects, which was surprising because the low choline levels are not considered very low in some prior studies (Moreno et al. 2013; Mellott et al. 2017; Velazquez et al. 2019). However, the past studies using 1.1 g/kg choline did not test Tg2576 mice. Past studies sometimes used the diet for only part of gestation rather than all of gestation and stopped after birth instead of continuing until weaning. We fed the diet throughout gestation and until weaning. Nevertheless, it is surprising. One possible explanation is an interaction of the low choline diet with the strain, SJL (Swiss James Lambert). To our knowledge this strain has not been tested with different diets before. SJL mice are descended from Swiss Webster, and are prone to reticulum cell sarcoma (Haran-Ghera et al. 1967). This strain is also characterized by vulnerability to infection causing a multiple sclerosis-like syndrome (Linzey et al. 2023).

One of the adverse effects was high mortality. Mice appeared to die in seizures. Consistent with chronic hyperexcitability, mice that had received the low choline diet had high ΔFosB expression in the GCs. Another adverse effect was weak NeuN in the hilus in mice that had received the low choline diet. For both WT and Tg2576 mice, NOL was impaired. These data suggest that the offspring of mothers fed the low choline diet were unhealthy, possibly leading to or because of seizures. If mice were unhealthy, IIS might have been reduced due to impaired excitatory synaptic function. Another reason for reduced IIS is that the mice that had the low choline diet had seizures which interrupted REM sleep. Thus, seizures in Tg2576 mice typically started in sleep. Less REM sleep would reduce IIS because IIS occur primarily in REM. Also, seizures in the Tg2576 mice were followed by a depression of the EEG (postictal depression; Supplemental Figure 3) that would transiently reduce IIS. A different, radical explanation is that the intermediate diet promoted IIS rather than low choline reducing IIS. Instead of choline, a constituent of the intermediate diet may have promoted IIS.

NeuN

As mentioned above, NeuN is a neuronal nuclear antigen that can be phosphorylated, and when that occurs the antibody to NeuN no longer binds to NeuN. NeuN is phosphorylated in response to oxidative damage, brain injury, and toxicity (Lind et al. 2005). Therefore, it was of interest when we saw that hilar neurons of the dentate gyrus showed reduced NeuN in Tg2576 mice. Hilar neurons are mainly glutamatergic mossy cells and somatostatin (SOM)/neuropeptide Y (NPY)-expressing GABAergic neurons (HIPP cells; (Houser 2007; Scharfman and Myers 2012; Scharfman 2016), and both neuronal types are implicated in spatial memory functions of the GCs, as well as their excitability (Myers and Scharfman 2009; 2011; Scharfman and Myers 2012; Jinde et al. 2013; Scharfman 2016; Raza et al. 2017; Bui et al. 2018; GoodSmith et al. 2019; Li et al. 2021). If damaged by the high intracellular APP/Aβ levels in young Tg2576 mice, the hilar neurons would be expected to show NeuN loss. This idea is consistent with deficits in SOM and NPY-stained cells (Chan-Palay 1987), indicating a vulnerability.

The reason for rescue of anterior hilar but not posterior hilar NeuN with the high choline diet is unclear. The greater sensitivity of dorsal neurons may be related to differences in gene expression patterns, since some of the genes that are differentially expressed along the dorsal-ventral axis could affect vulnerability to insult or injury (Cembrowski et al. 2016; Zhang et al. 2018). However, there was a similar pattern to the data for both anterior and posterior. The lack of statistical significance appeared to be greater variance in the posterior data.

ΔFosB

Given that seizures are associated with elevated GC expression of ΔFosB (Chen et al. 1997; McClung et al. 2004; Corbett et al. 2017; You et al. 2017), the differences in ΔFosB expression levels observed here could be the result of increased seizures in animals that received the low choline diet. Therefore, we suggest that the high ΔFosB in Tg2576 mice fed the low choline diet reflects more seizures than in the other two diets.

ΔFosB is a transcription factor that is linked to cognition. In the J20 mouse model of AD, elevated ΔFosB in GCs led to reduced cognition, and when ΔFosB was selectively reduced the cognition improved (Corbett et al. 2017). Therefore, the reduction in ΔFosB by the high choline diet was important to show hyperexcitability was reduced and also important because it showed how high choline may benefit cognition.

Choline and cholinergic neurons

There are many suggestions for the mechanisms that allow MCS to improve health of the offspring. One hypothesis that we are interested in is that MCS improves outcomes by reducing IIS. Reducing IIS would potentially reduce hyperactivity, which is significant because neuronal activity can increase release of Aβ and tau (Cirrito et al. 2005; Cirrito et al. 2008; Bero et al. 2011; Yamada et al. 2014; Yamamoto et al. 2015; Hettinger et al. 2018). IIS would also be likely to disrupt sleep since it represents aberrant synchronous activity over widespread brain regions. The disruption to sleep could impair memory consolidation since it is a notable function of sleep (Graves et al. 2001; Poe et al. 2010). Sleep disruption also has other negative consequences such as impairing normal clearance of Aβ (Nedergaard and Goldman 2020). Indeed, in patients with AD, IIS and similar events, IEDs, are correlated with memory impairment (Vossel et al. 2016).

How would choline supplementation in early life reduce IIS of the offspring? It may do so by making BFCNs more resilient. That is significant because BFCN abnormalities appear to cause IIS. Thus, selective silencing of BFCNs reduced IIS. The cholinergic antagonist atropine also reduces IIS when injected systemically (Kam et al., 2016), and it reduced elevated synaptic activity of GCs in young Tg2576 mice in vitro (Alcantara-Gonzalez et al. 2021). These studies are consistent with the idea that early in AD there is elevated cholinergic activity (DeKosky et al. 2002; Ikonomovic et al. 2003; Kelley et al. 2014; Mufson et al. 2015; Kelley et al. 2016), while later in life there is cholinergic degeneration. Indeed, the overactivity of cholinergic neurons of the first months of life could cause the degeneration at older ages.

Why would MCS make BFCNs resilient? There are several possibilities that have been explored, based on genes upregulated by MCS. One attractive hypothesis is that neurotrophic support for BFCNs is retained after MCS but in aging and AD it declines (Gautier et al. 2023). The neurotrophins, notably nerve growth factor (NGF) and brain-derived neurotrophic factor (BDNF) have been known for a long time to support the health of BFCNs (Mufson et al. 2003; Niewiadomska et al. 2011).

Limitations

One of the limitations of the study was that the maternal diets were not matched exactly. Although the low and high choline diets only had differences in choline, the intermediate diet had different amounts of choline as well as other constituents (Supplementary Table 1). Therefore, the only diet comparison with a difference restricted to choline is the low versus high choline diets (Supplemental Table 1). The intermediate diet was useful however, because numerous studies in AD mouse models employ this standard diet.

In addition, groups were not exactly matched. Although WT mice do not have IIS, a WT group for each of the Tg2576 groups would have been useful. Instead, we included WT mice for the behavioral tasks and some of the anatomical assays. Related to this point is that several mice died during the long-term EEG monitoring of IIS. However, this is unlikely to have had a major effect because IIS were low in frequency in all groups over the last months of the study.

Regarding sex differences, there may have been differences if females had been separated by stage of the estrous cycle at death. This possibility is raised by prior data showing that rats and mice during proestrous and estrous mornings have hyperexcitability but not at other cycle stages (Scharfman et al. 2003).

The Tg2576 mouse model is one of many murine models of AD, and as with all models there are inherent limitations. The Tg2576 model recapitulates familial AD, whereas the majority of AD is sporadic. Tg2576 mice also lack tau pathology. However, MCS has now demonstrated structural/functional benefits in several AD-relevant models, namely Tg2576 (this report), APP/PS1 (Alldred et al. 2021; Dave et al. 2023), Ts65Dn (Velazquez et al. 2013; Powers et al. 2016; Strupp et al. 2016; Powers et al. 2017; Alldred et al. 2021; Alldred et al. 2023) and 3xTg mice (Dave et al. 2023). Moreover, increasing evidence suggests that humans with AD have low serum choline and are improved by dietary choline (Dave et al. 2023; Judd et al. 2023a; b).

Conclusions

There is now a substantial body of evidence that MCS promotes learning and memory in the offspring of normal rats and improves behavior and other abnormalities in mice that simulate AD. There also is evidence that serum levels of choline are low in AD. Therefore, it is exciting to think that dietary choline might improve AD. Given the past work also shows benefits to mice that simulate DS, the benefits of choline supplementation appear to extend to DS. The present study adds to the growing consensus that MCS is restorative by showing that hyperexcitability and pathology in the DG of Tg2576 mice are improved by MCS. They also suggest that there may be some adverse effects of a relatively low level of choline in the diet.

eLife Assessment

In this fundamental work, the authors demonstrated that maternal choline supplementation improved spatial memory, reduced hyperexcitability, and restored NeuN expression in a familial Alzheimer's disease mouse model. Interestingly, choline deficiency increased mortality, while paradoxically reduced hyperexcitability. Using behavior, electrophysiological, and histological measures, the authors present convincing evidence supporting the significant role of maternal choline supplementation in protecting hippocampal functions vulnerable to Alzheimer's disease.

Reviewer #1 (Public Review):

Summary:

Chartampila et al. describe the effect of early-life choline supplementation on cognitive functions and epileptic activity in a mouse model of Alzheimer's disease. The cognitive abilities were assessed by the novel object recognition test and the novel object location test, performed in the same cohort of mice at 3 months and 6 months of age. Neuronal loss was tested using NeuN immunoreactivity, and neuronal hyperexcitability was examined using FosB and video-EEG recordings, providing multi-level correlations between these different parameters.

Strengths:

The study was designed as a 6-month follow-up, with repeated behavioral and EEG measurements through disease development and multilevel correlations providing valuable and interesting findings on AD progression and the effect of early-life choline supplementation. Moreover, the behavioral data that suggest an adverse effect of low choline in WT mice are interesting and important also beyond the context of AD, highlighting the dramatic effect of diet on the phenotypes of animal models.

Weaknesses:

The readability could be improved.

Author response:

The following is the authors’ response to the previous reviews.

Weaknesses:

The readability could be improved.

We have gone through the paper again and tried to revise the text to improve readability.

Reviewer #1 (Recommendations For The Authors):

(1) Thank you for adding the discrimination ratio. However, as Fig 2 and 3 depict the same experimental data, consider harmonizing the presentation (symbols and colors) and consolidating the Figs for clarity.“

This is an excellent point but it is actually very hard to harmonize symbols and colors because the data are divided in different ways. Upon considering this further, we actually don’t want to make the symbols and colors the same because it would be misleading. For example, WT and Tg training and testing session data are divided into grey and white throughout Figure 2, but in Figure 3, training and testing session data are pooled. To color code them grey and white in Figure 3 might make it seem that in Figure 3 training and testing were separated.

(2) Fig 5 is missing

We are not sure why Figure 5 was absent since it was present in our copy of the submitted pdf. We have double checked and in the revised manuscript we are sure Figure 5 is included.

(3) Fig 6 add raw data for WT

We have added raw WT data. Revised figure 6 includes the raw data in part A4.

(4) Fig 7 add raw data for WT

We have added raw WT data. Revised Figure 7 includes the raw data in part A4.

Version 3

Abstract

Abstract

Maternal choline supplementation (MCS) improves cognition in Alzheimer’s disease (AD) models. However, effects of MCS on neuronal hyperexcitability in AD are unknown. We investigated effects of MCS in a well-established mouse model of AD with hyperexcitability, the Tg2576 mouse. The most common type of hyperexcitability in Tg2576 mice are generalized EEG spikes (interictal spikes; IIS). IIS also are common in other mouse models and occur in AD patients. Im mouse models, hyperexcitability is also reflected by elevated expression of the transcription factor ΔFosB in the granule cells (GCs) of the dentate gyrus (DG), which are the principal cell type. Therefore we studied ΔFosB expression in GCs. We also studied the the neuronal marker NeuN within hilar neurons of the DG because other studies have reduced NeuN protein expression is a sign of oxidative stress or other pathology. This is potentially important because hilar neurons regulate GC excitability.

Tg2576 breeding pairs received a diet with a relatively low, intermediate or high concentration of choline. After weaning, all mice received the intermediate diet. In offspring of mice fed the high choline diet, IIS frequency declined, GC ΔFosB expression was reduced, and NeuN expression was restored. Using the novel object location task, spatial memory improved. In contrast, offspring exposed to the relatively low choline diet had several adverse effects, such as increased mortality. They had the weakest hilar NeuN immunoreactivity and greatest GC ΔFosB protein expression. However, their IIS frequency was low, which was surprising. The results provide new evidence that a diet high in choline in early life can improve outcomes in a mouse model of AD, and relatively low choline can have mixed effects. This is the first study showing that dietary choline can regulate hyperexcitability, hilar neurons, ΔFosB and spatial memory in an animal model of AD.

Introduction

Diet has been suggested to influence several aspects of brain health. One dietary intervention that has been studied extensively in rodents and humans is supplementation of the maternal diet with the nutrient choline (maternal choline supplementation; MCS). MCS improves many aspects of brain health in humans (Zeisel and da Costa 2009; Jiang et al. 2014). In normal rats, MCS also is beneficial, with numerous studies showing improved behavior in the adult offspring (for review see Meck and Williams 2003).

In Alzheimer’s disease (AD), changes to the diet have been recommended (Bourre 2006; Power et al. 2019; Mao et al. 2021; Lobo et al. 2022), including MCS (Strupp et al. 2016; Velazquez et al. 2020; Dave et al. 2023; Judd et al. 2023a; b). One reason for the recommendation is that serum levels of choline are low in individuals with AD (Dave et al. 2023; Judd et al. 2023b). In addition, using mouse models of AD, increased choline improved many characteristics of the disease, ranging from inflammation to glucose metabolism and the hallmark amyloid and tau pathology (Dave et al. 2023; Judd et al. 2023a; b). In the Ts65Dn mouse model of Down syndrome (DS) and AD, MCS led to improved memory and attention in Ts65Dn offspring (Strupp et al. 2016; Powers et al. 2017; Powers et al. 2021). In addition, degeneration of basal forebrain cholinergic neurons (BFCNs) in Ts65Dn mice, a hallmark of DS and AD, was reduced (Kelley et al. 2016; Powers et al. 2017; Alldred et al. 2023; Gautier et al. 2023).

Here we asked if MCS would improve Tg2576 mice, a model of familial AD where a mutation in the precursor to amyloid β (Aβ), amyloid precursor protein (APP) is expressed by the hamster prion protein promoter (Hsiao et al. 1996). This is a commonly used mouse model that simulates aspects of AD.

One of the reasons to use Tg2576 mice was to ask if MCS would improve the hyperexcitability found in the mice. The Tg2576 mouse is ideal because hyperexcitability is robust (Bezzina et al. 2015; Kam et al. 2016; Lisgaras and Scharfman 2023), and BFCNs are likely to play a role in the hyperexcitabililty (Kam et al. 2016; Lisgaras and Scharfman, 2023). In our past work, the primary type of hyperexcitability was interictal spikes (IIS), which are named because they occur in between seizures (ictal events) in epilepsy. They are studied by EEG. Importantly, IIS are found in numerous mouse models of AD (J20, Brown et al. 2018; Ts65Dn, Presenilin2-/-, Lisgaras and Scharfman 2023; Amyloid precursor protein (APP) /Presenilin1 (PS1), Shoob et al. 2023), as well as patients (Sanchez et al. 2012; Vossel et al. 2016; Beagle et al. 2017; Vossel et al. 2017; Vossel and Karageorgiou 2021; Vossel 2023).

We also studied hyperexcitability reflected by high protein expression of ΔFosB, a transcription factor that is increased when neurons have been highly active over the prior 10-14 days (McClung et al. 2004). We studied ΔFosB in granule cells (GCs) of the dentate gyrus (DG), because numerous ΔFosB-expressing GCs occur when there are IIS (You et al. 2017; You et al. 2018). We also focused on GCs because we found that the cell layer of GCs (GCL) is where IIS are largest relative to area CA1 and overlying neocortex (Lisgaras and Scharfman 2023). Moreover, closed loop optogenetic silencing of GCs reduced IIS (Lisgaras and Scharfman 2023).

We also asked if behavior would improve if MCS reduced IIS frequency in Tg2576 mice. The basis for this question is in studies of epilepsy, where IIS disrupt cognition (Rausch et al. 1978; Aarts et al. 1984; Holmes and Lenck-Santini 2006; Kleen et al. 2010; Kleen et al. 2013; Gelinas et al. 2016) . Also, both IIS and longer-lasting epileptiform discharges, interictal epileptiform discharges (IEDs), are increased when cognitive dysfunction is impaired in AD patients (Vossel et al. 2016).

We fed dams one of 3 diets, which were relatively low, intermediate or high in choline (Figure 1A; Supplemental Table 1). The high choline diet provided levels of choline similar to other studies of MCS in rodents (Meck et al. 1988; Loy et al. 1991; Holler et al. 1996; Meck and Williams 1999; Sandstrom et al. 2002; Mellott et al. 2004; Glenn et al. 2012; Kelley et al. 2019). After weaning, the intermediate diet was used. Offspring were implanted with electrodes for EEG at 1 month of age, and 24 hrs-long recordings were made every month shortly thereafter and ending at 6 months of age

Schematic of the behavioral procedures for the NOL and NOR tasks.

A. Experimental timeline. Dams were fed one of 3 diets for mating, gestation, and offspring consumed the diet until weaning at 25-30 days of age. At 1 month-old, offspring were implanted with electrodes for EEG. Recordings for 24 hrs started 1 week later (5 weeks-old; 1.2 months-old in the grahps).Additional recordings were made at, 2, 3, 4, 5, and 6 months of age. At 3 and 6 months of age behavior was tested, mice were perfused, brains were sectioned, and sections were processed with antibodies against NeuN and ΔFosB.

B. Prior to NOL or NOR, animals were acclimated to the testing arena. There were 3 acclimation sessions separated by 24 hrs during which animals were allowed to freely explore for 5 min.

C. In the NOL task, animals were placed in a cage with two identical objects and allowed to freely explore for 5 min (Training). After one hour they were brought back to the cage, where one object was displaced, and allowed to freely explore for 5 min (Testing).

D. In the NOR task. animals were placed in a cage with two identical objects and allowed to freely explore for 5 min (Training). After one hour they were brought back to the cage, where one object was replaced, and allowed to freely explore for 5 min (Testing).

The time span (1-6 months of age) is relevant because IIS are present at these ages (Bezzina et al., 2016; Kam et al., 2016; Lisgaras and Scharfman, 2023). The novel object location (NOL) task is impaired also (at 3-4 months of age in Tg2576 mice;(Duffy et al. 2015), so we evaluated NOL at 3 months. We also testeda related task, novel object recognition (NOR; Vogel-Ciernia and Wood 2014). To understand the persistence of any effects of the diets, we repeated NOL and NOR at 6 months. The interval between testing was sufficiently long that it is unlikely that testing at 3 months affected testing at 6 months, but we can not exclude the possibility. At 6 months of age we perfusion-fixed mice, sectioned the brains, and evaluated NeuN and ΔFosB protein expression. NeuN was studied because it is a neuronal marker that is reduced in numerous pathological conditions (Buckingham et al. 2008; Kadriu et al. 2009; Matsuda et al. 2009; Duffy et al. 2011; Duffy et al. 2015). We also assayed ΔFosB because it is a method complementary to EEG recordings of hyperexcitabililty. It also can be used to assess hyperexcitability of GCs.

The results showed a remarkable effect of the high choline diet. IIS and spatial memory were improved, as was NeuN and ΔFosB expression. Interestingly, the relatively low choline diet had mixed effects, reducing IIS frequency, but making NOL, ΔFosB, and NeuN worse. The mice also died prematurely relative to mice that were fed the high choline diet. We also report for the first time that there is loss of NeuN-ir in the DG hilus of Tg2576 mice, which is important because impaired hilar neurons could cause GC hyperexcitability (Sperk et al. 2007; Scharfman and Myers 2012). In summary, we make a strong argument for choline supplementation in early life to improve outcomes in an AD model, especially the DG.

Methods

Animals

All experimental procedures followed the guidelines set by the National Institute of Health (NIH) and were approved by the Institutional Animal Care and Use Committee (IACUC) of the Nathan Kline Institute for Psychiatric Research. Mice were housed in standard mouse cages (26 cm wide x 40 cm long x 20 cm high) with corn cob bedding and a 12 hrs-long light-dark cycle. Food and water were provided ad libitum.

Tg2576 mice express a mutant form of human APP (isoform 695) with a mutation found in a Swedish family with AD (Lys670Arg, Met671Leu), driven by the hamster prion protein promoter (Hsiao et al. 1996). Mice were bred in-house from heterozygous Tg2576 males and non-transgenic female mice (C57BL6/SJL F1 hybrid). Wildtype (WT) mice were obtained from Jackson Laboratories (Stock# 100012). Genotypes were determined by the New York University Mouse Genotyping Core facility using a protocol to detect APP695.

Breeding pairs were randomly assigned to receive one of 3 diets with different concentrations of choline chloride: 1.1 g/kg (AIN-76A, Dyets Inc.), 2.0 g/kg (Purina 5008, W.F. Fisher and Son Inc.) and 5.0 g/kg (AIN-76A, Dyets Inc.; Supplemental Table 1). These diets (“low choline” diet, “intermediate choline” diet and “high choline” diet, respectively) were used until weaning (25-30 days of age). After weaning, all mice were fed the intermediate diet. Mice were housed with others of the same sex (1-4 per cage). WT mice received the intermediate diet during breeding, gestation, and after birth.

Behavior

General information

Starting at least 24 hrs prior to all experiments, mice were housed in the room where they would be tested behaviorally . Both NOL and NOR were preceded by 3 acclimation sessions (5 min each). Acclimations were conducted between 10:00 a.m. and 12:00 p.m.NOL and NOR were composed of a training and a testing session (5 min each; Figure 1B-D).,Training and testing was conducted between 1:00 p.m. and 4:00 p.m. The interval between training and testing was 60 minutes and therefore the tasks tested short-term memory (Vogel-Ciernia and Wood 2014). NOL and NOR tests were separated by 7 days to ensure effects of one task did not affect the next. Prior studies have failed to find an effect on one task on the next when a 7 day-long interval is used (Botterill et al. 2021). The order of testing (NOL before NOR or NOR before NOL) was randomized.

Video recordings of all training and testing sessions were captured using a USB camera (Logitech HD Pro C920, Logitech). All equipment was cleaned using 70% ethanol before each use. In pilot studies, we confirmed that objects were explored equally by animals (see Results).. One object was composed of different 2 green LEGO pieces (approximately 5 cm x 7 cm x 7 cm); the other object was a pineapple (3 cm x 3 cm x 5 cm) and made of painted metal (Supplemental Figure 1; Botterill et al. 2021).

NOL

NOL was conducted in a standard rat cage (26 cm wide x 40 cm long x 20 cm high) with different pictures on 3 sides of the cage to provide a consistent context. Pictures included several shapes and colors. The dimensions of the pictures were 1) 10 cm x 21 cm, 2) 16 cm x 18 cm, and 3) 17 cm x 20 cm. During the acclimations, animals were allowed to freely explore the cage. During the training session, mice were placed in the cage where they had been acclimated, with two identical objects, 1 in a corner of the cage (e.g., left top) and the second in the adjacent corner (e.g., left bottom, Supplemental Figure 1). The mice were then removed and placed in their home cage for 1 hr. During the test session, one of the objects was moved to the opposite end of the cage (left top to right top; Supplemental Figure 1).

NOR

NOR was conducted in the cage described above for NOL. As for NOL, during the acclimations animals were allowed to freely explore the cage. During the training session, mice were placed in the cage used for acclimation with two identical objects centered along the shortest cage wall (Supplemental Figure 1). The mouse was then removed and placed in their home cage for 1 hr. During the testing session, one of the objects was replaced with a different one. The objects that were identical were 2 pineapple-like objects, and the new object was made of

Quantification

The experimenter who conducted the analysis was blind to the genotypeand which object was novel or familiar. Videos of the training and testing sessions were analyzed manually. A subset of data was analyzed by two independent blinded investigators and they were in agreement. Exploration was quantified based on the guidelines of Vogel-Ciernia and Wood (2014). However, in addition to the 2014 recommendations, we also analyzedthe number of approaches to an object. Time spent exploring each object was calculated as well, defined as the duration of time the nose was pointed at the object and the nose was located within 2 cm of the object. The time spent exploring also included the time animals spent on top of the object if they were looking down and sniffing it. An approach was defined by a movement toward the object and ending at least 2 cm from the edge of the object.

Animals that remembered the objects that were explored during the training phase were expected to demonstrate an increased preference for novelty in the test phase (Ennaceur and Delacour 1988). In other words, exploration of the novel object during testing was expected to be higher than 50% of the total time of object exploration. Cognitive impairment was defined as significantly less than 50%.

Anatomy

Perfusion-fixation and sectioning

Mice were initially anesthetized by isoflurane inhalation (NDC# 07-893-1389, Patterson Veterinary), followed by urethane (2.5 g/kg; i.p.; Cat# U2500, Sigma-Aldrich). Under deep anesthesia, the abdominal and heart cavities were opened with surgical scissors and a 23g needle inserted into the heart. The atria was clipped, the needle was clamped in place with a hemostat, and the animal was transcardially-perfused using 10 mL of room temperature (RT) saline (0.9% sodium chloride in double distilled (dd) H2O) using a peristaltic pump (Minipuls1; Gilson), followed by 30 mL of 4°C 4% paraformaldehyde (PFA, Cat# 19210, Electron Microscopy Sciences) in 0.1 M phosphate buffer (PB; pH 7.4). The brains were immediately removed, hemisected, and post-fixed for at least 24 hrs in a scintillation vial with 10 mL 4% PFA (at 4°C).

Following post-fixation, 50 µm-thick coronal sections were made using a vibratome (VT 1000P, Leica Biosystems). Sections were collected serially and stored at 4°C in 24-well tissue culture plates containing cryoprotectant solution (30% glycerol, 30% ethylene glycol, 40% 0.1 M phosphate buffer, pH 6.7).

Throughout the hippocampus, every 6th section was processed with an antibody to NeuN or ΔFosB. For analysis of anterior DG, 2-3 sections that were 300 µm apart were selected from ∼1.5-2.5 mm posterior to Bregma. For posterior DG, 2-3 sections that were 300 µm apart were selected starting at ∼3.5 mm posterior to Bregma. The values of the 2-3 sections were averaged so that there was one measurement for anterior and one for posterior for each mouse.

NeuN and ΔFosB immunohistochemistry

Free floating sections were washed in 0.1 M Tris Buffer (TB; 3 x 5 min), followed by a 3 min wash in 1% H2O2. Sections were then washed in 0.1 M TB (3 x 5 min) and incubated for 60 min in 5% horse serum for NeuN (Cat# S-2000, Vector) or 5% goat serum for ΔFosB (Cat# S-1000, Vector), diluted in a solution of 0.25% (volume/volume or v/v) Triton-X 100, and 1% (weight/volume or w/v) bovine serum albumin in 0.1 M TB. Sections were then incubated overnight at 4°C in primary antibody to NeuN (mouse monoclonal, 1:5000; Cat# MAB377, Chemicon International) or anti-ΔFosB (rabbit monoclonal, 1:1000; Cat# D3S8R, Cell Signaling), diluted in a solution of 0.25% (v/v) Triton-X 100, and 1% (w/v) bovine serum albumin in 0.1M TB. Both NeuN and ΔFosB have been well-characterized as antigens (Mullen et al. 1992; Wolf et al. 1996; Chen et al. 1997; Sarnat et al. 1998) and the antibodies we used have been commonly employed in the past (Chen et al. 1997; Duffy et al. 2013; Corbett et al. 2017). On the following day, sections were washed in 0.1 M TB (3 x 5 min) and then incubated for 60 min in biotinylated horse anti-mouse secondary antibody to NeuN (1:500, Cat# BA-2000, Vector) or biotinylated goat anti-rabbit secondary antibody to ΔFosB (1:500, Cat# BA-1000), diluted in a solution of 0.25% (v/v) Triton-X 100, and 1% (w/v) bovine serum albumin in 0.1 M TB. The sections were then washed in 0.1 M TB (3 x 5 min) and incubated in avidin-biotin complex for 2 hrs (1:1000; Cat# PK-6100, Vector). They were washed in 0.1 M TB (3 x 5 min) and then reacted in a solution containing 0.5 mg/mL 3, 3’-diaminobenzidine (DAB; Cat# D5905, Sigma-Aldrich), 40 µg/mL ammonium chloride (Cat# A4514, Sigma-Aldrich), 25 mg/mL D(+)-glucose (Cat# G5767, Sigma-Aldrich), and 3 g/mL glucose oxidase (Cat# G2133, Sigma-Aldrich) in 0.1 M TB. This method slowed the reaction time so that the reaction could be stopped when the immunoreactivity was robust but background was still low. The sections were then washed in 0.1 M TB (3 x 5 min), mounted on gelatin-coated slides (1% bovine gelatin, Cat# G9391, Sigma-Aldrich) and dried at RT overnight. The following day they were dehydrated in increasing concentrations of ethanol (90%, 10 min; 95% 10 min; 100%, 10 min; 100% again, 10 min), washed in Xylene (10 min; Cat# 534056, Sigma-Aldrich) and cover-slipped with Permount (Cat# 17986-01, Electron Microscopy Sciences).

Analysis

Photomicrographs were acquired using ImagePro Plus V7.0 (Media Cybernetics) and a digital camera (Model RET 2000R-F-CLR-12, Q-Imaging). NeuN and ΔFosB staining were quantified from micrographs using ImageJ (V1.44, National Institutes of Health). All images were first converted to grayscale and in each section, the hilus was traced, defined by zone 4 of Amaral (1978). A threshold was then calculated to identify the NeuN-stained cell bodies but not background. Then NeuN-stained cell bodies in the hilus were quantified manually. Note that the threshold was defined in ImageJ using the distribution of intensities in the micrograph. A threshold was then set using a slider in the histogram provided by Image J (see Figure 6B,7B). The slider was pushed from the low level of staining (similar to background) to the location where staining intensity made a sharp rise, reflecting stained cells. Cells with labeling that was above threshold were counted.

To quantify ΔFosB-stained cells, images were converted to grayscale and in each section, the GCL was outlined and defined as a region of interest (ROI). A threshold was then set to identify the well-stained cells but not the background, analogous to the method to define threshold in the for NeuN. Two thresholds were used because some cells were lightly stained and others much more robust in their staining. The first threshold was most inclusive of immunoreactive cells.

Presumably this group reflected cells with less neuronal activity as well as those with intense activity. Then a second analysis was done with the same sections using a higher threshold. This group corresponded to cells with the highest activity. ImageJ was used to calculate the area (in pixels) within the ROI that was above threshold (see Figure 7B).

Video-electroencephalographic (video-EEG) recordings

Stereotaxic surgery

EEG electrodes (Cat# 8209, 0.10” stainless steel screws, Pinnacle Technology) were implanted in Tg2576 animals at 4 weeks of age. WT littermates were not implanted because previous work from our laboratory has shown that WT mice have no IIS (Kam et al. 2016). The animals were anesthetized by isoflurane inhalation (3% isoflurane. 2% oxygen for induction) in a rectangular transparent plexiglas chamber (18 cm long x 10 cm wide x 8 cm high) made in-house. For maintenance during surgery, isoflurane was < 1.75% and flow rate was 1.2 L/min. Mice were placed in a stereotaxic apparatus (David Kopf Instruments). Prior to the implantation of the electrodes, animals were injected with the analgesic Buprenex (0.2 mg/kg, s.c.; buprenorphine hydroxide, NDC# 12496-0757-5, Reckitt Benckiser), which was diluted in saline (0.03 mg/mL in sterile 0.9% sodium chloride solution, Vedco Inc.). The skull was exposed with a midline incision and 6 holes were drilled for the placement of subdural screw electrodes.

The coordinates for the electrode placement were: right occipital cortex (AP -3.5 mm, ML 2.0 mm), left frontal cortex (AP -0.5 mm, ML -1.5 mm), left hippocampus (AP -2.5 mm, ML -2.0 mm), and right hippocampus (AP -2.5 mm, ML 2.0 mm (Kam et al. 2016). An additional screw over the right olfactory bulb was used as ground (AP 2.3 mm, ML 1.8 mm) and another screw was placed at the midline over the cerebellum (relative to Lambda: AP -1.5 mm, ML -0.5 mm) and used as reference. Screws were attached to an 8-pin connector (Cat# ED85100-ND, Digi-Key Corporation), which was placed over the skull and secured with dental cement (Cat# 4734FIB, Lang Dental Mfg. Co.).

After surgery, animals were placed on a heating blanket overnight and injected with lactated Ringer’s solution (50 mL/kg at 31°C; NDC# 099355000476, Aspen Veterinary Resources Ltd). They were then transferred to the room where the video-EEG was recorded.

Video-EEG recording

Video-EEG started 1 week after surgery, at 1.25 months (5 weeks) of age. The animals were recorded at 5 weeks, and subsequently 2, 3, 4, 5 and 6 months of age. Each recording session lasted 24 hrs so that a long period of sleep could be acquired, since IIS occur primarily in sleep (Kam et al. 2016).

Mice were placed into 21 cm x 19 cm transparent cages with food and water provided ad libitum and corncob bedding. A pre-amplifier was inserted into the 8-pin connector on the skull, which was connected to a multichannel commutator (Cat# 8408, Pinnacle Technology). This arrangement allowed for free range of movement throughout the recording. EEG signals were acquired at 2 kHz and bandpass filtered at 0.5 – 200 Hz using Sirenia Acquisition (V2.0.4, Pinnacle Technology). Simultaneous video recordings were captured using an infrared camera (Cat# AP-DCS100W, Apex CCTV).

Analysis

EEG recordings were analyzed offline with Neuroscore V3.2.1 (Data Science International). IIS were defined as large amplitude, 10-75 msec deflections occurring synchronously in all 4 leads (Kam et al. 2016). They were quantified by first setting an amplitude threshold (calculated separately for each recording) and a duration criterion (10-75 msec). To determine the amplitude threshold, we calculated the root mean square (RMS) amplitude for noise in each recording during a 60 sec segment of baseline activity. A period of baseline activity was selected that did not include any artifacts, which were defined as extremely large deflections much greater than any other activity, and composed of a waveform that was unlike any physiological activity. They typically were accompanied by a large shake in the animal, a cable hitting a part of the cage, or scratching the implant. The threshold for IIS was set at 9 standard deviations above the RMS amplitude. This threshold was selected because it identified IIS extremely well and excluded artifacts. The IIS in one of the hippocampal channels was used instead of another channel because of data suggesting IIS begin in the hippocampus (Kam et al. 2016; Lisgaras and Scharfman 2023). Following the automatic analysis of the IIS, the software’s detection accuracy was verified manually for each recording to ensure that all spikes appeared in all 4 channels and that no artifacts were included.

Statistical comparisons

Data are expressed as mean ± standard error of the mean (SEM). The significant was set to <0.05 prior to all experiments. Tests were conducted using Prism software (V 9.4.1, Graphpad).

Parametric data comparing two groups used unpaired two-tailed t-tests. For >2 groups, one-way ANOVA was used followed by Tukey-Kramer post-hoc tests that corrected for multiple comparisons. For data with two main factors, two-way ANOVA was followed by Tukey-Kramer post-hoc tests. Interactions are not reported in the Results unless they were significant. For the analysis of IIS frequency from 1.2 months of age up to 6 months-of age, a repeated measures ANOVA (RMANOVA) was used.

Tests for normality (Shapiro-Wilk) and homogeneity of variance (Bartlett’s test) were used to determine if parametric statistics could be used. When data were not normal, non-parametric data were used. When there was significant heteroscedasticity of variance, data were log transformed. If log transformation did not resolve the heteroscedasticity, non-parametric statistics were used. For non-parametric data, Mann-Whitney U tests were used to compare two groups and Kruskal-Wallis for >2 groups. The post-hoc tests were Dunn’s test for multiple comparisons.

For correlations, Pearson’s r was calculated. To compare survival curves, a Log rank (Mantel-Cox) test was performed.

Results

Approach

As shown in Figure 1A, animals were implanted with electrodes at 1 month of age and recorded for a continuous 24 hrs-long period at 5 weeks, 2, 3, 4, 5,and 6 months of age. At 3 and 6 months of age, mice were tested for NOL and NOR. Afterwards they were perfused and immunocytochemistry was conducted for NeuN and ΔFosB. Sections were made in the coronal plane and relatively anterior and posterior levels were compared to sample different parts of hippocampus. Some mice were not possible to include in all assays either because they died before reaching 6 months or for other reasons.

Behavior

NOL and the NOR were selected because they show deficits in Tg2576 mice at just 3 months of age (Duffy et al. 2015). This is before deficits have been shown in other tasks (i.e., Morris Water Maze, Radial Arm Water Maze, Y-Maze) which occur after 9 months of age (Yassine et al. 2013; Wolf et al. 2016). Figure 1B-D shows a schematic of the specific experimental procedures for NOL and NOR.

NOL

The results of the NOL task are presented in Figures 2-4. In Figure 2, lines connect the data for a given animal in training and testing. Data are the time spent exploring the objects. Exploration of the object that is moved in the training session (the novel oboject) is expressed as a percentage of the total exploration time of both objects. When the slope of the line increases between training and testing, the time spent exploring the moved object increased in testing relative to training. This increase reflects a preference to explore the moved object during the testing period, and suggests the mouse can recall the old object locations and can recognize the new object location. Supplemental Figure 2 shows the data with mean ± sem.

Choline enrichment reduced spatial memory deficits.

A. Three months-old mice. For A and B, breeders were fed the specified diet and offspring were tested.

1. Low choline diet. WT and Tg2576 offspring showed spatial memory deficits.

2. Intermediate diet. Tg2576 showed spatial memory deficits but not WT.

3. High choline diet. Spatial memory was improved..

B. Six months-old mice.

1. Low choline diet. WT and Tg2576 offspringshowed spatial memory deficits.

2. Intermediate diet. WT and Tg2576 mice showed spatial memory deficits.

3. High choline diet. WT and Tg2576 mice had improved spatial memory.

An important foundation for this task is equal preference for the objects during training. The lack of preference reflects no inherent bias for one object vs. the other. We confirmed the lack of bias for each genotype and each diet during training. To conduct statistics we first conducted a two-way ANOVA based on the training data with genotype or age (3 or 6 months) as factors. There were no effects of genotype (F(1,97)=0.40, p=0.530) or age (F(1,70)=1.94, p=0.167) on object exploration time. Then we conducted a two-way ANOVA to ask if diet played a role. The main factors were diet and age and there also was no effect of diet (F(2,97)=0.06, p=0.941) or age (F(2,70)=2.94, p=0.058) on object exploration time.. Thus, novel object exploration during training approached 50% in all treatment groups, independent of genotype and maternal diet, and at both ages.

Next, we confirmed that Tg2576 mice were deficient in the test phase of NOL when they were fed the standard mouse diet (the intermediate diet). First we studied 3 months-old mice (Figure 2A). We used a two-way ANOVA with genotype and task phase (training versus testing) as main factors. There was a significant main effect of the task phase (F(1,48)=8.51, p=0.005) and a trend for a genotype effect (F(1,48)=3.80, p=0.057). Consistent with previous studies (Duffy et al. 2015), Tukey-Kramer post-hoc analyses revealed a significant increase in the exploration of the novel object during testing in WT (p=0.017) but not Tg2576 mice (p=0.676; Figure 2A). Thus, Tg2576 mice were impaired in NOL when fed the intermediate diet.

In contrast to Tg2576 mice fed the intermediate diet, Tg2576 mice that received the high choline diet showed memory for object location. Thus, a two-way ANOVA revealed a main effect of the task phase (F(1,72)=46.16, p<0.0001) but not genotype (F(1,72)=1.74, p=0.191). Tukey-Kramer post-hoc analyses revealed a significant increase in the exploration of the novel object during testing in both WT (p=0.0003) and Tg2576 mice (p<0.0001; Figure 2A3). Therefore, high choline supplementation improved memory in Tg2576 mice.

In animals treated with the low choline diet, there was no effect of the phase of the task (F(1,74)=1.41, p=0.250) or genotype (F(1,74)=0.027, p=0.871). These data suggest that relatively low levels of choline during early life impaired spatial memory in both WT and Tg2576 mice. Thus, low choline had a significant adverse effect. Adverse effects are further supported by survival plots showing that there was more mortality at earlier ages in offspring exposed to the low choline diet (Supplemental Figure 1).

Figure 2B shows the results in 6 months-old WT and Tg2576 mice. In animals that received the intermediate diet, a two-way ANOVA showed no effect of genotype (F(1,36)=0.01, p=0.907) or task phase (F(1,36)=2.36, p=0.133), revealing memory deficits in WT and Tg2576 mice (Figure 2B). In contrast, the high choline group showed a main effect of task phase (F(1,56)=22.18, p<0.0001) but not genotype (F(1,56)=1.78, p=0.188). Tukey-Kramer post-hoc analyses showed a significant increase in the exploration of the novel object during testing in both WT (p<0.001) and Tg2576 (p=0.020; Figure 2B). Thus, the high choline-treated mice showed object location memory but the mice fed the intermediate diet did not. The mice that received the relatively low choline diet showed an effect of genotype (F(1,50)=4.36, p=0.042; Figure 2B) in that training in WT mice differed from testing in Tg2576 mice (p=0.031). However, there was no effect of task phase (F(1,50)=3.75, p=0.058). Thus, WT mice did not differ between training and testing (p=0.114) and Tg2576 mice did not either (p=0.921). Therefore, mice fed the low choline and intermediate diets were impaired and the high choline-treated mice were not.

The discrimination indices are shown in Figure 3 and results led to the same conclusions as the analyses in Figure 2. For the 3 months-old mice (Figure 3A), the low choline group did not show the ability to perform the task for WT or Tg2576 mice. Thus, a two-way ANOVA showed no effect of genotype (F(1,74)=0.027, p=0.870) or task phase (F(1,74)=1.41, p=0.239). For the intermediate diet-treated mice, there was no effect of genotype (F(1,50)=0.3.52, p=0.067) but there was an effect of task phase (F(1,50)=8.33, p=0.006). WT mice showed a greater discrimination index during testing relative to training (p=0.019) but Tg2576 mice did not (p=0.664). Therefore, Tg2576 mice fed the intermediate diet were impaired. In contrast, high choline-treated mice performed well. There was a main effect of task phase (F(1,68)=39.61, p=<0.001) with WT (p<0.0001) and Tg2576 mice (p=0.0002) showing preference for the moved object in the test phase. Interestingly, there was a main effect of genotype (F(1,68)=4.50, p=0.038) because the discrimination index for WT training was significantly different from Tg2576 testing (p<0.0001) and Tg2576 training was significantly different from WT testing (p=0.0003).

Novel object location results based on the discrimination index.

A. Results are shown for the 3 months-old WT and Tg2576 mice based on the discrimination index.

1. Mice fed the low choline diet showed object location memory only in WT.

2. Mice fed the intermediate diet showed object location memory only in WT.

3. Mice fed the high choline diet showed memory both for WT and Tg2576 mice. Therefore, the high choline diet improved memory in Tg2576 mice.

B. The results for the 6 months-old mice are shown.

1-2. There was no significant memory demonstrated by mice that were fed either the low or intermediate choline diet.

3. Mice fed a diet enriched in choline showed memory whether they were WT or Tg2576 mice. Therefore, choline enrichment improved memory in all mice.

The discrimination indices of 6 months-old mice led to the same conclusions as the results in Figure 2. There was no evidence of discrimination in low choline-treated mice by two-way ANOVA (no effect of genotype, (F(1,42)=3.25, p=0.079; no effect of task phase, F(1,42)=0.278, p=0.601). The same was true of mice fed the intermediate diet (genotype, F(1,12)=1.44, p=0.253; task phase, F(1,12)=2.64, p=0.130). However, both WT and Tg2576 mice performed well after being fed the high choline diet (effect of task phase, (F(1,52)=58.75, p=0.0001, but not genotype (F(1,52)=1.197, p=0.279). Tukey-Kramer post-hoc tests showed that both WT (p<0.0001) and Tg2576 mice that had received the high choline diet (p=0.0005) had elevated discrimination indices for the test session.

Taken together, these results demonstrate the lasting beneficial effects of the high choline diet and adverse effects of low choline on offspring in a spatial memory task in Tg2576 as well as WT littermates.

NOR

We first confirmed that animals did not show preference for one object over the other in training. Indeed, a two-way ANOVA revealed no main effects of genotype at 3 months (F(1,71)=2.59, p=0.536) and no effect of diet (F(2,71)=0.22, p=0.809). The same result was obtained at 6 months of age (genotype, (F(1,66)=0.11, p=0.746); diet, (F(2,66)=0.98, p=0.376). Thus, novel object exploration during training approached 50% in all treatment groups, independent of genotype and maternal diet, and at both ages.

Figure 4 shows the results for the NOR task. Lines connect the performance of a given animal in training and in testing. Supplemental Figure 3 shows the data with mean ± sem. Figure 4A shows that there were no impairments at 3 months of age. In animals treated with the low choline diet, there was a main effect of task phase (F(1,52)=19.81, p<0.0001), no effect of genotype (F(1,52)=0.020, p=0.887), and more exploration of the novel object during the test phase (WT, p=0.007; Tg2576, p<0.002). In animals that received the intermediate diet, there also was a significant main effect of task phase (F(1,38)=30.88, p<0.0001) and no effect of genotype (F(1,38)=0.97, p=0.330).

WT and Tg2576 mice showed object recognition memory regardless of diet.

A. Three months-old mice.

1-3. WT and Tg2576 mice performed the NOR task.

B. Six months-old mice.

1-3. WT and Tg2576 mice performed the NOR task.

Animals explored the novel object significantly more during testing in WT (p=0.014) and Tg2576 mice (p=0.002; Figure 3B). For the high choline group, there also was an effect of task phase (F(1,26)=17.51, p=0.0003) and not genotype (F(1,26)=3.53, p=0.072) and a significant increase in the exploration of the novel object during testing in both WT (p=0.006) and Tg2576 (p=0.027; Figure 3B).

Figure 4B shows the data for 6 months of age. Animals that received the low choline diet performed the NOR task at 6 months of age (two-way ANOVA, task phase (F(1,26)=32.66, p<0.0001; genotype (F(1,26)=0.048, p=0.821). Tukey-Kramer post-hoc tests showed that there was a significant increase in the exploration of the novel object in testing both in WT (p<0.001) and Tg2576 (p= 0.003). For mice that had been fed the intermediate choline diet, they also performed the NOR task at 6 months of age (two-way ANOVA, task phase (F(1,19)=13.65, p=0.002); genotype (F(1,19)=0.28, p=0.604). Tukey-Kramer post-hoc analyses revealed a significant increase in the exploration of the novel object during testing compared to training for both WT (p=0.0027) and Tg2576 (p=0.039; Figure 4B). Mice fed the high choline diet also showed no deficit in NOR at 6 months of age. Thus, a two-way ANOVA revealed a main effect of task phase (F(1,27)=16.26, p=0.0004) but not genotype (F(1,27)=0.24, p=0.625).

Tukey-Kramer post-hoc-hoc analyses revealed a significant increase in the exploration of the novel object during testing in both WT (p=0.003) and Tg2576 (p=0.007; Figure 4B).

Thus, at both 3 and 6 months of age, WT and Tg2576 mice performed well in the NOR task. These results suggest that young Tg2576 mice are less sensitive to NOR than NOL. The greater sensitivity to NOL is consistent with past demonstrations that the DG contributes to NOL (Sahay et al. 2011; Kesner et al. 2015; Spyrka and Hess 2018; Vandrey et al. 2020; Gulmez Karaca et al. 2021; GoodSmith et al. 2022). Also, our implementation of NOL may have increased the DG-dependence of the task by making the object locations relatively close together, because in studies by Pofahl and colleagues, it was shown that distances between objects like those we used made NOL DG-dependent (Pofahl et al. 2021).

Exploration time

Total object exploration (TOE) was measured to address any effects of genotype or diet on the total duration of exploration of objects (Figure 5). For NOL TOE at 3 months of age, there were effects of genotype (F(1,67)=6.891, p=0.01) but not diet (F(2,87)=0.67, p=0.63). Tukey-Kramer post-hoc tests showed that Tg2576 mice fed the low choline diet showed more exploration than WT mice fed the same diet but the effect was on the border of significance (p=0.049; data not shown). Three months-old mice tested with NOR showed no significant effect of genotype (F(1,76)=1.35, p=0.25) or diet (F(2,76)=0.30, p=0.61). Since there were weak or no effects of genotype, all genotypes are pooled for Figure 5A1 (NOL) and Figure 5B1 (NOR).

Tg2576 mice treated with the high choline diet spent more time with the objects at each approach.

A-B. NOL.

There was no effect of diet on time exploring or approaches to the novel object in the NOL task at 3 months of age (A) but there were significant differences at 6 months of age (B). Mice exposed to the high choline diet spent more time exploring than mice that had been fed the low choline or the intermediate diet.

C-D. NOR.

There was no effect of diet on time exploring or approaches to the novel object in the NOR task at 3 months of age (C) but there were significant differences at 6 months of age (D). Mice that had been fed the high choline diet spent more time exploring than mice that had been fed the low choline or the intermediate diet.

For NOL TOE at 6 months of age, there was no effect of genotype (F(1,66)=0.33; p=0.57). For NOR TOE, the same was true (genotype: F(1,72)=0.96, p=0.33; diet: F(2,72)=8.50, p=0.0005). Because genotype was not a significant factor, we pooled genotypes (NOL, Figure 5C1; NOR, Figure 5D2).

With genotypes pooled we determined how diets differed. For NOL TOE, mice treated with the low choline diet had less exploration than mice fed the high choline diet (one-way ANOVA, F(2,77)=6.90; p=0.002; Tukey post-hoc test, p=0.005; Figure 5C1) and the mice that were fed the intermediate diet also had less exploration than mice fed the high choline diet (Tukey post-hoc test, p=0.008; Figure 5C1). Results for NOR were similar: mice treated with the low choline diet had less exploration than mice fed the high choline diet (one-way ANOVA, F(2,74)=4.81; p=0.020; Tukey post-hoc test, p=0.016; Figure 5D1) and the mice that were fed the intermediate diet also had less exploration than mice fed the high choline diet (Tukey post-hoc test, p=0.03; Figure 5D1).

To gain insight into the potential reason for the effect of diet on exploration at 6 months of age, we measured total object approaches (TOA was defined as the number of approaches to the familiar object + number of approaches to the novel object (Figure 5). There was no effect of diet on TOA for NOL (F(2,75)=2.88; p=0.092; Figure 5C2) or NOR (F(2,74)=1.81, p=0.171; Figure 5D2).

Taken together, the results indicate that, in 6 months-old mice, animals that received the high choline diet spent more time with objects at each approach. This could explain the increased object memory in high choline-treated mice because there would bemore time for information processing during an approach. Another possibility is that as Tg2576 mice age they show compensatory changes that enhance memory. This has been suggested for procedural learning (Middei et al. 2004) but to the best of our knowledge, it has not been shown for memory of objects. A third possibility is that the high choline diet reduces anxiety (Glenn et al. 2012; Langley et al. 2015; McCall et al. 2015). Reduced anxiety may lessen fear of exploring objects, and as a result, animals may spend more time with objects.

Anatomy

NeuN

To further analyze the effects of dietary choline early in life, we used an antibody against a neuronal nuclear antigen, NeuN. Previous studies have found that reduced expression of NeuN often occurs in neurons after insults or injury (e.g., ischemia, toxicity and even aging; Lind et al. 2005; Portiansky et al. 2006; Buckingham et al. 2008; Kadriu et al. 2009; Matsuda et al. 2009; Won et al. 2009; Duffy et al. 2011). For example, when NeuN is reduced after a heterozygous deletion in ankyrin-rich membrane spanning kinase D-interacting substrate of 220K, ARMS/Kidins, a protein critical to neurotrophin signaling, neurons in the entorhinal cortex demonstrate loss of NeuN immunoreactivity and the cellular morphology is abnormal (Duffy et al. 2011; Duffy et al. 2015). NeuN is also reduced when Aβ levels are elevated (Wu et al. 2016) and NeuN is reduced in AD patients (Camporez et al. 2021). Therefore, we initially examined NeuN-ir in Tg2576 mice. We found it was reduced and then asked in the high choline diet could restore it.

As shown in Figure 6A, relatively weak NeuN-ir was observed in Tg2576 that received the low choline diet and the intermediate diet compared to the high choline diet (Figure 6A). This observation is consistent with the vulnerability of the hilus to insult and injury (Scharfman 1999). To examine the septotemporal axis, quantification from 2-3 anterior coronal sections were averaged to provide a value for the septal pole of the DG and 2-3 posterior sections were averaged to assess the caudal DG. The most ventral, temporal levels were not sampled.

Choline supplementation improved NeuN immunoreactivity (ir) in hilar cells in Tg2576 animals

A. Representative images of NeuN-ir staining in the anterior DG.

1. A section from a Tg2576 mouse fed the low choline diet. The area surrounded by a box is expanded below. Red arrows point to NeuN-ir hilar cells. Mol=molecular layer, GCL=granule cell layer. Calibration for the top row, 100 µm; for the bottom row, 50 µm.

2. A section from a Tg2576 mouse fed the intermediate diet. Same calibrations as for 1.

3. A section from a Tg2576 mouse fed the high choline diet. Same calibrations as for 1.

4. A section from a WT mouse fed the intermediate diet. Same calibrations as for 1.

B. Quantification methods. Representative images demonstrate the thresholding criteria used to quantify NeuN-ir.

1. A NeuN-stained section. The area surrounded by the white box is expanded in the inset (arrow) to show 3 hilar cells. The 2 NeuN-ir cells above threshold are marked by blue arrows. The 1 NeuN-ir cell below threshold is marked by a green arrow.

2. After converting the image to grayscale, the cells above threshold were designated as red. The inset shows that the two cells that were marked by blue arrows are red while the cell below threshold is not.

3. An example of the threshold menu from ImageJ showing the way the threshold was set. Sliders (red circles) were used to move the threshold to the left or right of the histogram of intensity values. The final position of the slider (red arrow) was positioned at the onset of the steep rise of the histogram.

C. NeuN-ir in Tg2576 and WT mice. Tg2576 mice had either the low, intermediate, or high choline diet in early life. WT mice were fed the standard diet (intermediate choline).

1. Tg2576 mice treated with the high choline diet had significantly more hilar NeuN-ir cells in the anterior DG compared to Tg2576 mice that had been fed the low choline or intermediate diet. The values for Tg2576 mice that received the high choline diet were not significantly different from WT mice, suggesting that the high choline diet restored NeuN-ir.

2. There was no effect of diet or genotype in the posterior DG.

There was no effect of diet (F(2,30)=2.11, p=0.137) nor rostral-caudal level (F(1,30)=0.02, p=0.877) but there was a significant interaction of diet and rostral-caudal level (F(2,30)=3.88, p=0.036). A one-way ANOVA using anterior values showed a significant effect of diet (F(2,21)=7.58, p=0.003) but this was not the case for posterior values (F(2,23)=0.13; p=0.876). In the anterior DG, there was less NeuN-ir in mice fed the low choline diet versus the high choline diet (Tukey-Kramer post-hoc test, p=0.029) and less NeuN in mice fed the intermediate vs high choline diet (p=0.003). The results suggest that hilar neurons of Tg2576 mice have reduced NeuN-ir and choline enrichment protected dorsal hilar neurons from NeuN loss in Tg2576 mice.

To ask if the improvement in NeuN after the high choline diet in Tg256 restored NeuN to WT levels, we used WT mice. The WT mice were fed the intermediate diet because it is the standard mouse chow, and this group was intended to reflect normal mice. For the analysis, we used a one-way ANOVA with 4 groups: Low choline Tg2576, Intermediate Tg2576, High choline Tg2576, and Intermediate WT (Figure 5C). Tukey-Kramer multiple comparisons tests were used as the post hoc tests. The results showed a significant group difference for anterior DG (F(3,25)=9.20; p=0.0003; Figure 5C1) but not posterior DG (F(3,28)=0.867; p=0.450; Figure 5C2). Regarding the anterior DG, there were more NeuN-ir cells in high choline-treated mice than both low choline (p=0.046) and intermediate choline-treated Tg2576 mice (p=0.003). WT mice had more NeuN-ir cells than Tg2576 mice fed the low (p=0.011) or intermediate diet (p=0.003). Tg2576 mice that were fed the high choline diet were not significantly different from WT (p=0.827). Thus, we found reduced hilar NeuN in Tg2576 mice and the high choline diet increased NeuN protein expression to WT levels.

ΔFosB

To complement the information from the video-EEG recordings (see below) we used a marker of elevated neuronal activity, ΔFosB (Figure 7). ΔFosB is a truncated variant of the transcription factor FosB, which is increased by enhanced neuronal activity; ΔFosB has a half-life of approximately 8 days (Ulery-Reynolds et al., 2009), so when ΔFosB is elevated, it reflects increased neuronal activity over the last 10-14 days (McClung et al. 2004). Previous studies have shown that when the J20 mouse model of AD is examined with an antibody to ΔFosB, the GC layer shows extremely high levels of ΔFosB expression (Corbett et al. 2017), similar to a mouse with chronic spontaneous seizures (Morris et al. 2000). This is not surprising since the J20 mice have recurrent seizures (Palop et al. 2007). Therefore, we asked if Tg2576 mice would have robust ΔFosB in the GC layer, and choline supplementation would reduce it. We also included WT mice fed the intermediate (standard) diet. These WT mice would allow us to address the possibility that the high choline diet restored ΔFosB back to normal.

Choline supplementation reduced ΔFosB expression in dorsal GCs of Tg2576 mice.

A. Representative images of ΔFosB staining in GCL..

1. A section from a low choline-treated Tg2576 mouse shows robust ΔFosB-ir in the GCL. The area outlined by the box is expanded below. Red arrows point to Δ FfosB-labeled cells. GCL=granule cell layer. Calibration for the top row, 100 µm; for the bottom row, 50 µm.2-3. Sections from intermediate (2) and high choline (3)-treated Tg2576 mice. Same calibrations as 1.

4, A section from a WT mouse treated with the intermediate diet. Same calibrations as 1.

B. Quantification methods. Representative images demonstrating the thresholding criteria established to quantify ΔFosB.

1. A ΔFosB -stained section shows strongly-stained cells (white arrows).

2. A strict thresholding criteria was used to make only the darkest stained cells red.

C. Use of the strict threshold to quantify ΔFosB-ir.

1. Anterior DG. Tg2576 mice treated with the choline supplemented diet had significantly less ΔFosB-ir compared to the Tg2576 mice fed the low or intermediate diets. Tg2576 mice fed the high choline diet were not significantly different from WT mice, suggesting a rescue of ΔFosB-ir.

2. There were no significant differences in ΔFosB-ir in posterior sections.

D. Methods are shown using a threshold that was less strict.

1. Some of the stained cells that were included are not as dark as those used for the strict threshold (white arrows).

2. All cells above the more permissive threshold are shown in red.

E. Use of the less strict threshold to quantify ΔFosB-ir.

1. Anterior DG. Tg2576 mice that were fed the high choline diet had less ΔFosB-ir pixels than the mice that were fed the other diets. There were no differences from WT mice, suggesting restoration of ΔFosB-ir by choline enrichment in early life.

2. Posterior DG. There were no significant differences between Tg2576 mice fed the 3 diets or WT mice.

There was strong expression of ΔFosB in Tg2576 GCs in mice fed the low choline diet (Figure 7A1). The high choline diet and intermediate diet appeared to show less GCL ΔFosB-ir (Figure 7A2-3). A two-way ANOVA was conducted with the experimental group (Tg2576 low choline diet, Tg2576 intermediate choline diet, Tg2576 high choline diet, WT intermediate choline diet) and location (anterior or posterior) as main factors. There was a significant effect of group (F(3,32)=13.80, p=<0.0001) and location (F(1,32)=8.69, p=0.006). Tukey-Kramer post-hoc tests showed that Tg2576 mice fed the low choline diet had significantly greater ΔFosB-ir than Tg2576 mice fed the high choline diet (p=0.0005) and WT mice (p=0.0007). Tg2576 mice fed the low and intermediate diets were not significantly different (p=0.275). Tg2576 mice fed the high choline diet were not significantly different from WT (p>0.999). There were no differences between groups for the posterior DG (all p>0.05).

ΔFosB quantification was repeated with a lower threshold to define ΔFosB-ir GCs (see Methods) and results were the same (Figure 7D). Two-way ANOVA showed a significant effect of group (F(3,32)=14.28, p< 0.0001) and location (F(1,32)=7.07, p=0.0122) for anterior DG but not posterior DG (Figure 7D). For anterior sections, Tukey-Kramer post hoc tests showed that low choline mice had greater ΔFosB-ir than high choline mice (p=0.0024) and WT mice (p=0.005) but not Tg2576 mice fed the intermediate diet (p=0.275); Figure 7D1). Mice fed the high choline diet were not significantly different from WT (p=0.993; Figure 7D1). These data suggest that high choline in the diet early in life can reduce neuronal activity of GCs in offspring later in life. In addition, low choline has an opposite effect, suggesting low choline in early life has adverse effects.

IIS

Previous research has shown that Tg2576 mice exhibit IIS (Figure 8A) starting at very young ages: 4 weeks (Kam et al. 2016) or 6 weeks (Bezzina et al. 2015). Therefore, we performed video-EEG using cortical (left frontal cortex, right occipital cortex) and hippocampal (left and right) electrodes in WT and Tg2576 mice (Figure 8B-C). Animals were recorded for 24 hrs each session so that the major behavioral states (exploration, awake rest, and sleep) were sampled well. Consistent with previous studies (Bezzina et al. 2015; Kam et al. 2016) we observed IIS in Tg2576 mice but not WT littermates. Therefore, analyses below were only in Tg2576 mice.

The high choline diet reduced IIS frequency in Tg2576 animals.

A. 1. Representative example of an IIS (red arrow). IIS were defined as occurring in all 4 channels (generalized) to distinguish them from focal spikes that can often be artifact.

2-4. The IIS shown in A1 is expanded.

B. 1. Scatter plot of IIS frequency at each age of recording.

2. Means and sem are plotted. The high choline diet group had fewer IIS than the intermediate diet group at ages 1-3 months ($, p<0.05) and the low choline group had less IIS than the intermediate diet at ages 1-4 months (@, p<0.05). The high and low choline diet groups were significantly different at ages 3 and 4 months (*, p<0.05).

As shown in Figure 8, animals that received the intermediate diet had a significantly higher number of IIS in the 24 hrs-long recording periods compared to animals that received the high choline and the low choline diets. A two-way ANOVA (mixed model analysis) showed that there was a significant effect of age (F(2,37)=3.38; p=0.036) and maternal diet (F(2,36)=8.12; p=0.089). At the 5 weeks-old recording, Tukey-Kramer post-hoc analyses showed that IIS frequency in animals treated with the intermediate diet was higher compared to animals treated with the low (p=0.027) or high choline diets (p=0.038). This also was true for 2 months and 3 months (low choline, high choline, p<0.05). At 4 months, the low choline group had significantly reduced IIS frequency compared to the mice that had received the intermediate diet (p=0.009) but this was not the case for the high choline group compared to the intermediate group (p=0.976). At 5-6 months, IIS frequencies were not significantly different in the mice fed the different diets (all p>0.05), probably because IIS frequency becomes increasingly variable with age (Kam et al. 2016). One source of variability is seizures, because there was a sharp increase in IIS during the day before and after a seizure (Supplemental Figure 4). Another reason that the diets failed to show differences was that the IIS frequency generally declined at 5-6 months. This can be appreciated in Figure 8B and Supplemental Figure 6B. These data are consistent with prior studies of Tg2576 mice where IIS increased from 1 to 3 months but then waxed and waned afterwards (Kam et al., 2016).

Seizures and premature mortality in mice fed the low choline diet

We found that mice fed the low choline diet had greater ΔFosB-ir in GCs and the hilus showed very low NeuN-ir. Therefore, we asked whether low choline-treated mice had more seizures than mice fed the other diets. We recorded 8 mice by video-EEG for 5 days each (at 6 months of age) to examine seizures and found 2 mice from the low choline group had seizures (11 seizures over 2 of the 5 days in one mouse, 1 seizure in the other mouse) whereas none of the other mice had seizures (n=0/4, 2 intermediate and 2 high choline, data not shown).

These values are probably an underestimate for the low choline group because many mice in this group appeared to die in a severe seizure prior to 6 months of age (Supplemental Figure 5). Therefore, the survivors at 6 months probably were the subset with few seizures. The reason we think that low choline-treated mice appeared to die in a seizure was that they were found in a specific posture in their cage which occurs during a severe seizure that leads to death, a prone posture with extended, rigid limbs (Supplemental Figure 5).

Regardless of how the mice died, there was greater mortality in the low choline group compared to mice that had been fed the high choline diet (Log-rank (Mantel-Cox) test, Chi square 5.36, df 1, p=0.021; Supplemental Figure 5A).

Correlations between IIS and other measurements

As shown in Figure 9A, IIS were correlated to behavioral performance in some conditions. For these correlations, only mice that were fed the low and high choline diets were included because mice that were fed the intermediate diet and were tested behaviorally were not always tested with EEG.

Correlations between IIS, Behavior, and hilar NeuN-ir.

A. IIS frequency over 24 hrs is plotted against the preference for the novel object in the test phase of NOL. A greater preference is reflected by a greater percentage of time exploring the novel object.

1. The mice fed the high choline diet (red) showed greater preference for the novel object when IIS were low. These data suggest IIS impaired object location memory in the high choline-treated mice. The low choline-treated mice had very weak preference and very few IIS, potentially explaining the lack of correlation in these mice.

2. There were no significant correlations for IIS and NOR. However, there were only 4 mice for the high choline group, which is a limitation.

B. IIS frequency over 24 hrs is plotted against the number of dorsal hilar cells expressing NeuN. The dorsal hilus was used because there was no effect of diet on the posterior hilus..

1. Hilar NeuN-ir is plotted against the preference for the novel object in the test phase of NOL. There were no significant correlations.

2. Hilar NeuN-ir was greater for mice that had better performance in NOR, both for the low choline (blue) and high choline (red) groups. These data support the idea that hilar cells contribute to object recognition (Kesner et al. 2015; Botterill et al. 2021; GoodSmith et al. 2022).

For NOL, IIS frequency over 24 hrs was plotted against the preference for the novel object in the test phase (Figure 9A). IIS were significantly less frequent when behavior was the best, but only for the high choline-treated mice (Pearson’s r, p=0.022). In the low choline group, behavioral performance was poor regardless of IIS frequency (Pearson’s r, p=0.933; Figure 9A1). For NOR, there were no significant correlations (low choliNe, p=0.202; high choline, p=0.680) but few mice were tested in the high choline-treated mice (Figure 9B2).

We also tested whether there were correlations between dorsal hilar NeuN-ir cell numbers and IIS frequency. In Figure 9B, IIS frequency over 24 hrs was plotted against the number of dorsal hilar cells expressing NeuN. The dorsal hilus was used because there was no effect of diet on the posterior hilus. For NOL, there was no significant correlation (low choline, p=0.273; high choline, p=0.159; Figure 9B1). However, for NOR, there were more NeuN-ir hilar cells when the behavioral performance was strongest (low choline, p=0.024; high choline, p=0.016; Figure 9B2). These data support prior studies showing that hilar cells, especially mossy cells (the majority of hilar neurons), contribute to object recognition (Botterill et al. 2021; GoodSmith et al. 2022).

Sex differences

As shown in Figures 2-3, there appeared to be no sex differences in NOL or NOR. For confirmation, we studied NOL further, since that was the task that showed effects of diet. A 3-way ANOVA with sex, genotype and task phase as factors showed no significant effect of sex at 3 months of age (low choline diet, F(1,35)=0.0001; p=0.99). Results were the same for the 3 months-old mice fed the intermediate diet (F(1,22)=0.062; p=0.810) and 3 months-old mice that were fed the high choline diet (F(1,34)=0.432; p=0.522). At 6 months of age there also were no significant effects of sex (low choline diet, F(1,23)=0.343; p=0.571; intermediate diet, F(1,16)=2.49; p=0.130; high choline diet, F(1,27)=0.29; p=0.873).

Regarding IIS, we did not find sex differences (Supplemental Figure 6). In Supplemental Figure 6A, comparisons are made of males and females fed the low or high choline diet at 5 weeks, 2, and 3 months-of age. A two-way ANOVA with sex and diet as factors showed no effect of sex (F(1,20) =2.48; p=0.132) or diet (F(1,20)=3.56; p=0.067). Note that for this comparison we used the low choline and high choline groups only because there were few mice for each sex at all ages. The results for months 2-3 also showed no effect of sex (2 months: F(1,17)=0.637; p=0.429; 3 months: F(1,15)=0.178; p=0.668).

In Supplemental Figure 6B we grouped mice at 1-2 months, 3-4 months and 5-6 months so that there were sufficient females and males to compare each diet. A two-way ANOVA with diet and sex as factors showed a significant effect of diet (F(2,47)=46.21; p<0.0001) at 1-2 months of age, but not sex (F1,47)=0.11, p=0.758). Post-hoc comparisons showed that the low choline group had fewer IIS than the intermediate group, and the same was true for the high choline-treated mice. Thus, female mice fed the low choline diet differed from the females (p<0.0001) and males (p<0.0001) fed the intermediate diet. Male mice that had received the low choline diet different from females (p<0.0001) and males (p<0.0001) fed the intermediate diet. Female mice fed the high choline diet differed from females (p=0.002) and males (p<0.0001) fed the intermediate diet, and males fed the high choline diet differed from females (p<0.0001) and males (p<0.0001) fed the intermediate diet.

For the 3-4 months-old mice there was also a significant effect of diet (F(2,32)=10.82, p=0.0003) but not sex (F(1,32)=1.05, p=0.313). Post-hoc tests showed that low choline females were different from males fed the intermediate diet (p=0.007), and low choline males were also significantly different from males that had received the intermediate diet (p=0.006). There were no significant effects of diet (F(2,23)=1.21, p=0.317) or sex (F(1,23)=0.84, p=0.368) at 5-6 months of age.

Discussion

Summary

This study showed that choline supplementation in early life had several beneficial effects in Tg2576 mice. The high choline diet led to improved behavior in the NOL task, improved expression of ΔFosB, and reduced IIS frequency.

We also found surprising effects of treating mice with relatively low choline in the maternal diet. The low choline-treated Tg2576 mice had the least hilar NeuN-ir and most GCL ΔFosB protein expression. There was an impairment in the NOL task not only in low choline-treated Tg2576 mice but also WT mice. In addition, the low choline-treated Tg2576 mice showed premature mortality compared to Tg2576 mice fed the high choline diet. These results were surprising because the relatively low levels of choline have been considered by others to be a control (Moreno et al. 2013; Mellott et al. 2017; Velazquez et al. 2019). Our data suggest the low choline diet led to several adverse effects. Yet the IIS frequency was low in the low choline-treated Tg2576 mice, suggesting a beneficial effect. It is possible the chronic overexpression of GC ΔFosB led to reduced IIS by altering GC gene expression, since ΔFosB regulates the GC transcriptome (Corbett et al. 2017; You et al. 2018; Stephens et al. 2020). Alterations in GC transcription could impair IIS generation because the GCs appear to be a site where IIS are generated (Lisgaras and Scharfman 2023). Altered GC gene expression might also explain the reduced performance in the NOL task of low choline-treated mice, because the DG plays an important role in this detecting novel object locations normally (Lee et al. 2005).

We showed for the first time that Tg2576 mice have pathology in the hilus an early age in that hilar neurons are deficient in immunostaining with an anti-NeuN antibody, and this occurs before amyloid plaque develops (after 6 months of age; Hsaio et al., 1996). Tg2576 offspring that were fed the high choline diet showed a restoration of hilar NeuN staining. Because hilar neurons play a role in NOL (Bui et al. 2018), it is possible that rescue of hilar neurons by the high choline diet led to improved spatial memory. However, there was no significant correlation between the number of NeuN-ir hilar cells and testing for the NOL task. There was a significant correlation for the NOR task, with both low and high choline groups performing better when hilar NeuN-ir cell numbers were greater. The significant correlation with NOR test performance is consistent with a role of somatostatin-expressing hilar GABAergic neurons (Nagarajan et al. 2024) and hilar glutamatergic mossy cells in NOR and object-related activity (Botterill et al. 2021; GoodSmith et al. 2022). Somatostatin-expressing hilar neurons (Tallent 2007; Savanthrapadian et al. 2014; Hofmann et al. 2016) and mossy cells (Sloviter 1994; Scharfman 1999; Scharfman and Myers 2012; Jinde et al. 2013; Bui et al. 2018; Botterill et al. 2019) have been suggested to contribute to excitability of GCs, so phenotypic rescue may also have contributed to the ability of the high choline diet to reduce hyperexcitability.

Benefits of high choline

The results of this study are consistent with previous reports that prenatal or postnatal choline supplementation improves object memory. One study of iron deficiency showed that choline supplementation improved NOR (Kennedy et al. 2014). Wistar rats treated with a diet high (5.0 g/kg) in choline had improved NOR relative to rats that had been fed a diet lower (1.1 g/kg) in choline (Moreno et al. 2013). In rats that were aged to 24 months, animals that had been fed a high choline diet showed improved NOR, but only in females (Glenn et al. 2008).

Although few studies have examined the effects of MCS on hyperexcitability, our results showing the benefits of high choline are consistent with reports that a methyl-enriched diet (including choline but also betaine, folic acid, vitamin B12, L-methionine and zinc) reduced spike wave discharges (Sarkisova et al. 2023) and audiogenic seizure severity (Poletaeva et al. 2014) in the offspring. Perinatal ethanol treatment that increased excitability was mitigated by choline chloride injection from postnatal day 10-30 (Grafe et al. 2022).

To our knowledge, our study is the first to show that MCS can exert effects on IIS and ΔFosB. However, prior studies have shown that MCS in rats improved memory after severe seizures. Using a convulsant to induce several hours of severe seizures (status epilepticus, SE), there is usually memory impairment in subsequent weeks; MCS reduced the impairment (Yang et al. 2000; Holmes et al. 2002; Wong-Goodrich et al. 2011).

Our results are consistent with previous studies of mouse models of AD. In the APP/PS1 mouse model, where mutations found in AD are present in APP and PS1, it was shown that life-long choline supplementation, using the high choline diet, improved memory in aged mice compared to a low choline diet (Velazquez et al. 2019). Postnatal choline supplementation also improved memory in APP/PS1 mice (Wang et al. 2019). Other improvements were also shown in the APP/PS1 mice, such as increased choline acetyltransferase, the major enyzme for acetylcholine synthesis (Mellott et al. 2017; Velazquez et al. 2019). In 3xFAD mice, where there are 3 familial AD mutations, a diet without choline had numerous deleterious consequences, including increased Aβ and tau phosporylation (Dave et al. 2023)

Adverse effects of the diet with relatively low choline

The relatively low choline diet had several adverse effects, which was surprising because the low choline levels are not considered very low in some prior studies (Moreno et al. 2013; Mellott et al. 2017; Velazquez et al. 2019). However, the past studies using low choline did not test Tg2576 mice. Past studies sometimes used the diet for only part of gestation rather than all of gestation and stopped after birth instead of continuing until weaning. In contrast, we fed the diet throughout gestation and until weaning. Nevertheless, it is surprising. One possible explanation is an interaction of the low choline diet with the strain of the Tg2576 mice, SJL (Swiss James Lambert). To our knowledge this strain has not been tested with different diets before. SJL mice are descended from Swiss Webster, and are prone to reticulum cell sarcoma (Haran-Ghera et al. 1967). This strain is also characterized by vulnerability to infection causing a multiple sclerosis-like syndrome (Linzey et al. 2023).

One of the adverse effects was high mortality. Mice appeared to die in seizures. Mice also had high ΔFosB expression in the GCs. Another adverse effect was weak NeuN-ir in the hilus.. For both WT and Tg2576 mice, NOL was impaired. These data suggest that the offspring of mothers fed the low choline diet were unhealthy.. If mice were unhealthy, IIS might have been reduced due to impaired synchronization (despite high excitability in the DG). Another reason for reduced IIS is that the mice that had the low choline diet had seizures which interrupted REM sleep. Less REM sleep would reduce IIS because IIS occur primarily in REM. Also, seizures in the Tg2576 mice were followed by a depression of the EEG (postictal depression; Supplemental Figure 3) that would transiently reduce IIS. A different perspective is that the intermediate diet promoted IIS (rather than low choline reducing IIS). Instead of choline, a constituent of the intermediate diet may have promoted IIS.

NeuN

As mentioned above, NeuN is a neuronal nuclear antigen that can be phosphorylated, and when that occurs the antibody to NeuN no longer binds to NeuN. NeuN is phosphorylated in response to oxidative damage, brain injury, and toxicity (Lind et al. 2005). Therefore, it was of interest when we saw that hilar neurons of the dentate gyrus showed reduced NeuN in Tg2576 mice. Hilar neurons are mainly glutamatergic mossy cells and somatostatin (SOM)/neuropeptide Y (NPY)-expressing GABAergic neurons (HIPP cells; (Houser 2007; Scharfman and Myers 2012; Scharfman 2016), and both neuronal types are implicated in spatial memory functions of the GCs, as well as their excitability (Myers and Scharfman 2009; 2011; Scharfman and Myers 2012; Jinde et al. 2013; Scharfman 2016; Raza et al. 2017; Bui et al. 2018; GoodSmith et al. 2019; Li et al. 2021). If damaged by the high intracellular APP and Aβ levels in young Tg2576 mice, the hilar neurons would be expected to show NeuN loss. This idea is consistent with deficits in SOM and NPY-stained cells in AD (Chan-Palay 1987), indicating a vulnerability.

The reason for rescue of anterior hilar but not posterior hilar NeuN with the high choline diet is unclear. The greater sensitivity of dorsal neurons may be related to differences in gene expression patterns, since some of the genes that are differentially expressed along the dorsal-ventral axis could affect vulnerability to insult or injury (Cembrowski et al. 2016; Zhang et al. 2018). However, there was a similar trend for the data in both anterior and posterior regions. The lack of statistical significance in posterior DG may simply have been due to be greater variance in the posterior DG data.

ΔFosB

Given that seizures are associated with elevated GC expression of ΔFosB (Chen et al. 1997; McClung et al. 2004; Corbett et al. 2017; You et al. 2017), the differences in ΔFosB protein expression levels observed here could be the result of increased seizures in animals that received the low choline diet. However, we did not detect a significant difference in seizures at 6 months of age, although there may have been a difference earlier in life or if we had recorded for a longer time. Interestingly, there may have been more seizures due to a reduced number of IIS. Thus, some evidence suggests that IIS can abort a seizure (Staley and Dudek 2006).

ΔFosB is a transcription factor that is linked to cognition. In the J20 mouse model of AD, elevated ΔFosB in GCs led to reduced cognition, and when ΔFosB was selectively reduced the cognition improved (Corbett et al. 2017). Therefore, the reduction in ΔFosB by the high choline diet was important to show hyperexcitability was reduced and also important because it showed how high choline may benefit cognition.

Choline and cholinergic neurons

There are many suggestions for the mechanisms that allow MCS to improve health of the offspring. One hypothesis that we are interested in is that MCS improves outcomes by reducing IIS. Reducing IIS would potentially reduce hyperactivity, which is significant because neuronal activity can increase release of Aβ and tau (Cirrito et al. 2005; Cirrito et al. 2008; Bero et al. 2011; Yamada et al. 2014; Yamamoto et al. 2015; Hettinger et al. 2018). IIS would also be likely to disrupt sleep since it represents aberrant synchronous activity over widespread brain regions. The disruption to sleep could impair memory consolidation since consolidation is a notable function of sleep (Graves et al. 2001; Poe et al. 2010). Indeed, in patients with AD, IIS and similar events, IEDs, are correlated with memory impairment (Vossel et al. 2016). Sleep disruption also has other negative consequences such as impairing normal clearance of Aβ (Nedergaard and Goldman 2020).

How would choline supplementation in early life reduce IIS of the offspring? It may do so by making BFCNs more resilient. That is significant because BFCN abnormalities appear to cause IIS. Thus, selective silencing of BFCNs reduced IIS (Lisgaras and Scharfman, 2023). The cholinergic antagonist atropine also reduces IIS when injected systemically (Kam et al., 2016), and it reduced elevated synaptic activity of GCs in young Tg2576 mice in vitro (Alcantara-Gonzalez et al. 2021). These studies are consistent with the idea that early in AD there is elevated cholinergic activity (DeKosky et al. 2002; Ikonomovic et al. 2003; Kelley et al. 2014; Mufson et al. 2015; Kelley et al. 2016), while later in life there is cholinergic degeneration. Indeed, the overactivity of cholinergic neurons of the first months of life could cause the degeneration at older ages.

Why would MCS make BFCNs resilient? There are several possibilities that have been explored, based on genes upregulated by MCS. One attractive hypothesis is that neurotrophic support for BFCNs is retained after MCS but in aging and AD it declines (Gautier et al. 2023). The neurotrophins, notably nerve growth factor (NGF) and brain-derived neurotrophic factor (BDNF) have been known for a long time to support the health of BFCNs (Mufson et al. 2003; Niewiadomska et al. 2011).

Limitations

One of the limitations of the study was that the maternal diets were not matched exactly. Although the low and high choline diets only had differences in choline, the intermediate diet had different amounts of choline as well as other constituents (Supplementary Table 1). Therefore, the only diet comparison with a difference restricted to choline is the low versus high choline diets (Supplemental Table 1). The intermediate diet was useful however, because numerous studies in AD mouse models employ this standard diet.

In addition, groups were not exactly matched. Although WT mice do not have IIS, a WT group for each of the diets would have been useful. Sample sizes were also not matched exactly, because several mice died.. However, this is unlikely to have had a major effect because IIS were low in frequency in all groups over the last months of the study.

Regarding sex differences, there may have been differences if females had been separated by stage of the estrous cycle at death. This possibility is raised by prior data showing that rats and mice during proestrous and estrous mornings have hyperexcitability but not at other cycle stages (Scharfman et al. 2003).

The Tg2576 mouse model is one of many murine models of AD, and as with all models there are inherent limitations. The Tg2576 model recapitulates familial AD, whereas the majority of AD is sporadic. Tg2576 mice also lack tau pathology. However, MCS has now demonstrated structural/functional benefits in several AD-relevant models, namely Tg2576 (this report), APP/PS1 (Alldred et al. 2021; Dave et al. 2023), Ts65Dn (Velazquez et al. 2013; Powers et al. 2016; Strupp et al. 2016; Powers et al. 2017; Alldred et al. 2021; Alldred et al. 2023) and 3xTg mice (Dave et al. 2023). Moreover, increasing evidence suggests that humans with AD have low serum choline and are improved by dietary choline (Dave et al. 2023; Judd et al. 2023a; b).

Conclusions

There is now a substantial body of evidence that MCS promotes learning and memory in the offspring of normal rats and improves behavior and other abnormalities in mice that simulate AD. There also is evidence that serum levels of choline are low in AD. Therefore, it is exciting to think that dietary choline might improve AD. Given the past work also shows benefits to mice that simulate DS, the benefits of choline supplementation appear to extend to DS. The present study adds to the growing consensus that MCS is restorative by showing that hyperexcitability and pathology in the DG of Tg2576 mice are improved by MCS. They also suggest that there may be some adverse effects of a relatively low level of choline in the diet.